### 2018-01-29 cabios-template/main.tex
start drafting manuscript

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

%https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3463124/

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[Rethomics]{Rethomics: an R framework to analyse high-throughput behavioural data}
\author[Geissmann \textit{et~al}.]{Quentin Geissmann\,$^{\text{\sfb 1,}*}$, Luis Garcia\,$^{\text{\sfb 2}}$, Esteban Beckwith\,$^{\text{\sfb 1}}$,  and Giorgio Gilestro\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
	\textbf{Motivation:}
		Ethomics, a quantitative and high-throughput approach to ethology, is a novel and exciting field.
		The recent development of methods that automatically score variables in multiple animals provides an unprecedented insight into the study of behaviours. 
		The analysis of ethomics data presents many challenges that are conceptually independent of the acquisition platform.
		However, there is, little effort in providing a general framework to analyse multiple and long behavioural time series.
		We developped the  Rethomics framework, a suite of R packages that altogether provide utilities to:
		import, store, visualise and analyse behavioural data.\\
	\textbf{Availability and Implementation:}
	All packages in the rethomics framework are available under the terms of the GPLv3 license.
	Exaustive installation instructions and tutorials are available at
 		\href{https://github.com/rethomics}{https://github.com/rethomics/} 
 		\href{https://rethomics.github.io/}{https://rethomics.github.io/}.\\
	\textbf{Contact:}
		\href{qgeissmann@gmail.com}{qgeissmann@gmail.com}
	}

\maketitle

\section{Introduction}
TODOs:

	1. Ethomics, a scaling up in time but also in number of animals.
	2. A diversity of platform to do the same thing.
	3. Need for well documented programmatic interface  innovative/advances analysis
	4. The idea of unifying behavioural analysis, so community (experimental and quantitative biologists) collaborates

%\enlargethispage{12pt}

\section{Approach}

Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
Text Text Text Text Text Text Text Text Text Text Text Text Text.
Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
Text Text Text  Text Text Text Text Text Text  Text Text.
\citealp{Boffelli03} might want to know about text text text text

\section{Framework  features}
\subsection{Data import}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Internal data structure}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Circadian analysis}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Sleep scoring}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Visualisation}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl

\section{Conclusion}

xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl

\section*{Acknowledgements}

Han Kim. Patrick Kr{\"a}tschmer

\section*{Funding}

This work has been supported by the... Text Text  Text Text.\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}

\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01}
Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

### 2018-01-29 cabios-template/main.tex
draft a third of paper!

@@
 		However, there is, little effort in providing a general framework to analyse multiple and long behavioural time series.
-		We developped the  Rethomics framework, a suite of R packages that altogether provide utilities to:
+		We developped the \texttt{rethomics} framework, a suite of R packages that altogether provide utilities to:
 		import, store, visualise and analyse behavioural data.\\
 	\textbf{Availability and Implementation:}
-	All packages in the rethomics framework are available under the terms of the GPLv3 license.
+	All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license.
 	Exaustive installation instructions and tutorials are available at
@@
 \section{Introduction}
-TODOs:
+Animal behaviours are complex phenotypes that are manifestations of the interaction between the nervous system and the external or internal environment.
+In the last few decades, our ability to record large amounts of various phenotypical data has tremendously increased.
+Behaviour scoring is certainly not an exception to this trend.
+Indeed, many platforms (todo citations) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
 
-	1. Ethomics, a scaling up in time but also in number of animals.
-	2. A diversity of platform to do the same thing.
-	3. Need for well documented programmatic interface  innovative/advances analysis
-	4. The idea of unifying behavioural analysis, so community (experimental and quantitative biologists) collaborates
+The availability of ever so large amounts of data is very exciting as it paves the way for refined analyses.
+Clearly, the multiplicity of model organisms, hypotheses and paradigms makes the diversity of recording tools crucial.
+However, regarding subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
+Instead, tools tend to consist of graphical interfaces with strictly defined functionalities that only import data from a single platform.
+There are three issues with this approach.
+First of all, state of the art analysis and visualisation requires a flexibility than only a programmatic interface can provide.
+Secondly, it favours replicated work as developers need to provide their own implementation to very similar problems.
+Lastly, it links analysis and visualisation to the target platform, which makes it very difficult to share cross-platform tools.
+
+Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
+Typically, the behaviour of each individual is a long time series (possibly multivariate and heterogeneous).
+In addition, each individual has to be unambiguously identified and associated with arbitrary metadata defined by the experimenter (sex, treatment, genotype and so on).
+Efficiently combining and manipulating these information, on datasets of hundreds of individuals, each recorded for weeks, is not trivial.
+The availability of such a tool help promoting the analysis of behaviour as a data science.
+
+Here, we describe \texttt{rethomics}, a framework that unifies analysis of large behavioural dataset in an efficient and manner.
+It offers an elegant solution to store, manipulate and visualise a large amount of behavioural data.
+Rethomics comes with a extensive documentation and a set of both practical and theoretical tutorials.
 
@@
 \section{Approach}
-
-Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text
-Text Text Text Text Text Text Text Text Text Text Text Text Text.
-Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
-Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about text text text text
+%Figure~2\vphantom{\ref{fig:02}} shows that the above method  Text
+\texttt{rethomics} is implemented as a collection of small packages linked to one another.
+This development model follows modern frameworks such as the \texttt{tidyverse}, which results in increases testability and maintainability.
+The different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
+are explicitly handled by different packages (Fig.~1\vphantom{\ref{fig:01}}).
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers an very flexible and efficient solution to store both large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table} derived object.
+Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the input platform.
+Analyses results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state visualisation and statistics tools.
 

### 2018-01-29 plos-latex-template/manuscript.tex
change of plan, we are now writing a PLoS paper, big time!

% Template for PLoS
% Version 3.4 January 2017
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

\usepackage{amsmath,amssymb}
\usepackage{changepage}
\usepackage[utf8x]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{cite}
\usepackage{nameref,hyperref}
\usepackage[right]{lineno}
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
\usepackage[table]{xcolor}
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

\usepackage[colorinlistoftodos]{todonotes}
%\usepackage[colorinlistoftodos,disable]{todonotes}

%% Include all macros below
\newcommand{\citationneeded}[2][]{\todo[color=brown, fancyline, #1]{\textbf{Citation Needed:} #2}}

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Rethomics: an R framework to analyse high-throughput behavioural data} 
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Quentin Geissmann\textsuperscript{1*},
Luis Garcia Rodriguez\textsuperscript{2},
Esteban J Beckwith\textsuperscript{1},
Giorgio F Gilestro\textsuperscript{1*}
\\
\bigskip
\textbf{1} Department of Life Sciences, Imperial College London, London, United Kingdom
\\
\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country %TODO @luis
\\
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* qgeissmann@gmail.com, giorgio@gilest.ro

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Ethomics, a quantitative and high-throughput approach to ethology, is a novel and exciting field.
The recent development of methods that automatically score variables in multiple animals provides an unprecedented insight into the study of behaviours. 
The analysis of ethomics data presents many challenges that are conceptually independent of the acquisition platform.
However, there is, little effort in providing a general framework to analyse multiple and long behavioural time series.
We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether provide utilities to:
import, store, visualise and analyse behavioural data.
The \texttt{rethomics} framework is available and documented at \href{https://github.com/rethomics}{https://github.com/rethomics/}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author summary}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non %sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque %pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

Animal behaviours are complex phenotypical manifestations of the interaction between nervous systems and their external or internal environment.
In the last few decades, our ability to record large amounts of various phenotypical data has tremendously increased.
Behaviour scoring is certainly not an exception to this trend.
Indeed, many platforms (todo citations) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).

The availability of ever so large amounts of data is very exciting as it paves the way for refined analyses.
Clearly, the multiplicity of model organisms, hypotheses and paradigms makes the diversity of recording tools crucial.
However, regarding subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
Instead, tools tend to consist of graphical interfaces with strictly defined functionalities that only import data from a single platform.
There are three issues with this approach.
First of all, state of the art analysis and visualisation requires a flexibility than only a programmatic interface can provide.
Secondly, it favours replicated work as developers need to provide their own implementation to address very similar problems.
Lastly, it links analysis and visualisation to the target platform, which makes it very difficult to share cross-platform tools.

Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
Typically, the behaviour of each individual is a long time series (possibly multivariate and heterogeneous).
In addition, each individual has to be unambiguously identified and associated with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype). Efficiently combining and manipulating these information, on datasets of hundreds of individuals, each recorded for weeks, is not trivial. The availability of a unified ethomics toolbox would help promoting the analysis of behaviour as a data science.

Here, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
It offers an elegant solution to store, manipulate and visualise a large amount of data.
Rethomics comes with a extensive documentation and a set of both practical and theoretical tutorials.

\section*{Design and Implementation}

\texttt{rethomics} is implemented as a collection of small packages linked to one another.
This development model follows modern frameworks such as the \texttt{tidyverse}, which results in increases testability and maintainability.
The different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
are explicitly handled by different packages (Fig.~1\vphantom{\ref{fig:01}}).
At the core of \texttt{rethomics}, the \texttt{behavr} package offers an very flexible and efficient solution to store both large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table} derived object.
Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
Analyses results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state visualisation and statistics tools.

\subsection{Data import}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Internal data structure}
The baha
\subsection{Circadian analysis}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Sleep scoring}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
\subsection{Visualisation}
xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl

\section*{Results}

%\todo{write something here!}

\section*{Availability and Future Directions}
All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at  		\href{https://github.com/rethomics}{https://github.com/rethomics/}.
Extensive installation instructions as well as reproducible demos and tutorials are available at
\href{https://rethomics.github.io/}{https://rethomics.github.io/}.

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
%\paragraph*{S1 Fig.}
%\label{S1_Fig}
%{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\begin{thebibliography}{10}

\bibitem{bib1}
Conant GC, Wolfe KH.
\newblock {{T}urning a hobby into a job: how duplicated genes find new
  functions}.
\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.

\bibitem{bib2}
Ohno S.
\newblock Evolution by gene duplication.
\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
  Springer-Verlag.; 1970.

\bibitem{bib3}
Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
  infection through a transposon insertion followed by a {D}uplication}.
\newblock PLoS Genet. 2011 Oct;7(10):e1002337.

\end{thebibliography}

\end{document}

### 2018-01-30 plos-latex-template/manuscript.tex
add ppl in acknoledgments
description of implementation, layout of results

@@
 \section*{Abstract}
-Ethomics, a quantitative and high-throughput approach to ethology, is a novel and exciting field.
+Ethomics, a quantitative and high-throughput approach to ethology, is a new and exciting field.
 The recent development of methods that automatically score variables in multiple animals provides an unprecedented insight into the study of behaviours. 
@@
 
-The availability of ever so large amounts of data is very exciting as it paves the way for refined analyses.
+The availability large amounts of data is very exciting as it paves the way for in depth analyses.
 Clearly, the multiplicity of model organisms, hypotheses and paradigms makes the diversity of recording tools crucial.
@@
 
-Here, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
+In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
 It offers an elegant solution to store, manipulate and visualise a large amount of data.
@@
 
-\texttt{rethomics} is implemented as a collection of small packages linked to one another.
-This development model follows modern frameworks such as the \texttt{tidyverse}, which results in increases testability and maintainability.
+\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:rethomics}).
+This development model follows modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
 The different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
-are explicitly handled by different packages (Fig.~1\vphantom{\ref{fig:01}}).
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers an very flexible and efficient solution to store both large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table} derived object.
+are explicitly handled by different packages.
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store both large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
-Analyses results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state visualisation and statistics tools.
+Analyses results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistics tools.
 
-\subsection{Data import}
-xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
-\subsection{Internal data structure}
-The baha
-\subsection{Circadian analysis}
-xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
-\subsection{Sleep scoring}
-xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl 
-\subsection{Visualisation}
-xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl xxx xxxxx xxxx dwjdkl
+\subsection*{Internal data structure}
+We created \texttt{behavr} (Fig~\ref{fig:data-table}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
+two challenges that inherent to manipulating ethomics results.
+
+Firstly, there could be very long (typically $k_i > 10^8$), multivariate (often, $q > 10$), time series for each individual.
+For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second over days.
+Therefore, data structure must be computationally efficient -- both in term of memory footprint and processing speed.
+
+Secondly, a large amount of individuals are often studied (typically $n > 100$).
+Each individual ($i$) is associated with metadata, that is a set of $p$ ``metavariables'' that describe experimental conditions.
+For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, post-hock observations and other arbitrary metavariables.
+It is good practice to record as many metavariables as possible so they can later be used as covariates. 
+Therefore, typically $p > 10$.
+
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to access, and manipulate metadata.
+This approach guaranties that any data point can be mapped correctly to its parent metadata.
+It also allows implicit update of metadata when data is altered.
+For instance, when is data filtered, only the remaining individuals should be in the new metadata. 
+Metadata and data also have to interoperate.
+For instance, when one wants to update variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male''$).
+
+\subsection*{Data import}
+Data import package translate results from a recording platform (e.g. text files and databases) into a single \texttt{behavr} object.
+Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data  and another one for Ethoscope data.
+Although the structure of the raw results if very different, conceptually, loading data is very similar.
+In all cases the user is asked to generate a metadata table (one row per individual). 
+The columns are of two types: mandatory and optional.
+The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
+The optional columns are user-defined arbitrary fields that relate to the experiment itself (\emph{e.g.} condition and sex).
+
+In this respect the metadata file is standardised an comprehensive data frame describing of an experiment.
+Using such a structure comes with multiple advantages.
+For instance, it simplify collaboration and data exchange as all treatments and individuals are very explicit.
+Then, it promotes good experimental practices such as interspersion of treatments (indeed, without it, users are tempted to simplify their design, for instance, confounding device/location and treatment).
+Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
+Indeed, additional replicates can simply be added as new rows, and replicate number later user, if needed, as a covariate.
+
+\subsection*{Visualisation}
+Long time series often need to be preprocessed before visualisation.
+Typically, users are interested in understanding the individual or population trends over time.
+To integrate visualisation in \texttt{rethomics},
+we implemented \texttt{ggetho}, a package extending the widely adopted \texttt{ggplot2} by providing preprocessing tools as well as new layers and scales.
+Our tools make full use of the internal \texttt{behavr} structure to deliver efficient representations of temporal trends.
+It particularly applies to the visualisation of long experiments, with the ability to, for instance, annotate light and dark phases, wrap time over a circadian day, display ``double-plotted actograms'' and periodograms.
+Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot}.
+
+\subsection*{Circadian and sleep analysis}
+The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
+Together, they offer a suite of methods to compute periodograms and define find peaks, score sleep from inactivity (\emph{e.g.} using the ``five minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
 
 \section*{Results}
+We present an small example examining the circadian behaviour of 64 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
+A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
+In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
+Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
+Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
+Two replicates were performed at different times.
+raw data and metadata files are publicly available TODO cite zenodo
 
-%\todo{write something here!}
+\subsection*{Data loading}
+
+\subsection*{Quality control}
+
+\subsection*{Actograms}
+
+\subsection*{Spectogram}
 
@@
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
+All packages are continuously integrated and unit tested on several version of \texttt{R} to minimise resent and future issues.
+
+\section*{Acknowledgements}
+TODO:
+
+Han Kim
+
+Maite
+
+Patrick Kr{\"a}tschmer
 
@@
 %
+
 \begin{thebibliography}{10}

### 2018-02-02 plos-latex-template/results/scipt.R
add maite's result and analysis. needs some work

@@ -1,0 +1,39 @@
+rm(list=ls())
+library(ggetho)
+library(damr)
+library(zeitgebr)
+
+met <- damr::link_dam2_metadata("./metadata.csv", ".")
+dt <- load_dam2(met)
+
+summary(dt)
+dt[,t := t - days(xmv(baseline_days))]
+dt[, moving :=  activity > 0]
+dt[, label := interaction(1:.N, genotype), meta=T]
+dt
+
+ggetho(dt, aes(y=label, z=moving)) + 
+  stat_tile_etho() + 
+  stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
+  stat_ld_annotations(x_limits = c(0, dt[,max(t)]), ld_colours = c("white", "grey"))
+
+# remove dead animals here
+
+ggetho(dt, aes(z=moving), multiplot=2) + 
+  stat_tile_etho() +
+  facet_wrap(~label) 
+
+# from the overview, we decide to 
+dt_ll <- dt[t > days(1)]
+
+ggetho(dt_ll, aes(z=moving/2), multiplot=2) + 
+  stat_tile_etho() +
+  facet_wrap(~label) 
+
+per_dt <- periodogram(moving, 
+						dt_ll, 
+						FUN=chi_sq_periodogram, 
+						period_range = c(hours(15), hours(32)))
+per_dt <- find_peaks(per_dt)
+
+ggperio(per_dt, aes(y = power, peak=peak)) + geom_line() + geom_peak() + facet_wrap(~label) 

### 2018-02-02 plos-latex-template/manuscript.tex

@@
 \section*{Abstract}
-Ethomics, a quantitative and high-throughput approach to ethology, is a new and exciting field.
-The recent development of methods that automatically score variables in multiple animals provides an unprecedented insight into the study of behaviours. 
-The analysis of ethomics data presents many challenges that are conceptually independent of the acquisition platform.
-However, there is, little effort in providing a general framework to analyse multiple and long behavioural time series.
-We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether provide utilities to:
+Ethomics, a quantitative and high-throughput approach to animal behaviour, is a new and exciting field.
+The recent development of automatised methods that can score various behaviours on a large number of animals
+% brings the promise of a powerful and original insight
+provides biologists with an unprecedented tools to decipher these complex phenotypes. 
+Analysing ethomics data comes with many challenges that are, conceptually, independent of the acquisition platform used.
+However, there is, little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
+We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offer utilities to:
 import, store, visualise and analyse behavioural data.
-The \texttt{rethomics} framework is available and documented at \href{https://github.com/rethomics}{https://github.com/rethomics/}.
+In this article, we describe it and show an example of its application to the blooming field of sleep and circadian rhythm in fruit fly.
+The \texttt{rethomics} framework is available and documented at \href{https://rethomics.github.io}{https://rethomics.github.io}.
 
@@
 Animal behaviours are complex phenotypical manifestations of the interaction between nervous systems and their external or internal environment.
-In the last few decades, our ability to record large amounts of various phenotypical data has tremendously increased.
+In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
 Behaviour scoring is certainly not an exception to this trend.
@@
 
-The availability large amounts of data is very exciting as it paves the way for in depth analyses.
-Clearly, the multiplicity of model organisms, hypotheses and paradigms makes the diversity of recording tools crucial.
-However, regarding subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces with strictly defined functionalities that only import data from a single platform.
-There are three issues with this approach.
-First of all, state of the art analysis and visualisation requires a flexibility than only a programmatic interface can provide.
-Secondly, it favours replicated work as developers need to provide their own implementation to address very similar problems.
-Lastly, it links analysis and visualisation to the target platform, which makes it very difficult to share cross-platform tools.
+The availability of large amounts of data is very exciting as it paves the way for in-depth analyses.
+Clearly, the multiplicity of model organisms, hypotheses and paradigms should be reflected in  diverse range of recording tools.
+However, when it comes to the subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
+Instead, tools tend to consist of graphical interfaces with rigid functionalities that only import data from a single platform.
+There are, at least, three issues with this approach.
+First of all, state-of-the-art analysis and visualisation requires a level of reproducibility, flexibility and scalability than only a programmatic interface can provide.
+Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
+Lastly, it links analysis and visualisation to the target tool, which makes it very difficult to share cross-tool utilities and concepts.
 
 Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
-Typically, the behaviour of each individual is a long time series (possibly multivariate and heterogeneous).
-In addition, each individual has to be unambiguously identified and associated with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating these information, on datasets of hundreds of individuals, each recorded for weeks, is not trivial.
-The availability of a unified ethomics toolbox would help promoting the analysis of behaviour as a data science.
+Typically, the behaviour of each individual is described by a long time series (possibly multivariate and heterogeneous).
+Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
+Efficiently combining and manipulating these two types of information, on datasets of hundreds of individuals, each recorded for weeks, is not trivial.
 
 In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
-It offers an elegant solution to store, manipulate and visualise a large amount of data.
-Rethomics comes with a extensive documentation and a set of both practical and theoretical tutorials.
+It offers an elegant computational solution to store, manipulate and visualise a large amount of data.
+We expect it to fill the gap between behavioural biology and data sciences.
+\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical tutorials.
 
@@
 
-\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:rethomics}).
-This development model follows modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
-The different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
+\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:framework}).
+This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
+In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
 are explicitly handled by different packages.
@@
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
-Analyses results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistics tools.
+Results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistics tools.
+
+% Place figure captions after the first paragraph in which they are cited.
+
+\begin{figure}[!h]
+    \includegraphics[width=1\textwidth]{fig/framework.pdf}
+	\caption{{\bf The \texttt{rethomics} workflow.}
+		Diagram representing, from left to right, the interplay between the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
+	\label{fig:framework}
+\end{figure}
 
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:data-table}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
-two challenges that inherent to manipulating ethomics results.
+We created \texttt{behavr} (Fig~\ref{fig:behavr}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
+two challenges that are inherent to manipulating ethomics results.
 
-Firstly, there could be very long (typically $k_i > 10^8$), multivariate (often, $q > 10$), time series for each individual.
-For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second over days.
+Firstly, there could be very long (typically $k_i > 10^8, \forall i \in [1,n]$), multivariate (often, $q > 10$), time series for each individual.
+For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second, over multiple days.
 Therefore, data structure must be computationally efficient -- both in term of memory footprint and processing speed.
@@
 Each individual ($i$) is associated with metadata, that is a set of $p$ ``metavariables'' that describe experimental conditions.
-For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, post-hock observations and other arbitrary metavariables.
+For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
 It is good practice to record as many metavariables as possible so they can later be used as covariates. 
@@
 
-\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to access, and manipulate metadata.
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata.
 This approach guaranties that any data point can be mapped correctly to its parent metadata.
@@
 For instance, when is data filtered, only the remaining individuals should be in the new metadata. 
-Metadata and data also have to interoperate.
-For instance, when one wants to update variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male''$).
+It is also important that metadata and data can interoperate.
+For instance, when one wants to update variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
+
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/behavr.pdf}
+	\caption{{\bf Example of a \texttt{behavr} table.}
+		The metadata holds a single row for each of the $n$ individuals. 
+		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to map the data) or user-defined (i.e. arbitrary).
+		In the data, each row is a ``read'' (\emph{i.e.} information about one animal at one time point).
+		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual.
+		Data and metadata are implicitly joined with the \texttt{id} field.
+		Notably, the names used for variables and metavariable are only plausible examples that will likely differ in practice. 
+	}
+	\label{fig:behavr}
+\end{figure}
 
 \subsection*{Data import}
-Data import package translate results from a recording platform (e.g. text files and databases) into a single \texttt{behavr} object.
+Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
 Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data  and another one for Ethoscope data.
@@
 In all cases the user is asked to generate a metadata table (one row per individual). 
-The columns are of two types: mandatory and optional.
+In it, there will be both mandatory and optional columns.
 The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
@@
 
-In this respect the metadata file is standardised an comprehensive data frame describing of an experiment.
+In this respect the metadata file is a standardised an comprehensive data frame describing an experiment.
 Using such a structure comes with multiple advantages.
@@
 Long time series often need to be preprocessed before visualisation.
-Typically, users are interested in understanding the individual or population trends over time.
+Typically, users are interested in understanding individual or population trends over time.
 To integrate visualisation in \texttt{rethomics},
@@
 The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute periodograms and define find peaks, score sleep from inactivity (\emph{e.g.} using the ``five minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
+Together, they offer a suite of methods to compute periodograms and find their peaks, score sleep from inactivity (\emph{e.g.} using the ``five minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
 
 \section*{Results}
-We present an small example examining the circadian behaviour of 64 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
+We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
 A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
@@
 
+* Other inputs
+
+* Position analysis
+
+* GUI
+
+* ...
+
 \section*{Acknowledgements}
@@
 Maite
+
+Hannah
 

### 2018-02-08 plos-latex-template/results/script.Rmd
result script

---
output:
  pdf_document:
    keep_tex: true
---

# Results

## Description
In order to illustrate some mechanisms in rethomics, we provide a simple and reproducible example of analysis of 
circadian phenotype recorded with DAM2 monitors -- a widely addopted paradigm.

TODO:
* data source -- The data was obtained from ...(citation) TODO.
* description (genotypes etc) -- exhaustive description in the `metadata.csv`

## Data loading
First of all, the necessary `rethomics` packages are loaded.

```{r, message=FALSE}
library(damr)      # input DAM2 data
library(zeitgebr)  # periodogram computation
library(sleepr)    # sleep analysis
library(ggetho)    # behaviour visualisation
```

Then, the metadata file is read and linked to the `.txt` result files.
```{r}
met <- damr::link_dam2_metadata("./metadata.csv", ".") # linking
dt <- load_dam2(met)                                   # loading
summary(dt)                                            # quick summary
```

## Preprocessing

We notice, from the metadata, that the two replicates do not have the same time in baseline.
We would like to express the time relative to the important event: thetransition to `LL`. 
The best way is to sustract the `baseline_days` metavariable from the `t` variable.
This gives us an opportunity to illustrate the use `xmv()` that maps metavariables as variables.
In addition, we use the `data.table` syntax to create, in place, a `moving` variable.
It is `TRUE` when and only when `activity` is greater than zero:
```{r}
dt[,t := t - days(xmv(baseline_days))]    # baseline sustraction. not the use of xmv
dt[, moving :=  activity > 0]  
```

To simplify visualisation, we create our own `label` metavariable, as combination of a number and `genotype`.
In the restricted context of this analysis, this acts a unique identifyier.
Importanly, we keep `id` , which is more rigourous and universal.

```{r}
dt[, label := interaction(1:.N, genotype), meta=T]
# print(dt)
```

## Curation

It is important to see an overview of how each individual and experiment behaved and, if necessary, alter the data accordingly. We save this figure as Fig 3A.

```{r}
# make a ggplot object with label on the y and moving on the z axis
fig3A <- ggetho(dt, aes(y=label, z=moving)) +  
  # show data as a tile plot. That is z is a pixel whose intensity maps moving
  stat_tile_etho() +                  
  # add layers to draw annotations to show L and D phases as white and black
  # the first layer is for the baseline (until t=0)
  stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
  # in the 2nd one, we start at 0 and use grey instead of black as we work in LL
  stat_ld_annotations(x_limits = c(0, dt[,max(t)]), ld_colours = c("white", "grey"))
```

Dead or escaped animals are falsely scored as long series of zero-activity.
Our `sleepr` packges offer a tool to detect and remove this artifactual data:

```{r}
dt <- sleepr::curate_dead_animals(dt, moving)
# make a ggplot object with label on the y and moving on the z axis
fig3B <- ggetho(dt, aes(y=label, z=moving)) +  
    stat_tile_etho() +                  
    stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
    stat_ld_annotations(x_limits = c(0, dt[,max(t)]), ld_colours = c("white", "grey"))
```
The updated version can be visualised in Fig 3B.

For the purpose of this example, we keep only individuals that have at least five days in LL.
```{r}
valid_dt <- dt[ , .(valid = max(t) > days(5)), by=id]
valid_ids <- valid_dt[valid == T, id]
dt <- dt[id %in% valid_ids]
summary(dt)
```
Note that as a result, we now have `r nrow(dt[meta=T])` "valid" individuals.

## Double plotted actograms

A common way of representing rythmicity in circadian experiments is to compute "double-plotted actograms".
In Fig S1A, we show all double plotted actograms layed out in a grid.

```{r}
figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
            stat_bar_tile_etho() + 
            facet_wrap( ~ label, ncol=4) +
            scale_y_discrete(name="Day")
```

## Periodograms

For each indidual, we compute a $\chi{}^2$ periodogram and we layout all of them in a grid (Fig S1B).

```{r}
dt_ll <- dt[t > days(1)]
per_dt <- periodogram(moving, 
						dt_ll, 
						resample_rate = 1/mins(10),
						FUN=chi_sq_periodogram)

per_dt <- find_peaks(per_dt)

figS1B <- ggperio(per_dt, aes(y = power, peak=peak)) + 
      geom_line() +
      geom_line(aes(y=signif_threshold), colour="red") + 
      geom_peak() + 
      facet_wrap( ~ label, ncol=4)

```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
################################ CODE BELOW FOR COSMETIC ALTERATIONS OF FIGURES

library(cowplot)

my_theme <- theme(axis.title.y=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.line = element_blank(),
            panel.border = element_blank(),
            legend.position="none")

legend <- fig3A + theme(axis.title=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks=element_blank(),
                        axis.line = element_blank(),
                        #axis.ticks.y=element_blank(),
                        panel.border = element_blank(),
                        #legend.text = "moving",
                        legend.position="left") + 
                  coord_cartesian(xlim=-days(c(10, 9))) + scale_y_discrete(position = "right")

fig3 <- cowplot::plot_grid(
                          legend,
                          fig3A +  my_theme,
                          fig3B +  my_theme,
                          nrow=1,
                          labels=c("", "A", "B"),
                          rel_widths = c(1, 2, 2),
                          label_size = 20,
                          axis="b",
                          align="h"
                          )
pdf("fig3.pdf",w=12,h=12)
print(fig3)
dev.off()

figS1A <- figS1A +  theme(strip.text.x = element_text(size =8))
figS1B <- figS1B +  theme(strip.text.x = element_text(size =8))

figS1 <- cowplot::plot_grid(
                          figS1A,
                          figS1B,
                          nrow=1,
                          labels=c( "A", "B"),
                          label_size = 20,
                          align="v"
                          )

pdf("figS1.pdf",w=16,h=16)
print(figS1)
dev.off()
```

### 2018-02-08 plos-latex-template/results/script.tex
result script

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\begin{document}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{description}{%
\subsection{Description}\label{description}}

In order to illustrate some mechanisms in rethomics, we provide a simple
and reproducible example of analysis of circadian phenotype recorded
with DAM2 monitors -- a widely addopted paradigm.

TODO: * data source -- The data was obtained from \ldots{}(citation)
TODO. * description (genotypes etc) -- exhaustive description in the
\texttt{metadata.csv}

\hypertarget{data-loading}{%
\subsection{Data loading}\label{data-loading}}

First of all, the necessary \texttt{rethomics} packages are loaded.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(damr)      }\CommentTok{# input DAM2 data}
\KeywordTok{library}\NormalTok{(zeitgebr)  }\CommentTok{# periodogram computation}
\KeywordTok{library}\NormalTok{(sleepr)    }\CommentTok{# sleep analysis}
\KeywordTok{library}\NormalTok{(ggetho)    }\CommentTok{# behaviour visualisation}
\end{Highlighting}
\end{Shaded}

Then, the metadata file is read and linked to the \texttt{.txt} result
files.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{met <-}\StringTok{ }\NormalTok{damr}\OperatorTok{::}\KeywordTok{link_dam2_metadata}\NormalTok{(}\StringTok{"./metadata.csv"}\NormalTok{, }\StringTok{"."}\NormalTok{) }\CommentTok{# linking}
\NormalTok{dt <-}\StringTok{ }\KeywordTok{load_dam2}\NormalTok{(met)                                   }\CommentTok{# loading}
\KeywordTok{summary}\NormalTok{(dt)                                            }\CommentTok{# quick summary}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## behavr table with:
##  58  individuals
##  8   metavariables
##  2   variables
##  1.58722e+05 measurements
##  1   key (id)
\end{verbatim}

\hypertarget{preprocessing}{%
\subsection{Preprocessing}\label{preprocessing}}

We notice, from the metadata, that the two replicates do not have the
same time in baseline. We would like to express the time relative to the
important event: thetransition to \texttt{LL}. The best way is to
sustract the \texttt{baseline\_days} metavariable from the \texttt{t}
variable. This gives us an opportunity to illustrate the use
\texttt{xmv()} that maps metavariables as variables. In addition, we use
the \texttt{data.table} syntax to create, in place, a \texttt{moving}
variable. It is \texttt{TRUE} when and only when \texttt{activity} is
greater than zero:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt[,t }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\NormalTok{t }\OperatorTok{-}\StringTok{ }\KeywordTok{days}\NormalTok{(}\KeywordTok{xmv}\NormalTok{(baseline_days))]    }\CommentTok{# baseline sustraction. not the use of xmv}
\NormalTok{dt[, moving }\OperatorTok{:}\ErrorTok{=}\StringTok{  }\NormalTok{activity }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{]  }
\end{Highlighting}
\end{Shaded}

To simplify visualisation, we create our own \texttt{label}
metavariable, as combination of a number and \texttt{genotype}. In the
restricted context of this analysis, this acts a unique identifyier.
Importanly, we keep \texttt{id} , which is more rigourous and universal.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt[, label }\OperatorTok{:}\ErrorTok{=}\StringTok{ }\KeywordTok{interaction}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{.N, genotype), meta=T]}
\CommentTok{# print(dt)}
\end{Highlighting}
\end{Shaded}

\hypertarget{curation}{%
\subsection{Curation}\label{curation}}

It is important to see an overview of how each individual and experiment
behaved and, if necessary, alter the data accordingly. We save this
figure as Fig 3A.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# make a ggplot object with label on the y and moving on the z axis}
\NormalTok{fig3A <-}\StringTok{ }\KeywordTok{ggetho}\NormalTok{(dt, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{label, }\DataTypeTok{z=}\NormalTok{moving)) }\OperatorTok{+}\StringTok{  }
\StringTok{  }\CommentTok{# show data as a tile plot. That is z is a pixel whose intensity maps moving}
\StringTok{  }\KeywordTok{stat_tile_etho}\NormalTok{() }\OperatorTok{+}\StringTok{                  }
\StringTok{  }\CommentTok{# add layers to draw annotations to show L and D phases as white and black}
\StringTok{  }\CommentTok{# the first layer is for the baseline (until t=0)}
\StringTok{  }\KeywordTok{stat_ld_annotations}\NormalTok{(}\DataTypeTok{x_limits =} \KeywordTok{c}\NormalTok{(dt[,}\KeywordTok{min}\NormalTok{(t)], }\DecValTok{0}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\CommentTok{# in the 2nd one, we start at 0 and use grey instead of black as we work in LL}
\StringTok{  }\KeywordTok{stat_ld_annotations}\NormalTok{(}\DataTypeTok{x_limits =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, dt[,}\KeywordTok{max}\NormalTok{(t)]), }\DataTypeTok{ld_colours =} \KeywordTok{c}\NormalTok{(}\StringTok{"white"}\NormalTok{, }\StringTok{"grey"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Dead or escaped animals are falsely scored as long series of
zero-activity. Our \texttt{sleepr} packges offer a tool to detect and
remove this artifactual data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt <-}\StringTok{ }\NormalTok{sleepr}\OperatorTok{::}\KeywordTok{curate_dead_animals}\NormalTok{(dt, moving)}
\CommentTok{# make a ggplot object with label on the y and moving on the z axis}
\NormalTok{fig3B <-}\StringTok{ }\KeywordTok{ggetho}\NormalTok{(dt, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{label, }\DataTypeTok{z=}\NormalTok{moving)) }\OperatorTok{+}\StringTok{  }
\StringTok{    }\KeywordTok{stat_tile_etho}\NormalTok{() }\OperatorTok{+}\StringTok{                  }
\StringTok{    }\KeywordTok{stat_ld_annotations}\NormalTok{(}\DataTypeTok{x_limits =} \KeywordTok{c}\NormalTok{(dt[,}\KeywordTok{min}\NormalTok{(t)], }\DecValTok{0}\NormalTok{)) }\OperatorTok{+}
\StringTok{    }\KeywordTok{stat_ld_annotations}\NormalTok{(}\DataTypeTok{x_limits =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, dt[,}\KeywordTok{max}\NormalTok{(t)]), }\DataTypeTok{ld_colours =} \KeywordTok{c}\NormalTok{(}\StringTok{"white"}\NormalTok{, }\StringTok{"grey"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

The updated version can be visualised in Fig 3B.

For the purpose of this example, we keep only individuals that have at
least five days in LL.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{valid_dt <-}\StringTok{ }\NormalTok{dt[ , .(}\DataTypeTok{valid =} \KeywordTok{max}\NormalTok{(t) }\OperatorTok{>}\StringTok{ }\KeywordTok{days}\NormalTok{(}\DecValTok{5}\NormalTok{)), by=id]}
\NormalTok{valid_ids <-}\StringTok{ }\NormalTok{valid_dt[valid }\OperatorTok{==}\StringTok{ }\NormalTok{T, id]}
\NormalTok{dt <-}\StringTok{ }\NormalTok{dt[id }\OperatorTok{%in%}\StringTok{ }\NormalTok{valid_ids]}
\KeywordTok{summary}\NormalTok{(dt)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## behavr table with:
##  52  individuals
##  9   metavariables
##  3   variables
##  1.40609e+05 measurements
##  1   key (id)
\end{verbatim}

Note that as a result, we now have 52 ``valid'' individuals.

\hypertarget{double-plotted-actograms}{%
\subsection{Double plotted actograms}\label{double-plotted-actograms}}

A common way of representing rythmicity in circadian experiments is to
compute ``double-plotted actograms''. In Fig S1A, we show all double
plotted actograms layed out in a grid.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{figS1A <-}\StringTok{ }\KeywordTok{ggetho}\NormalTok{(dt, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{z =}\NormalTok{ moving), }\DataTypeTok{multiplot =} \DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{            }\KeywordTok{stat_bar_tile_etho}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{            }\KeywordTok{facet_wrap}\NormalTok{( }\OperatorTok{~}\StringTok{ }\NormalTok{label, }\DataTypeTok{ncol=}\DecValTok{4}\NormalTok{) }\OperatorTok{+}
\StringTok{            }\KeywordTok{scale_y_discrete}\NormalTok{(}\DataTypeTok{name=}\StringTok{"Day"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{periodograms}{%
\subsection{Periodograms}\label{periodograms}}

For each indidual, we compute a \(\chi{}^2\) periodogram and we layout
all of them in a grid (Fig S1B).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt_ll <-}\StringTok{ }\NormalTok{dt[t }\OperatorTok{>}\StringTok{ }\KeywordTok{days}\NormalTok{(}\DecValTok{1}\NormalTok{)]}
\NormalTok{per_dt <-}\StringTok{ }\KeywordTok{periodogram}\NormalTok{(moving, }
\NormalTok{                        dt_ll, }
                        \DataTypeTok{resample_rate =} \DecValTok{1}\OperatorTok{/}\KeywordTok{mins}\NormalTok{(}\DecValTok{10}\NormalTok{),}
                        \DataTypeTok{FUN=}\NormalTok{chi_sq_periodogram)}

\NormalTok{per_dt <-}\StringTok{ }\KeywordTok{find_peaks}\NormalTok{(per_dt)}

\NormalTok{figS1B <-}\StringTok{ }\KeywordTok{ggperio}\NormalTok{(per_dt, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ power, }\DataTypeTok{peak=}\NormalTok{peak)) }\OperatorTok{+}\StringTok{ }
\StringTok{      }\KeywordTok{geom_line}\NormalTok{() }\OperatorTok{+}
\StringTok{      }\KeywordTok{geom_line}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{y=}\NormalTok{signif_threshold), }\DataTypeTok{colour=}\StringTok{"red"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{      }\KeywordTok{geom_peak}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{      }\KeywordTok{facet_wrap}\NormalTok{( }\OperatorTok{~}\StringTok{ }\NormalTok{label, }\DataTypeTok{ncol=}\DecValTok{4}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\end{document}

### 2018-02-08 plos-latex-template/results/scipt.R

@@ -1,39 +1,0 @@
-rm(list=ls())
-library(ggetho)
-library(damr)
-library(zeitgebr)
-
-met <- damr::link_dam2_metadata("./metadata.csv", ".")
-dt <- load_dam2(met)
-
-summary(dt)
-dt[,t := t - days(xmv(baseline_days))]
-dt[, moving :=  activity > 0]
-dt[, label := interaction(1:.N, genotype), meta=T]
-dt
-
-ggetho(dt, aes(y=label, z=moving)) + 
-  stat_tile_etho() + 
-  stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
-  stat_ld_annotations(x_limits = c(0, dt[,max(t)]), ld_colours = c("white", "grey"))
-
-# remove dead animals here
-
-ggetho(dt, aes(z=moving), multiplot=2) + 
-  stat_tile_etho() +
-  facet_wrap(~label) 
-
-# from the overview, we decide to 
-dt_ll <- dt[t > days(1)]
-
-ggetho(dt_ll, aes(z=moving/2), multiplot=2) + 
-  stat_tile_etho() +
-  facet_wrap(~label) 
-
-per_dt <- periodogram(moving, 
-						dt_ll, 
-						FUN=chi_sq_periodogram, 
-						period_range = c(hours(15), hours(32)))
-per_dt <- find_peaks(per_dt)
-
-ggperio(per_dt, aes(y = power, peak=peak)) + geom_line() + geom_peak() + facet_wrap(~label) 

### 2018-02-08 plos-latex-template/manuscript.Rnw
include results in manuscript and sort things around a bit

--- 2018-02-02 plos-latex-template/manuscript.tex
@@
 
-\section*{Results}
-We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
-A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
-In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
-Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
-Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
-Two replicates were performed at different times.
-raw data and metadata files are publicly available TODO cite zenodo
+<<'child-results.Rnw', child='results.Rnw'>>=
+@
 
-\subsection*{Data loading}
-
-\subsection*{Quality control}
-
-\subsection*{Actograms}
-
-\subsection*{Spectogram}
+% We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
+% A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
+% In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
+% Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
+% Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
+% Two replicates were performed at different times.
+% raw data and metadata files are publicly available TODO cite zenodo
+%
 

### 2018-02-08 plos-latex-template/results.Rnw
include results in manuscript and sort things around a bit

<<echo=FALSE, message=FALSE>>=
opts_knit$set(root.dir = "results")
@

\section*{Results}

<<message=FALSE>>=
library(damr)      # input DAM2 data
library(zeitgebr)  # periodogram computation
library(sleepr)    # sleep analysis
library(ggetho)    # behaviour visualisation
@

Then, the metadata file is read and linked to the \texttt{.txt} result files.

<<>>=
metadata <- link_dam2_metadata("metadata.csv",".")    # linking
# print(metadata)                                     # check metadata
dt <- load_dam2(metadata)                             # loading
summary(dt)                                           # quick summary
@

\subsection*{Preprocessing}
We notice, from the metadata, that the two replicates do not have the same time in baseline.
We would like to express the time relative to the important event: the transition to \texttt{LL}. 
To do so, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:

<<message=F, results="hide">>=
# baseline subtraction -- note the use of xmv
dt[,t := t - days(xmv(baseline_days))]
dt[, moving :=  activity > 0]  
summary(dt)
@

To simplify visualisation, we create our own \texttt{label} metavariable, as combination of a number and \texttt{genotype}.
In the restricted context of this analysis, \texttt{label} acts a unique identifier.
Importantly, we keep \texttt{id} , which is more rigorous and universal.

<<message=FALSE, rsults="hide">>=
dt[, label := interaction(1:.N, genotype), meta=T]
# print(dt)
@

\subsection*{Curation}
It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:tile-plot}A).

<<>>=
# make a ggplot object with label on the y and moving on the z axis
fig3A <- ggetho(dt, aes(y=label, z=moving)) +  
  # show data as a tile plot
  # that is z is a pixel whose intensity maps moving
  stat_tile_etho() +                  
  # add layers to draw annotations to show L and D phases
  # as white and black, respectivelly
  # the first layer is for the baseline (until t=0)
  stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
  # in the 2nd one, we start at 0 and use grey 
  # instead of black as we work in LL
  stat_ld_annotations(x_limits = c(0, dt[,max(t)]), 
                      ld_colours = c("white", "grey"))
@

Activity of dead or escaped animals is falsely scored as long series of zeros.
Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data:

<<>>=
# remove data after death
dt <- sleepr::curate_dead_animals(dt, moving)
# same as above
fig3B <- ggetho(dt, aes(y=label, z=moving)) +  
    stat_tile_etho() +                  
    stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
    stat_ld_annotations(x_limits = c(0, dt[,max(t)]), 
                        ld_colours = c("white", "grey"))
@
The updated version can be visualised in Fig 3B.

For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
<<>>=
# for each id, we check for validity
valid_dt <- dt[ , .(valid = max(t) > days(5)), by=id]
# a vector of all valid ids
valid_ids <- valid_dt[valid == T, id]
# filter dt with the valid ids
dt <- dt[id %in% valid_ids]
summary(dt)
@

Note that as a result, we now have \Sexpr{nrow(dt[meta=T])} ``valid''individuals.

\subsection*{Double plotted actograms}
``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
In Fig~S1A, we show the double-plotted actograms of each animals:

<<>>=
figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
            stat_bar_tile_etho() + 
            facet_wrap( ~ label, ncol=4) +
            scale_y_discrete(name="Day")
@

\subsection*{Periodograms}
Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (Fig~S1B).

<<>>=
dt_ll <- dt[t > days(1)]
per_dt <- periodogram(moving, 
                    	dt_ll, 
                    	resample_rate = 1/mins(10),
                    	FUN=chi_sq_periodogram)

per_dt <- find_peaks(per_dt)

figS1B <- ggperio(per_dt, aes(y = power, peak=peak)) + 
                  geom_line() +
                  geom_line(aes(y=signif_threshold), colour="red") + 
                  geom_peak() + 
                  facet_wrap( ~ label, ncol=4) 
@

<<r, eval=TRUE, echo=FALSE, message=FALSE, result="hide">>=
################################ CODE BELOW FOR COSMETIC ALTERATIONS OF FIGURES

library(cowplot)
my_theme <- theme(axis.title.y=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.line = element_blank(),
            panel.border = element_blank(),
            legend.position="none")

legend <- fig3A + theme(axis.title=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks=element_blank(),
                        axis.line = element_blank(),
                        #axis.ticks.y=element_blank(),
                        panel.border = element_blank(),
                        #legend.text = "moving",
                        legend.position="left") + 
                  coord_cartesian(xlim=-days(c(10, 9))) + scale_y_discrete(position = "right")

fig3 <- cowplot::plot_grid(
                          legend,
                          fig3A +  my_theme,
                          fig3B +  my_theme,
                          nrow=1,
                          labels=c("", "A", "B"),
                          rel_widths = c(1, 2, 2),
                          label_size = 20,
                          axis="b",
                          align="h"
                          )
pdf("../fig/fig3.pdf",w=12,h=12)
print(fig3)
dev.off()

figS1A <- figS1A +  theme(strip.text.x = element_text(size =8))
figS1B <- figS1B +  theme(strip.text.x = element_text(size =8))

figS1 <- cowplot::plot_grid(
                          figS1A,
                          figS1B,
                          nrow=1,
                          labels=c( "A", "B"),
                          label_size = 20,
                          align="v"
                          )

pdf("../fig/figS1.pdf",w=16,h=16)
print(figS1)
dev.off()
@

### 2018-02-08 plos-latex-template/manuscript.tex
result script
cleaner manuscript

@@
 
-\documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 %% END MACROS SECTION
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:framework}).
+\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:fig-1}).
 This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
@@
 \begin{figure}[!h]
-    \includegraphics[width=1\textwidth]{fig/framework.pdf}
+    \includegraphics[width=1\textwidth]{fig/fig-1.pdf}
 	\caption{{\bf The \texttt{rethomics} workflow.}
 		Diagram representing, from left to right, the interplay between the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
-	\label{fig:framework}
+	\label{fig:fig-1}
 \end{figure}
@@
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:behavr}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
+We created \texttt{behavr} (Fig~\ref{fig:fig-2}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
 two challenges that are inherent to manipulating ethomics results.
@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/behavr.pdf}
+	\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
 	\caption{{\bf Example of a \texttt{behavr} table.}
 		The metadata holds a single row for each of the $n$ individuals. 
-		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to map the data) or user-defined (i.e. arbitrary).
+		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to map the data) or user-defined (\emph{i.e.} arbitrary).
 		In the data, each row is a ``read'' (\emph{i.e.} information about one animal at one time point).
@@
 	}
-	\label{fig:behavr}
+	\label{fig:fig2}
 \end{figure}
@@
 \section*{Results}
-We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
-A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
-In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
-Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
-Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
-Two replicates were performed at different times.
-raw data and metadata files are publicly available TODO cite zenodo
 
-\subsection*{Data loading}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlkwd{library}\hlstd{(damr)}      \hlcom{# input DAM2 data}
+\hlkwd{library}\hlstd{(zeitgebr)}  \hlcom{# periodogram computation}
+\hlkwd{library}\hlstd{(sleepr)}    \hlcom{# sleep analysis}
+\hlkwd{library}\hlstd{(ggetho)}    \hlcom{# behaviour visualisation}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
 
-\subsection*{Quality control}
+Then, the metadata file is read and linked to the \texttt{.txt} result files.
 
-\subsection*{Actograms}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam2_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,}\hlstr{"."}\hlstd{)}    \hlcom{# linking}
+\hlcom{# print(metadata)                                     # check metadata}
+\hlstd{dt} \hlkwb{<-} \hlkwd{load_dam2}\hlstd{(metadata)}                             \hlcom{# loading}
+\hlkwd{summary}\hlstd{(dt)}                                           \hlcom{# quick summary}
+\end{alltt}
+\begin{verbatim}
+## behavr table with:
+##  58	individuals
+##  8	metavariables
+##  2	variables
+##  1.58722e+05	measurements
+##  1	key (id)
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
 
-\subsection*{Spectogram}
+\subsection*{Preprocessing}
+We notice, from the metadata, that the two replicates do not have the same time in baseline.
+We would like to express the time relative to the important event: the transition to \texttt{LL}. 
+To do so, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
+This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
+In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
+It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# baseline subtraction -- note the use of xmv}
+\hlstd{dt[,t} \hlkwb{:=} \hlstd{t} \hlopt{-} \hlkwd{days}\hlstd{(}\hlkwd{xmv}\hlstd{(baseline_days))]}
+\hlstd{dt[, moving} \hlkwb{:=}  \hlstd{activity} \hlopt{>} \hlnum{0}\hlstd{]}
+\hlkwd{summary}\hlstd{(dt)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+To simplify visualisation, we create our own \texttt{label} metavariable, as combination of a number and \texttt{genotype}.
+In the restricted context of this analysis, \texttt{label} acts a unique identifier.
+Importantly, we keep \texttt{id} , which is more rigorous and universal.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{dt[, label} \hlkwb{:=} \hlkwd{interaction}\hlstd{(}\hlnum{1}\hlopt{:}\hlstd{.N, genotype),} \hlkwc{meta}\hlstd{=T]}
+\hlcom{# print(dt)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Curation}
+It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:tile-plot}A).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# make a ggplot object with label on the y and moving on the z axis}
+\hlstd{fig3A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=label,} \hlkwc{z}\hlstd{=moving))} \hlopt{+}
+  \hlcom{# show data as a tile plot}
+  \hlcom{# that is z is a pixel whose intensity maps moving}
+  \hlkwd{stat_tile_etho}\hlstd{()} \hlopt{+}
+  \hlcom{# add layers to draw annotations to show L and D phases}
+  \hlcom{# as white and black, respectivelly}
+  \hlcom{# the first layer is for the baseline (until t=0)}
+  \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,}\hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
+  \hlcom{# in the 2nd one, we start at 0 and use grey }
+  \hlcom{# instead of black as we work in LL}
+  \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,}\hlkwd{max}\hlstd{(t)]),}
+                      \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+Activity of dead or escaped animals is falsely scored as long series of zeros.
+Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# remove data after death}
+\hlstd{dt} \hlkwb{<-} \hlstd{sleepr}\hlopt{::}\hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
+\hlcom{# same as above}
+\hlstd{fig3B} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=label,} \hlkwc{z}\hlstd{=moving))} \hlopt{+}
+    \hlkwd{stat_tile_etho}\hlstd{()} \hlopt{+}
+    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,}\hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
+    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,}\hlkwd{max}\hlstd{(t)]),}
+                        \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+The updated version can be visualised in Fig 3B.
+
+For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# for each id, we check for validity}
+\hlstd{valid_dt} \hlkwb{<-} \hlstd{dt[ ,} \hlkwd{.}\hlstd{(}\hlkwc{valid} \hlstd{=} \hlkwd{max}\hlstd{(t)} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{5}\hlstd{)),} \hlkwc{by}\hlstd{=id]}
+\hlcom{# a vector of all valid ids}
+\hlstd{valid_ids} \hlkwb{<-} \hlstd{valid_dt[valid} \hlopt{==} \hlstd{T, id]}
+\hlcom{# filter dt with the valid ids}
+\hlstd{dt} \hlkwb{<-} \hlstd{dt[id} \hlopt{%in%} \hlstd{valid_ids]}
+\hlkwd{summary}\hlstd{(dt)}
+\end{alltt}
+\begin{verbatim}
+## behavr table with:
+##  52	individuals
+##  9	metavariables
+##  3	variables
+##  1.40609e+05	measurements
+##  1	key (id)
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Note that as a result, we now have 52 ``valid''individuals.
+
+\subsection*{Double plotted actograms}
+``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
+In Fig~S1A, we show the double-plotted actograms of each animals:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
+            \hlkwd{stat_bar_tile_etho}\hlstd{()} \hlopt{+}
+            \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol}\hlstd{=}\hlnum{4}\hlstd{)} \hlopt{+}
+            \hlkwd{scale_y_discrete}\hlstd{(}\hlkwc{name}\hlstd{=}\hlstr{"Day"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\subsection*{Periodograms}
+Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
+In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (see \nameref{S1-Fig}).
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{dt_ll} \hlkwb{<-} \hlstd{dt[t} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{1}\hlstd{)]}
+\hlstd{per_dt} \hlkwb{<-} \hlkwd{periodogram}\hlstd{(moving,}
+                        \hlstd{dt_ll,}
+                        \hlkwc{resample_rate} \hlstd{=} \hlnum{1}\hlopt{/}\hlkwd{mins}\hlstd{(}\hlnum{10}\hlstd{),}
+                        \hlkwc{FUN}\hlstd{=chi_sq_periodogram)}
+
+\hlstd{per_dt} \hlkwb{<-} \hlkwd{find_peaks}\hlstd{(per_dt)}
+
+\hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak}\hlstd{=peak))} \hlopt{+}
+                  \hlkwd{geom_line}\hlstd{()} \hlopt{+}
+                  \hlkwd{geom_line}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=signif_threshold),} \hlkwc{colour}\hlstd{=}\hlstr{"red"}\hlstd{)} \hlopt{+}
+                  \hlkwd{geom_peak}\hlstd{()} \hlopt{+}
+                  \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol}\hlstd{=}\hlnum{4}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## pdf 
+##   2
+## pdf 
+##   2
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+% We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 
+% A less formal description of this case and others are explained at \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
+% In order to prodive a more comprehensive and didactic example, data was altered and simplified.	
+% Fig~\ref{fig:experiment} describes our case experiment (A) and the corresponding metadata (B).
+% Briefly, three monitors were used each containing 16 males and 16 females of a different genotype (A, B and C).
+% Two replicates were performed at different times.
+% raw data and metadata files are publicly available TODO cite zenodo
+%
 
 \section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at  		\href{https://github.com/rethomics}{https://github.com/rethomics/}.
+All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at
+\href{https://github.com/rethomics}{https://github.com/rethomics/}.
 Extensive installation instructions as well as reproducible demos and tutorials are available at
@@
 * ...
+
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1-Fig}
+{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
 

### 2018-02-08 plos-latex-template/results.Rnw
cleaner manuscript

@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (Fig~S1B).
+In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (see \nameref{S1-Fig}).
 
@@
                           )
-pdf("../fig/fig3.pdf",w=12,h=12)
+pdf("../fig/fig-3.pdf",w=12,h=12)
 print(fig3)
@@
 
-pdf("../fig/figS1.pdf",w=16,h=16)
+pdf("../fig/fig-S1.pdf",w=12,h=16)
 print(figS1)

### 2018-02-08 plos-latex-template/manuscript.Rnw
another push

@@
 Behaviour scoring is certainly not an exception to this trend.
-Indeed, many platforms (todo citations) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
+Indeed, many platforms (TODO citations) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
 
@@
 	}
-	\label{fig:behavr}
+	\label{fig:fig-2}
 \end{figure}
@@
 \section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at  		\href{https://github.com/rethomics}{https://github.com/rethomics/}.
+All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at
+\href{https://github.com/rethomics}{https://github.com/rethomics/}.
 Extensive installation instructions as well as reproducible demos and tutorials are available at

### 2018-02-08 plos-latex-template/manuscript.tex
another push

@@
 	}
-	\label{fig:fig2}
+	\label{fig:fig-2}
 \end{figure}
@@
 \subsection*{Curation}
-It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:tile-plot}A).
+It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
+
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
+	\caption{{\bf Experiment quality control.}
+			Tile plot representing the fraction of time spent moving as a colour intensity.
+			Each individual is represented by a row and time, on the x axis, is binned in 30 minute.}
+	\label{fig:fig-3}
+\end{figure}
 
@@
 Activity of dead or escaped animals is falsely scored as long series of zeros.
-Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data:
-
+Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data.
+The updated version can be visualised in Fig~\ref{fig:fig-3}B.
 \begin{knitrout}
@@
 \end{knitrout}
-The updated version can be visualised in Fig 3B.
 

### 2018-02-08 plos-latex-template/results.Rnw
another push

@@
 Activity of dead or escaped animals is falsely scored as long series of zeros.
-Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data:
-
+Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data.
+The updated version can be visualised in Fig~\ref{fig:fig-3}B.
 <<>>=
@@
 @
-The updated version can be visualised in Fig 3B.
 

### 2018-02-09 plos-latex-template/manuscript.Rnw
cache chunks for faster writing

@@
 provides biologists with an unprecedented tools to decipher these complex phenotypes. 
-Analysing ethomics data comes with many challenges that are, conceptually, independent of the acquisition platform used.
-However, there is, little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
+Analysing ethomics data comes with many challenges that are independent of these acquisition platform.
+However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
 We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offer utilities to:

### 2018-02-09 plos-latex-template/manuscript.tex
cache chunks for faster writing

@@
 \hlstd{dt[, label} \hlkwb{:=} \hlkwd{interaction}\hlstd{(}\hlnum{1}\hlopt{:}\hlstd{.N, genotype),} \hlkwc{meta}\hlstd{=T]}
-\hlcom{# print(dt)}
+\hlkwd{print}\hlstd{(dt)}
 \end{alltt}
@@
 The updated version can be visualised in Fig~\ref{fig:fig-3}B.
+
 \begin{knitrout}
@@
 \end{alltt}
-\end{kframe}
-\end{knitrout}
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-## pdf 
-##   2
-## pdf 
-##   2
-\end{verbatim}
 \end{kframe}

### 2018-02-09 plos-latex-template/results.Rnw
cache chunks for faster writing

@@
-<<echo=FALSE, message=FALSE>>=
+<<echo=FALSE, message=FALSE, cache=TRUE>>=
 opts_knit$set(root.dir = "results")
@@
 
-<<>>=
+<<cache=TRUE>>=
 metadata <- link_dam2_metadata("metadata.csv",".")    # linking
@@
 
-<<message=F, results="hide">>=
+<<message=F, results="hide", cache=TRUE>>=
 # baseline subtraction -- note the use of xmv
@@
 
-<<message=FALSE, rsults="hide">>=
+<<message=FALSE, results="hide", cache=TRUE>>=
 dt[, label := interaction(1:.N, genotype), meta=T]
-# print(dt)
+print(dt)
 @
@@
 
-<<>>=
+<<cache=TRUE>>=
 # make a ggplot object with label on the y and moving on the z axis
@@
 The updated version can be visualised in Fig~\ref{fig:fig-3}B.
-<<>>=
+
+<<cache=TRUE>>=
 # remove data after death
@@
 For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
-<<>>=
+<<cache=TRUE>>=
 # for each id, we check for validity
@@
 
-<<>>=
+<<cache=TRUE>>=
 figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
@@
 
-<<>>=
+<<cache=TRUE>>=
 dt_ll <- dt[t > days(1)]
@@
 
-<<r, eval=TRUE, echo=FALSE, message=FALSE, result="hide">>=
+<<r, eval=TRUE, echo=FALSE, message=FALSE, result="hide",cache=TRUE>>=
 ################################ CODE BELOW FOR COSMETIC ALTERATIONS OF FIGURES
@@
 print(fig3)
-dev.off()
+tmp  <- dev.off()
 
@@
 print(figS1)
-dev.off()
+tmp <- dev.off()
 @

### 2018-02-09 plos-latex-template/manuscript.Rnw
draft almost there!

@@
 % brings the promise of a powerful and original insight
-provides biologists with an unprecedented tools to decipher these complex phenotypes. 
-Analysing ethomics data comes with many challenges that are independent of these acquisition platform.
+provides biologists with an unprecedented set of tools to decipher these complex phenotypes. 
+Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
 However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
@@
 The availability of large amounts of data is very exciting as it paves the way for in-depth analyses.
-Clearly, the multiplicity of model organisms, hypotheses and paradigms should be reflected in  diverse range of recording tools.
+Clearly, the multiplicity of model organisms, hypotheses and paradigms should be matched by a diverse range of recording tools.
 However, when it comes to the subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
@@
 Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
-Lastly, it links analysis and visualisation to the target tool, which makes it very difficult to share cross-tool utilities and concepts.
+Lastly, it links analysis and visualisation to the target acquisition tool, which makes it very difficult to share cross-tool utilities and concepts.
 
@@
 It offers an elegant computational solution to store, manipulate and visualise a large amount of data.
-We expect it to fill the gap between behavioural biology and data sciences.
+We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural biologists.
 \texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical tutorials.
@@
 are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store both large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object.
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
@@
 		The metadata holds a single row for each of the $n$ individuals. 
-		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to map the data) or user-defined (\emph{i.e.} arbitrary).
-		In the data, each row is a ``read'' (\emph{i.e.} information about one animal at one time point).
+		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to fetch the data) or user-defined (\emph{i.e.} arbitrary).
+		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time point).
 		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual.
 		Data and metadata are implicitly joined with the \texttt{id} field.
-		Notably, the names used for variables and metavariable are only plausible examples that will likely differ in practice. 
+		In this figure, the names used for variables and metavariable are only plausible examples that will likely differ in practice. 
 	}
@@
 
-In this respect the metadata file is a standardised an comprehensive data frame describing an experiment.
+In this respect, the metadata file is a standardised an comprehensive data frame describing an experiment.
 Using such a structure comes with multiple advantages.
 For instance, it simplify collaboration and data exchange as all treatments and individuals are very explicit.
-Then, it promotes good experimental practices such as interspersion of treatments (indeed, without it, users are tempted to simplify their design, for instance, confounding device/location and treatment).
+Then, it promotes good experimental practices such as interspersion of treatments (indeed, without it, users are tempted to simplify their design by, for instance, confounding device/location and treatment).
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
-Indeed, additional replicates can simply be added as new rows, and replicate number later user, if needed, as a covariate.
+Indeed, additional replicates can simply be added as new rows and replicate ID later used, if needed, as a covariate.
 
@@
 It particularly applies to the visualisation of long experiments, with the ability to, for instance, annotate light and dark phases, wrap time over a circadian day, display ``double-plotted actograms'' and periodograms.
-Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot}.
+Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot2}.
 

### 2018-02-09 plos-latex-template/manuscript.tex
draft almost there!

@@
 
+TODO description of the dataset here
+
+* experiment goal
+
+* design
+
+* source
+
+* express the idea that shis is just simple example that could be scaled up
+
+First of all, we load the necessary libraries (see availability section for installation instructions):
 \begin{knitrout}
@@
 \hlstd{dt[, moving} \hlkwb{:=}  \hlstd{activity} \hlopt{>} \hlnum{0}\hlstd{]}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
 \hlkwd{summary}\hlstd{(dt)}
 \end{alltt}
+\begin{verbatim}
+## behavr table with:
+##  58	individuals
+##  8	metavariables
+##  3	variables
+##  1.58722e+05	measurements
+##  1	key (id)
+\end{verbatim}
 \end{kframe}
@@
 \hlcom{# make a ggplot object with label on the y and moving on the z axis}
-\hlstd{fig3A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=label,} \hlkwc{z}\hlstd{=moving))} \hlopt{+}
+\hlstd{fig3A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= label,} \hlkwc{z} \hlstd{= moving))} \hlopt{+}
   \hlcom{# show data as a tile plot}
@@
   \hlcom{# as white and black, respectivelly}
-  \hlcom{# the first layer is for the baseline (until t=0)}
+  \hlcom{# the first layer is for the baseline (until t = 0)}
   \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,}\hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
@@
   \hlcom{# instead of black as we work in LL}
-  \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,}\hlkwd{max}\hlstd{(t)]),}
+  \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,} \hlkwd{max}\hlstd{(t)]),}
                       \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
@@
 
-Activity of dead or escaped animals is falsely scored as long series of zeros.
+Activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
 Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data.
@@
 \hlcom{# same as above}
-\hlstd{fig3B} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=label,} \hlkwc{z}\hlstd{=moving))} \hlopt{+}
+\hlstd{fig3B} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= label,} \hlkwc{z} \hlstd{= moving))} \hlopt{+}
     \hlkwd{stat_tile_etho}\hlstd{()} \hlopt{+}
-    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,}\hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
-    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,}\hlkwd{max}\hlstd{(t)]),}
+    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(dt[,} \hlkwd{min}\hlstd{(t)],} \hlnum{0}\hlstd{))} \hlopt{+}
+    \hlkwd{stat_ld_annotations}\hlstd{(}\hlkwc{x_limits} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{0}\hlstd{, dt[,} \hlkwd{max}\hlstd{(t)]),}
                         \hlkwc{ld_colours} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"white"}\hlstd{,} \hlstr{"grey"}\hlstd{))}
@@
 \hlcom{# for each id, we check for validity}
-\hlstd{valid_dt} \hlkwb{<-} \hlstd{dt[ ,} \hlkwd{.}\hlstd{(}\hlkwc{valid} \hlstd{=} \hlkwd{max}\hlstd{(t)} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{5}\hlstd{)),} \hlkwc{by}\hlstd{=id]}
+\hlstd{valid_dt} \hlkwb{<-} \hlstd{dt[,} \hlkwd{.}\hlstd{(}\hlkwc{valid} \hlstd{=} \hlkwd{max}\hlstd{(t)} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{5}\hlstd{)),} \hlkwc{by} \hlstd{= id]}
 \hlcom{# a vector of all valid ids}
@@
 ``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In Fig~S1A, we show the double-plotted actograms of each animals:
+In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals:
 
@@
 \hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
+            \hlcom{# bars n the z axis could }
+            \hlcom{#  one could also use stat_tile_etho}
             \hlkwd{stat_bar_tile_etho}\hlstd{()} \hlopt{+}
-            \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol}\hlstd{=}\hlnum{4}\hlstd{)} \hlopt{+}
+            \hlcom{# divide plot by individual}
+            \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol} \hlstd{=} \hlnum{4}\hlstd{)} \hlopt{+}
+            \hlcom{# rename the y axis}
             \hlkwd{scale_y_discrete}\hlstd{(}\hlkwc{name}\hlstd{=}\hlstr{"Day"}\hlstd{)}
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (see \nameref{S1-Fig}).
+In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (see \nameref{S1-Fig}B).
 
@@
 \hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak}\hlstd{=peak))} \hlopt{+}
+                  \hlcom{# periododogram drawn as a line}
                   \hlkwd{geom_line}\hlstd{()} \hlopt{+}
-                  \hlkwd{geom_line}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{y}\hlstd{=signif_threshold),} \hlkwc{colour}\hlstd{=}\hlstr{"red"}\hlstd{)} \hlopt{+}
+                  \hlcom{# the significance line in red}
+                  \hlkwd{geom_line}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= signif_threshold),} \hlkwc{colour}\hlstd{=}\hlstr{"red"}\hlstd{)} \hlopt{+}
+                  \hlcom{# point and text at the peak}
                   \hlkwd{geom_peak}\hlstd{()} \hlopt{+}
-                  \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol}\hlstd{=}\hlnum{4}\hlstd{)}
+                  \hlcom{# divide plot by individual}
+                  \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol} \hlstd{=} \hlnum{4}\hlstd{)}
 \end{alltt}
@@
 \end{knitrout}
+
+\subsection*{Population statistics}
+Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
+arhythmic in LL whilst Tim/NKCCOX appear to have a consistant, long-period rhythm.
+To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-4}A):
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{fig4A} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power} \hlopt{-} \hlstd{signif_threshold,}
+                             \hlkwc{colour} \hlstd{= genotype))} \hlopt{+}
+          \hlcom{# periododogram a line for population mean}
+          \hlcom{# and bootstrap error bars}
+          \hlkwd{stat_pop_etho}\hlstd{(}\hlkwc{method} \hlstd{= ggplot2}\hlopt{::}\hlstd{mean_cl_boot)} \hlopt{+}
+          \hlcom{# rename x and y axis }
+          \hlkwd{scale_y_continuous}\hlstd{(}\hlkwc{name}\hlstd{=}\hlstr{"Relative power"}\hlstd{)} \hlopt{+}
+          \hlkwd{scale_x_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-4}B). First of all, we comput a summary per individual (\texttt{by=id}):
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{summary_dt} \hlkwb{<-} \hlstd{per_dt[,}\hlkwd{.}\hlstd{(}
+                        \hlkwc{first_peak_period} \hlstd{= period[peak}\hlopt{==}\hlnum{1}\hlstd{],}
+                        \hlkwc{first_peak_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{],}
+                        \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak}\hlopt{==}\hlnum{1}\hlstd{)),}
+                      \hlkwc{by}\hlstd{=id]}
+
+\hlcom{# rejoin metadata}
+\hlstd{summary_dt} \hlkwb{<-} \hlkwd{rejoin}\hlstd{(summary_dt)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\texttt{summary\_dt} is just a regular data frame with one row per individual, contianing both metadata and our summary statistics. It can therefore be used directly by \texttt{ggplot} and such:
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# standard ggplot}
+\hlstd{fig4B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
+                                \hlkwc{x} \hlstd{= genotype))} \hlopt{+}
+              \hlcom{# draw the mean of each genotype group}
+              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.y} \hlstd{= mean,} \hlkwc{geom} \hlstd{=} \hlstr{"point"}\hlstd{,} \hlkwc{shape}\hlstd{=}\hlnum{3}\hlstd{)} \hlopt{+}
+              \hlcom{# draw bootstrap confidence intervals}
+              \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.data} \hlstd{= mean_cl_boot,} \hlkwc{geom} \hlstd{=} \hlstr{"errorbar"}\hlstd{)} \hlopt{+}
+              \hlcom{# shows all individuals as points. }
+              \hlcom{# the size of the point expresses the power of the peak}
+              \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour}\hlstd{= genotype,}
+                              \hlkwc{size}\hlstd{=first_peak_power),}
+                          \hlkwc{alpha}\hlstd{=}\hlnum{.67}\hlstd{)} \hlopt{+}
+              \hlcom{# We would like to convert time in hour}
+              \hlkwd{scale_y_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhytmic flies (see Fig~\ref{fig:fig-4}C):
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# standard ggplot}
+\hlstd{fig4C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
+                                \hlkwc{fill} \hlstd{= genotype,}
+                                \hlkwc{alpha} \hlstd{= is_rhythmic}
+                                \hlstd{))} \hlopt{+}
+              \hlkwd{geom_bar}\hlstd{(}\hlkwc{colour}\hlstd{=}\hlstr{"black"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\texttt{R} provide one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
+One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic vs arhythmic flies in both genotypes.
+To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of ($p$-value $< 10^{-5}$):
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family}\hlstd{=}\hlstr{"binomial"}\hlstd{)}
+
+\hlkwd{summary}\hlstd{(fit)}\hlopt{$}\hlstd{coefficients}
+\end{alltt}
+\begin{verbatim}
+##                     Estimate Std. Error   z value     Pr(>|z|)
+## (Intercept)        -1.504077  0.5527708 -2.720978 6.508902e-03
+## genotypeTim/NKCCOX  4.143135  0.9172057  4.517127 6.268433e-06
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{result_dt} \hlkwb{<-} \hlstd{summary_dt[,}
+          \hlkwd{.}\hlstd{(}
+            \hlkwc{mean_period} \hlstd{=} \hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
+            \hlkwc{sd_period} \hlstd{=} \hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
+            \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
+            \hlkwc{n} \hlstd{= .N}
+            \hlstd{),}
+          \hlkwc{by} \hlstd{= genotype}
+          \hlstd{]}
+\hlstd{result_dt}
+\end{alltt}
+\begin{verbatim}
+##      genotype mean_period sd_period n_rhythmic  n
+## 1:   NKCCOX/+    25.92500  4.133098          4 22
+## 2: Tim/NKCCOX    26.36071  1.834513         28 30
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\caption{{\bf Population statistics on circadian phenotype.}
+			TODO}
+	\label{fig:fig-4}
+\end{figure}
+
+* high scalability
+
+* reproducibility
+
+* from raw data to publication-quality figures
+
+* flexibility
 

### 2018-02-09 plos-latex-template/results.Rnw
draft almost there!

@@
-<<echo=FALSE, message=FALSE, cache=TRUE>>=
+<<echo=FALSE, message=FALSE>>=
+library(knitr)
 opts_knit$set(root.dir = "results")
+setwd("results")
 @
@@
 
+TODO description of the dataset here
+
+* experiment goal
+
+* design
+
+* source
+
+* express the idea that shis is just simple example that could be scaled up
+
+First of all, we load the necessary libraries (see availability section for installation instructions):
 <<message=FALSE>>=
@@
 dt[, moving :=  activity > 0]  
+@
+
+<<cache=TRUE>>=
 summary(dt)
@@
 # make a ggplot object with label on the y and moving on the z axis
-fig3A <- ggetho(dt, aes(y=label, z=moving)) +  
+fig3A <- ggetho(dt, aes(y = label, z = moving)) +  
   # show data as a tile plot
@@
   # as white and black, respectivelly
-  # the first layer is for the baseline (until t=0)
+  # the first layer is for the baseline (until t = 0)
   stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
@@
   # instead of black as we work in LL
-  stat_ld_annotations(x_limits = c(0, dt[,max(t)]), 
+  stat_ld_annotations(x_limits = c(0, dt[, max(t)]), 
                       ld_colours = c("white", "grey"))
@@
 # same as above
-fig3B <- ggetho(dt, aes(y=label, z=moving)) +  
+fig3B <- ggetho(dt, aes(y = label, z = moving)) +  
     stat_tile_etho() +                  
-    stat_ld_annotations(x_limits = c(dt[,min(t)], 0)) +
-    stat_ld_annotations(x_limits = c(0, dt[,max(t)]), 
+    stat_ld_annotations(x_limits = c(dt[, min(t)], 0)) +
+    stat_ld_annotations(x_limits = c(0, dt[, max(t)]), 
                         ld_colours = c("white", "grey"))
@@
 # for each id, we check for validity
-valid_dt <- dt[ , .(valid = max(t) > days(5)), by=id]
+valid_dt <- dt[, .(valid = max(t) > days(5)), by = id]
 # a vector of all valid ids
@@
 figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
+            # bars n the z axis could 
+            #  one could also use stat_tile_etho
             stat_bar_tile_etho() + 
-            facet_wrap( ~ label, ncol=4) +
+            # divide plot by individual
+            facet_wrap( ~ label, ncol = 4) +
+            # rename the y axis
             scale_y_discrete(name="Day")
@@
 figS1B <- ggperio(per_dt, aes(y = power, peak=peak)) + 
+                  # periododogram drawn as a line
                   geom_line() +
-                  geom_line(aes(y=signif_threshold), colour="red") + 
+                  # the significance line in red
+                  geom_line(aes(y = signif_threshold), colour="red") + 
+                  # point and text at the peak
                   geom_peak() + 
-                  facet_wrap( ~ label, ncol=4) 
+                  # divide plot by individual
+                  facet_wrap( ~ label, ncol = 4) 
 @
 
-<<r, eval=TRUE, echo=FALSE, message=FALSE, result="hide",cache=TRUE>>=
+\subsection*{Population statistics}
+Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
+arhythmic in LL whilst Tim/NKCCOX appear to have a consistant, long-period rhythm.
+To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-4}A):
+
+<<cache=TRUE, warning=FALSE,message=FALSE, result="hide">>=
+fig4A <- ggperio(per_dt, aes(y = power - signif_threshold,
+                             colour = genotype)) + 
+          # periododogram a line for population mean
+          # and bootstrap error bars
+          stat_pop_etho(method = ggplot2::mean_cl_boot) +
+          # rename x and y axis 
+          scale_y_continuous(name="Relative power") + 
+          scale_x_hours("Period")
+@
+
+To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-4}B). First of all, we comput a summary per individual (\texttt{by=id}):
+
+<<cache=TRUE, warning=FALSE, results="hide">>=
+
+summary_dt <- per_dt[,.(
+                        first_peak_period = period[peak==1],
+                        first_peak_power = power[peak==1],
+                        is_rhythmic = any(peak==1)),
+                      by=id]
+
+# rejoin metadata
+summary_dt <- rejoin(summary_dt)
+@
+
+\texttt{summary\_dt} is just a regular data frame with one row per individual, contianing both metadata and our summary statistics. It can therefore be used directly by \texttt{ggplot} and such:
+
+<<>>=
+# standard ggplot
+fig4B <- ggplot(summary_dt, aes(y = first_peak_period,
+                                x = genotype)) + 
+              # draw the mean of each genotype group
+              stat_summary(fun.y = mean, geom = "point", shape=3) +
+              # draw bootstrap confidence intervals
+              stat_summary(fun.data = mean_cl_boot, geom = "errorbar") +
+              # shows all individuals as points. 
+              # the size of the point expresses the power of the peak
+              geom_jitter(aes(colour= genotype, 
+                              size=first_peak_power),
+                          alpha=.67) +
+              # We would like to convert time in hour
+              scale_y_hours("Period")
+@
+
+To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhytmic flies (see Fig~\ref{fig:fig-4}C):
+<<cache=TRUE, warning=FALSE, results="hide">>=
+# standard ggplot
+fig4C <- ggplot(summary_dt, aes(x = genotype,
+                                fill = genotype,
+                                alpha = is_rhythmic
+                                )) +
+              geom_bar(colour="black")
+
+@
+
+\texttt{R} provide one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
+One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic vs arhythmic flies in both genotypes.
+To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of ($p$-value $< 10^{-5}$):
+<<>>=
+fit <- glm(is_rhythmic ~ genotype, summary_dt, family="binomial")
+
+summary(fit)$coefficients
+@
+
+Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
+<<warning=FALSE>>=
+result_dt <- summary_dt[,
+          .(
+            mean_period = mean(first_peak_period, na.rm = T) / hours(1),
+            sd_period = sd(first_peak_period, na.rm = T) / hours(1),
+            n_rhythmic = sum(is_rhythmic),
+            n = .N
+            ),
+          by = genotype
+          ]
+result_dt
+@
+
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\caption{{\bf Population statistics on circadian phenotype.}
+			TODO}
+	\label{fig:fig-4}
+\end{figure}
+
+* high scalability
+
+* reproducibility
+
+* from raw data to publication-quality figures
+
+* flexibility
+
+<<r, eval=TRUE, echo=FALSE, message=FALSE, result="hide",cache=TRUE, warning=FALSE>>=
 ################################ CODE BELOW FOR COSMETIC ALTERATIONS OF FIGURES
-
 library(cowplot)
@@
                           )
-pdf("../fig/fig-3.pdf",w=12,h=12)
+pdf("../fig/fig-3.pdf",w=10,h=10)
 print(fig3)
@@
 
+#=================================================
 figS1A <- figS1A +  theme(strip.text.x = element_text(size =8))
@@
 tmp <- dev.off()
+#=================================================
+
+fig4 <- cowplot::plot_grid(fig4A + theme(legend.position="none"),
+                           fig4B + theme(legend.position="bottom"),
+                           fig4C + theme(legend.position="none"),
+                           nrow=1,
+                           label_size = 20,
+                           align = "hv",
+                           axis="b",
+                           labels=c( "A", "B", "C")
+                           #?cowplot::plot_grid
+                           )
+pdf("../fig/fig-4.pdf",w=10,h=5)
+print(fig4)
+tmp <- dev.off()
 @

### 2018-02-09 manuscript.Rnw
cleanup repo

--- 2018-02-09 plos-latex-template/manuscript.Rnw
@@
 
-Animal behaviours are complex phenotypical manifestations of the interaction between nervous systems and their external or internal environment.
+Animal behaviours are complex phenotypical manifestations of the interaction between nervous systems and external or internal environments.
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
@@
 
-The availability of large amounts of data is very exciting as it paves the way for in-depth analyses.
+The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
 Clearly, the multiplicity of model organisms, hypotheses and paradigms should be matched by a diverse range of recording tools.
@@
 Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating these two types of information, on datasets of hundreds of individuals, each recorded for weeks, is not trivial.
+Efficiently combining and manipulating metadata and data of hundreds of individuals, each recorded for weeks, is not trivial.
 
@@
 It offers an elegant computational solution to store, manipulate and visualise a large amount of data.
-We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural biologists.
-\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical tutorials.
+We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
+\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.
 
@@
 
-\texttt{rethomics} is implemented as a collection of small packages linked to one another (Fig~\ref{fig:fig-1}).
+\texttt{rethomics} is implemented as a collection of small packages related to one another (Fig~\ref{fig:fig-1}).
 This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
@@
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
-Results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistics tools.
+Results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
 
@@
 	\caption{{\bf The \texttt{rethomics} workflow.}
-		Diagram representing, from left to right, the interplay between the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
+		Diagram representing the interplay between, from left to right, the raw data, the \texttt{rethomics} packages (in blue) and the rest of the \texttt{R} ecosystem.}
 	\label{fig:fig-1}
@@
 Secondly, a large amount of individuals are often studied (typically $n > 100$).
-Each individual ($i$) is associated with metadata, that is a set of $p$ ``metavariables'' that describe experimental conditions.
+Each individual ($i$) is associated with metadata: a set of $p$ ``metavariables'' that describe experimental conditions.
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
@@
 		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time point).
-		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual.
+		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual $i$.
 		Data and metadata are implicitly joined with the \texttt{id} field.
@@
 The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
-The optional columns are user-defined arbitrary fields that relate to the experiment itself (\emph{e.g.} condition and sex).
+The optional columns are user-defined arbitrary fields that translate experimental conditions (\emph{e.g.} treatment, genotype and sex).
 
@@
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
-Indeed, additional replicates can simply be added as new rows and replicate ID later used, if needed, as a covariate.
+Indeed, additional replicates can simply be added as new rows -- and the ID of the replicate later used, if needed, as a covariate.
 
@@
 
+Rob
+
+Markus
+
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.

### 2018-02-09 manuscript.tex
cleanup repo

--- 2018-02-09 plos-latex-template/manuscript.tex
@@
 
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\caption{{\bf Population statistics on circadian phenotype.}
+			A: Average periodograms. 
+			      The aggregated relative power of the periodogram of all animals.
+			      The lines shows and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
+			B: Peak periodicity power and average.
+			      Values of the peak period for animals with a significant peak.
+			      Individual animals are shown by dots whose size represent power of the peak period.
+			      The error bars are 95\% bootstrap confidence interval on the population mean.
+			C: Frequencies of rhythmic animals.
+			      Number of rhythmic animals (\emph{i.e.} with a significant peak) in each genotypes.
+			      Dark and clear fillings indicate rhythmic and arhythmic animals, respectively.
+			      }
+	\label{fig:fig-4}
+\end{figure}
+
 To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-4}B). First of all, we comput a summary per individual (\texttt{by=id}):
@@
 
-\begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
-	\caption{{\bf Population statistics on circadian phenotype.}
-			TODO}
-	\label{fig:fig-4}
-\end{figure}
+Some conclusion here, TODO:
 

### 2018-02-09 manuscript.tex
add figure (fig4). A handpicked subset of S1.

@@
 \begin{alltt}
+\hlcom{# we also show a subset of this figure in supplementary materials}
 \hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid (see \nameref{S1-Fig}B).
+In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid.
+A representative sample of all animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).
 
@@
 \hlstd{per_dt} \hlkwb{<-} \hlkwd{find_peaks}\hlstd{(per_dt)}
-
+\hlcom{# we also show a subset of this figure in supplementary materials}
 \hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak}\hlstd{=peak))} \hlopt{+}
@@
 
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
+		A: Double plotted actograms showing activity over the experiment. Transition to LL happens at day 0.
+		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
+		Titles on top of each facet refer to the label allocated to each individual.
+		A version of this figure with all animals is available \nameref{S1-Fig}.
+	}
+	\label{fig:fig-4}
+\end{figure}
+
 \subsection*{Population statistics}
 Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
-arhythmic in LL whilst Tim/NKCCOX appear to have a consistant, long-period rhythm.
-To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-4}A):
+arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
+To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
 
@@
 \begin{alltt}
-\hlstd{fig4A} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power} \hlopt{-} \hlstd{signif_threshold,}
+\hlstd{fig5A} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power} \hlopt{-} \hlstd{signif_threshold,}
                              \hlkwc{colour} \hlstd{= genotype))} \hlopt{+}
@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
 	\caption{{\bf Population statistics on circadian phenotype.}
@@
 			      }
-	\label{fig:fig-4}
+	\label{fig:fig-5}
 \end{figure}
 
-To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-4}B). First of all, we comput a summary per individual (\texttt{by=id}):
+To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-5}B). First of all, we compute a summary per individual (\texttt{by=id}):
 
@@
 
-\texttt{summary\_dt} is just a regular data frame with one row per individual, contianing both metadata and our summary statistics. It can therefore be used directly by \texttt{ggplot} and such:
+\texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics. It can therefore be used directly by \texttt{ggplot} and such:
 
@@
 \hlcom{# standard ggplot}
-\hlstd{fig4B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
+\hlstd{fig5B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
                                 \hlkwc{x} \hlstd{= genotype))} \hlopt{+}
@@
 
-To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhytmic flies (see Fig~\ref{fig:fig-4}C):
+To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhythmic flies (see Fig~\ref{fig:fig-5}C):
 \begin{knitrout}
@@
 \hlcom{# standard ggplot}
-\hlstd{fig4C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
+\hlstd{fig5C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
                                 \hlkwc{fill} \hlstd{= genotype,}
@@
 \label{S1-Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+{\bf Complete version of Fig~\ref{fig:fig-4}.}
+See Fig~\ref{fig:fig-4} for legend.
 

### 2018-02-09 results.Rnw
add figure (fig4). A handpicked subset of S1.

@@
 <<cache=TRUE>>=
+# we also show a subset of this figure in supplementary materials
 figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
@@
 @
+<<cache=TRUE, echo=FALSE>>=
+fig_4_picked_labels = paste(c(5,10,33,39, 13,17,22,51), rep(c("NKCCOX/+", "Tim/NKCCOX"), each=4), sep=".")
+
+fig4A <- ggetho(dt[xmv(label) %in% fig_4_picked_labels],
+						 aes(z = moving), multiplot = 2) +
+	stat_bar_tile_etho() + 
+	facet_wrap( ~ label, ncol = 4) +
+	scale_y_discrete(name="Day")
+@
 
@@
 per_dt <- find_peaks(per_dt)
-
+# we also show a subset of this figure in supplementary materials
 figS1B <- ggperio(per_dt, aes(y = power, peak=peak)) + 
@@
 
+<<cache=TRUE, echo=FALSE>>=
+fig4B <- ggperio(per_dt[xmv(label) %in% fig_4_picked_labels],
+aes(y = power, peak=peak)) + 
+# periododogram drawn as a line
+geom_line() +
+# the significance line in red
+geom_line(aes(y = signif_threshold), colour="red") + 
+# point and text at the peak
+geom_peak() + 
+# divide plot by individual
+facet_wrap( ~ label, ncol = 4) 
+@
+
+\begin{figure}[!h]
+	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
+		A: Double plotted actograms showing activity over the experiment. Transition to LL happens at day 0.
+		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
+		Titles on top of each facet refer to the label allocated to each individual.
+		A version of this figure with all animals is available \nameref{S1-Fig}.
+	}
+	\label{fig:fig-4}
+\end{figure}
+
 \subsection*{Population statistics}
 Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
-arhythmic in LL whilst Tim/NKCCOX appear to have a consistant, long-period rhythm.
-To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-4}A):
+arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
+To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
 
 <<cache=TRUE, warning=FALSE,message=FALSE, result="hide">>=
-fig4A <- ggperio(per_dt, aes(y = power - signif_threshold,
+fig5A <- ggperio(per_dt, aes(y = power - signif_threshold,
                              colour = genotype)) + 
@@
 # standard ggplot
-fig4B <- ggplot(summary_dt, aes(y = first_peak_period,
+fig5B <- ggplot(summary_dt, aes(y = first_peak_period,
                                 x = genotype)) + 
@@
 
-To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhytmic flies (see Fig~\ref{fig:fig-4}C):
+To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhythmic flies (see Fig~\ref{fig:fig-5}C):
 <<cache=TRUE, warning=FALSE, results="hide">>=
 # standard ggplot
-fig4C <- ggplot(summary_dt, aes(x = genotype,
+fig5C <- ggplot(summary_dt, aes(x = genotype,
                                 fill = genotype,
@@
 tmp <- dev.off()
+
+#===================================================
+
+fig4 <- cowplot::plot_grid(
+				fig4A,
+				fig4B,
+				nrow=1,
+				labels=c("A", "B"),
+				align="v"
+)
+
+pdf("../fig/fig-4.pdf",w=10,h=3)
+print(fig4)
+tmp  <- dev.off()
+
 #=================================================
 
-fig4 <- cowplot::plot_grid(fig4A + theme(legend.position="none"),
-                           fig4B + theme(legend.position="bottom"),
-                           fig4C + theme(legend.position="none"),
+fig5 <- cowplot::plot_grid(fig5A + theme(legend.position="none"),
+                           fig5B + theme(legend.position="bottom"),
+                           fig5C + theme(legend.position="none"),
                            nrow=1,
@@
                            )
-pdf("../fig/fig-4.pdf",w=10,h=5)
-print(fig4)
+pdf("../fig/fig-5.pdf",w=10,h=5)
+print(fig5)
 tmp <- dev.off()

### 2018-02-10 results.Rnw
work on #1

@@
                           nrow=1,
-                          labels=c("", "A", "B"),
+                          labels=c("A", "", "B"),
                           rel_widths = c(1, 2, 2),
-                          label_size = 20,
+                          label_size = 32,
                           axis="b",
-                          align="h"
+                          align="h",
+                          hjust=0, vjust=1.01
                           )
-pdf("../fig/fig-3.pdf",w=10,h=10)
+pdf("../fig/fig-3.pdf",w=12,h=12)
 print(fig3)
@@
                           labels=c( "A", "B"),
-                          label_size = 20,
-                          align="v"
+                          label_size = 32,
+                          align="v",
+                          axis="b",
+                          hjust=0, vjust=1.01
                           )
@@
 				labels=c("A", "B"),
-				align="v"
+				align="v",
+				axis="b",
+				label_size = 32,
+				hjust=0, vjust=1.01
 )
 
-pdf("../fig/fig-4.pdf",w=10,h=3)
+pdf("../fig/fig-4.pdf",w=12,h=4)
 print(fig4)
@@
                            nrow=1,
-                           label_size = 20,
+                           label_size = 32,
                            align = "hv",
                            axis="b",
-                           labels=c( "A", "B", "C")
+                           labels=c( "A", "B", "C"),
+                           hjust=0, vjust=1.01
                            #?cowplot::plot_grid
                            )
-pdf("../fig/fig-5.pdf",w=10,h=5)
+pdf("../fig/fig-5.pdf",w=12,h=6)
 print(fig5)

### 2018-02-10 manuscript.Rnw
add figure (fig4). A handpicked subset of S1.
fixes #1

@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
+	%\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
 	\caption{{\bf The \texttt{rethomics} workflow.}
@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
-	\caption{{\bf Example of a \texttt{behavr} table.}
+	%\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
+	\caption{{\bf \texttt{behavr} table.}
+	A: Illustration of a \texttt{behavr} object, core data structure in \texttt{rethomics}.
 		The metadata holds a single row for each of the $n$ individuals. 
@@
 		Data and metadata are implicitly joined with the \texttt{id} field.
-		In this figure, the names used for variables and metavariable are only plausible examples that will likely differ in practice. 
+		Names used for variables and metavariable are only plausible examples that will likely differ in practice. 
+		B: Non exhaustive list of use of a \texttt{behavr} table, refered as \texttt{dt}. 
+		In addition of operations on data inherited from \texttt{data.table}, we provide utilities designed specifically to act on both            metadata and data.  
 	}

### 2018-02-10 manuscript.tex
fixes #1

@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
+	%\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
 	\caption{{\bf Experiment quality control.}
@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+	%\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
 	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
@@
 \begin{figure}[!h]
-	\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
+	%\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
 	\caption{{\bf Population statistics on circadian phenotype.}

### 2018-02-10 results.Rnw
fixes #1
fixes #6, fixes #3

@@
 
-<<cache=TRUE>>=
+<<cache=TRUE, warning=FALSE>>=
 # we also show a subset of this figure in supplementary materials
@@
 @
+
 <<cache=TRUE, echo=FALSE>>=
-fig_4_picked_labels = paste(c(5,10,33,39, 13,17,22,51), rep(c("NKCCOX/+", "Tim/NKCCOX"), each=4), sep=".")
+#fig_4_picked_labels = paste(c(5,10,33,39, 13,17,22,51), rep(c("NKCCOX/+", "Tim/NKCCOX"), each=4), sep=".")
+helper_dt_picked_labs <- data.table(num = c(5,10,33,39, 13,17,22,51),
+                                    genotype =  rep(c("NKCCOX/+", "Tim/NKCCOX"), each=4))
+helper_dt_picked_labs[,label := interaction(num, genotype)]
+
+  # add a bg that depends on labels
+bg_layer =   geom_rect(data=helper_dt_picked_labs, aes(fill=genotype),
+            xmin = -Inf,xmax = Inf,
+            ymin = -Inf,ymax = Inf,
+            alpha = 0.3, inherit.aes = F)
+
+fig_4_picked_labels = helper_dt_picked_labs[,label]
 
@@
 						 aes(z = moving), multiplot = 2) +
+  bg_layer+
 	stat_bar_tile_etho() + 
 	facet_wrap( ~ label, ncol = 4) +
-	scale_y_discrete(name="Day")
+	scale_y_discrete(name="Day") +
+  theme(strip.text= element_text(size = 11))
+
+ylabs <- ggplot_build(fig4A)$layout$panel_ranges[[1]]$y.labels
+fig4A <- fig4A + scale_y_discrete(name="Day",
+                                  limits = ifelse(1:length(ylabs) %%2, ylabs, ""))
 @
@@
 
-<<cache=TRUE, echo=FALSE>>=
+<<cache=TRUE, echo=FALSE, warning=FALSE>>=
 fig4B <- ggperio(per_dt[xmv(label) %in% fig_4_picked_labels],
@@
             geom_line() +
+            bg_layer+
             # the significance line in red
@@
             # divide plot by individual
-facet_wrap( ~ label, ncol = 4) 
+            facet_wrap( ~ label, ncol = 4) + 
+            scale_y_continuous(name="Power") +
+            theme(strip.text= element_text(size = 11))
+
 @
@@
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
+		The blue annotation represents the first peak (if present) above the significance threshold (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
@@
                         #legend.text = "moving",
-                        legend.position="left") + 
+                        legend.position="none") + 
                   coord_cartesian(xlim=-days(c(10, 9))) + scale_y_discrete(position = "right")
@@
                           )
+fill_legend <- 
+  cowplot::plot_grid(NULL,
+    cowplot::get_legend(
+        fig3A + theme(legend.position = "bottom", legend.direction = "horizontal") + scale_fill_continuous(labels = scales::percent, name="Moving ", breaks=c(0,.5,1))
+        ), NULL
+    )
+#?theme
+
+fig3_final <- cowplot::plot_grid(fig3, fill_legend,nrow = 2, rel_heights = c(15,1))
+
 pdf("../fig/fig-3.pdf",w=12,h=12)
-print(fig3)
+print(fig3_final)
 tmp  <- dev.off()
@@
 fig4 <- cowplot::plot_grid(
-				fig4A,
-				fig4B,
+				fig4A + theme(legend.position = "none"),
+				fig4B+ theme(legend.position = "none"),
 				nrow=1,
@@
 
+fill_legend <- 
+  cowplot::plot_grid(NULL,
+    cowplot::get_legend(
+        fig4A + theme(legend.position = "bottom", legend.direction = "horizontal") + scale_fill_discrete(name="Genotype "))
+        )
+#?theme
+
+fig4_final <- cowplot::plot_grid(fig4, fill_legend,nrow = 2, rel_heights = c(15,1))
 pdf("../fig/fig-4.pdf",w=12,h=4)
-print(fig4)
+print(fig4_final)
 tmp  <- dev.off()
@@
 fig5 <- cowplot::plot_grid(fig5A + theme(legend.position="none"),
-                           fig5B + theme(legend.position="bottom"),
-                           fig5C + theme(legend.position="none"),
+                           fig5B + theme(legend.position="none") + scale_x_discrete(name="Genotype"),
+                           fig5C + theme(legend.position="none") + scale_y_continuous(name="Number") +
+                             scale_x_discrete(name="Genotype"),
                            nrow=1,
@@
                            )
+
+dummy <- ggplot(summary_dt, aes(is_rhythmic, first_peak_period, size= first_peak_power, fill= genotype, colour=genotype)) + 
+    geom_bar(aes(x=genotype, alpha= ifelse(is_rhythmic, "Rhythmic", "Arhythmic"), fill=genotype), inherit.aes = F) + 
+  geom_jitter(aes(is_rhythmic, first_peak_period, size=first_peak_power), inherit.aes = F) +
+   scale_fill_discrete(name="Genotype")+scale_size(name= "Peak power") +  scale_alpha_discrete(name= "") +
+  theme(legend.position = "bottom")  + 
+  guides(fill = guide_legend(order=1),
+         size = guide_legend(order=2),
+         alpha = guide_legend(order=3))
+
+legend <- plot_grid(NULL, get_legend(dummy), NULL, nrow=1)
+
+fig5_final <- cowplot::plot_grid(fig5, legend, nrow = 2, rel_heights = c(15,1))
 pdf("../fig/fig-5.pdf",w=12,h=6)
-print(fig5)
+print(fig5_final)
 tmp <- dev.off()

### 2018-02-10 manuscript.Rnw
random updates

@@
 		Names used for variables and metavariable are only plausible examples that will likely differ in practice. 
-		B: Non exhaustive list of use of a \texttt{behavr} table, refered as \texttt{dt}. 
-		In addition of operations on data inherited from \texttt{data.table}, we provide utilities designed specifically to act on both            metadata and data.  
+		B: Non exhaustive list of uses of a \texttt{behavr} table, refered as \texttt{dt}. 
+		In addition to operations on data, which are inherited from \texttt{data.table},
+		we provide utilities designed specifically to act on both metadata and data.  
 	}

### 2018-02-10 manuscript.tex
fixes #6, fixes #3
random updates

@@
 
-\texttt{R} provide one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
+\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic vs arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of ($p$-value $< 10^{-5}$):
+To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):
 \begin{knitrout}

### 2018-02-10 results.Rnw
random updates

@@
 
-<<cache=TRUE, warning=FALSE>>=
+<<cache=TRUE, warning=FALSE, result="hide", message=FALSE>>=
 # we also show a subset of this figure in supplementary materials
@@
 
-<<cache=TRUE, echo=FALSE>>=
+<<cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, result="hide">>=
 #fig_4_picked_labels = paste(c(5,10,33,39, 13,17,22,51), rep(c("NKCCOX/+", "Tim/NKCCOX"), each=4), sep=".")
@@
 
-\texttt{R} provide one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
+\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic vs arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of ($p$-value $< 10^{-5}$):
+To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):
 <<>>=

### 2018-02-10 manuscript.Rnw
ypos et al

@@
 	\caption{{\bf \texttt{behavr} table.}
-	A: Illustration of a \texttt{behavr} object, core data structure in \texttt{rethomics}.
+	A: Illustration of a \texttt{behavr} object, the core data structure in \texttt{rethomics}.
 		The metadata holds a single row for each of the $n$ individuals. 
-		Its columns, the $p$ metavariables, are of two kind: either required and defined by the acquisition platform (\emph{i.e.} used to fetch the data) or user-defined (\emph{i.e.} arbitrary).
+		Its columns, the $p$ metavariables, are one of two kinds: either required -- and defined by the acquisition platform (\emph{i.e.} used to fetch the data) -- or user-defined (\emph{i.e.} arbitrary).
 		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time point).
@@
 		Data and metadata are implicitly joined with the \texttt{id} field.
-		Names used for variables and metavariable are only plausible examples that will likely differ in practice. 
-		B: Non exhaustive list of uses of a \texttt{behavr} table, refered as \texttt{dt}. 
+		Note that the names used in this for variables and metavariable in this example are only plausible cases that will likely differ in practice. 
+		B: Non exhaustive list of uses of a \texttt{behavr} table (refered as \texttt{dt}). 
 		In addition to operations on data, which are inherited from \texttt{data.table},
@@
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
-All packages are continuously integrated and unit tested on several version of \texttt{R} to minimise resent and future issues.
+All packages are continuously integrated and unit tested on several version of \texttt{R} to minimise the risk of present and future issues.
 
@@
 
-Hannah
+Hannah, Alice, Diana
 
-Patrick Kr{\"a}tschmer
-
-Rob
-
-Markus
+Markus, Patrick Kr{\"a}tschmer Rob
 

### 2018-02-10 manuscript.tex
ypos et al

@@
 ``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals:
+In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.
 
@@
 \begin{alltt}
-\hlcom{# we also show a subset of this figure in supplementary materials}
+\hlcom{# we also show a subset of this figure in 4A}
 \hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-In several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid.
-A representative sample of all animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).
+Several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid.
+A subset of eight animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).
 
@@
                              \hlkwc{colour} \hlstd{= genotype))} \hlopt{+}
-          \hlcom{# periododogram a line for population mean}
+          \hlcom{# periododogram shown as a line for population mean}
           \hlcom{# and bootstrap error bars}
@@
 			      The aggregated relative power of the periodogram of all animals.
-			      The lines shows and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
+			      The solid lines and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
 			B: Peak periodicity power and average.
@@
 
-\texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics. It can therefore be used directly by \texttt{ggplot} and such:
+\texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics.
+It can therefore be used directly by \texttt{ggplot} and other tools:
 
@@
 \texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
-One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic vs arhythmic flies in both genotypes.
+One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
 To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):

### 2018-02-10 results.Rnw
ypos et al

@@
 ``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals:
+In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.
 
 <<cache=TRUE, warning=FALSE, result="hide", message=FALSE>>=
-# we also show a subset of this figure in supplementary materials
+# we also show a subset of this figure in 4A
 figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
@@
 ylabs <- ggplot_build(fig4A)$layout$panel_ranges[[1]]$y.labels
+
+new_labs <- ifelse(1:length(ylabs) %%2, ylabs, "")
+
 fig4A <- fig4A + scale_y_discrete(name="Day",
-                                  limits = ifelse(1:length(ylabs) %%2, ylabs, ""))
+                                  labels=new_labs
+                                  )
 @
@@
                              colour = genotype)) + 
-          # periododogram a line for population mean
+          # periododogram shown as a line for population mean
           # and bootstrap error bars

### 2018-02-10 manuscript.Rnw
start stuffing with refs

@@
 
-\texttt{rethomics} is implemented as a collection of small packages related to one another (Fig~\ref{fig:fig-1}).
-This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}, which results in increased testability and maintainability.
+\texttt{rethomics} is implemented as a collection of small \texttt{R}\cite{r_core_team_r:_2017} packages related to one another (Fig~\ref{fig:fig-1}).
+This paradigm follows the model of modern frameworks such as the \texttt{tidyverse} \cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
 In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
 are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object.
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
@@
 To integrate visualisation in \texttt{rethomics},
-we implemented \texttt{ggetho}, a package extending the widely adopted \texttt{ggplot2} by providing preprocessing tools as well as new layers and scales.
+we implemented \texttt{ggetho}, a package extending the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016} by providing preprocessing tools as well as new layers and scales.
 Our tools make full use of the internal \texttt{behavr} structure to deliver efficient representations of temporal trends.
@@
 
-\begin{thebibliography}{10}
+\bibliography{manuscript}{}
 
-\bibitem{bib1}
-Conant GC, Wolfe KH.
-\newblock {{T}urning a hobby into a job: how duplicated genes find new
-  functions}.
-\newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-
-\bibitem{bib2}
-Ohno S.
-\newblock Evolution by gene duplication.
-\newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-  Springer-Verlag.; 1970.
-
-\bibitem{bib3}
-Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-\newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-  infection through a transposon insertion followed by a {D}uplication}.
-\newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-
-\end{thebibliography}
+%\begin{thebibliography}{10}
+%\end{thebibliography}
 

### 2018-02-13 manuscript.Rnw
typos --

@@
 The recent development of automatised methods that can score various behaviours on a large number of animals
-% brings the promise of a powerful and original insight
 provides biologists with an unprecedented set of tools to decipher these complex phenotypes. 
@@
 import, store, visualise and analyse behavioural data.
-In this article, we describe it and show an example of its application to the blooming field of sleep and circadian rhythm in fruit fly.
+In this article, we describe it and show an example of its application to the blooming field of sleep and circadian rhythm in fruit flies.
 The \texttt{rethomics} framework is available and documented at \href{https://rethomics.github.io}{https://rethomics.github.io}.
@@
 There are, at least, three issues with this approach.
-First of all, state-of-the-art analysis and visualisation requires a level of reproducibility, flexibility and scalability than only a programmatic interface can provide.
+First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programmatic interface can provide.
 Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
@@
 \texttt{rethomics} is implemented as a collection of small \texttt{R}\cite{r_core_team_r:_2017} packages related to one another (Fig~\ref{fig:fig-1}).
-This paradigm follows the model of modern frameworks such as the \texttt{tidyverse} \cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
+This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
 In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
@@
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
-Results and plots integrate seamlessly within the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
+Results and plots integrate seamlessly into the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
 
@@
 For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second, over multiple days.
-Therefore, data structure must be computationally efficient -- both in term of memory footprint and processing speed.
+Therefore, the data structure must be computationally efficient -- both in term of memory footprint and processing speed.
 
-Secondly, a large amount of individuals are often studied (typically $n > 100$).
+Secondly, a large number of individuals are often studied (typically $n > 100$).
 Each individual ($i$) is associated with metadata: a set of $p$ ``metavariables'' that describe experimental conditions.
@@
 \texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata.
-This approach guaranties that any data point can be mapped correctly to its parent metadata.
+This approach guarantees that any data point can be mapped correctly to its parent metadata.
 It also allows implicit update of metadata when data is altered.
@@
 		Its columns, the $p$ metavariables, are one of two kinds: either required -- and defined by the acquisition platform (\emph{i.e.} used to fetch the data) -- or user-defined (\emph{i.e.} arbitrary).
-		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time point).
+		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time-point).
 		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual $i$.
@@
 Although the structure of the raw results if very different, conceptually, loading data is very similar.
-In all cases the user is asked to generate a metadata table (one row per individual). 
+In all cases, the user is asked to generate a metadata table (one row per individual). 
 In it, there will be both mandatory and optional columns.
@@
 Using such a structure comes with multiple advantages.
-For instance, it simplify collaboration and data exchange as all treatments and individuals are very explicit.
+For instance, it helps collaboration and data exchange as all treatments and individuals are very explicit.
 Then, it promotes good experimental practices such as interspersion of treatments (indeed, without it, users are tempted to simplify their design by, for instance, confounding device/location and treatment).
@@
 The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute periodograms and find their peaks, score sleep from inactivity (\emph{e.g.} using the ``five minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
+Together, they offer a suite of methods to compute periodograms and find their peaks, score sleep from inactivity (\emph{e.g.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
 

### 2018-02-13 manuscript.tex
typos --

@@
 \subsection*{Preprocessing}
-We notice, from the metadata, that the two replicates do not have the same time in baseline.
+We notice, from the metadata, that the two replicates do not have the same baseline time.
 We would like to express the time relative to the important event: the transition to \texttt{LL}. 
@@
 
-To simplify visualisation, we create our own \texttt{label} metavariable, as combination of a number and \texttt{genotype}.
-In the restricted context of this analysis, \texttt{label} acts a unique identifier.
-Importantly, we keep \texttt{id} , which is more rigorous and universal.
+To simplify visualisation, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
+Importantly, we keep \texttt{id}, which is more rigorous and universal.
 
@@
 			Tile plot representing the fraction of time spent moving as a colour intensity.
-			Each individual is represented by a row and time, on the x axis, is binned in 30 minute.}
+			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.}
 	\label{fig:fig-3}
@@
 
-Activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
-Our \texttt{sleepr} package offer a tool to detect and remove such artefactual data.
+The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
+Our \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
 The updated version can be visualised in Fig~\ref{fig:fig-3}B.
@@
 
-\subsection*{Double plotted actograms}
+\subsection*{Double-plotted actograms}
 ``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animals. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.
+In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.
 
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-Several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodrograms and lay them out in a grid.
+Several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
 A subset of eight animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).
@@
 
-\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in he analysis of the extracted variables.
+\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.

### 2018-02-13 manuscript.Rnw
todolist

@@
 
+\reversemarginpar
 \usepackage[colorinlistoftodos]{todonotes}
@@
 \newcommand{\citationneeded}[2][]{\todo[color=brown, fancyline, #1]{\textbf{Citation Needed:} #2}}
+\newcommand{\TODO}[2][]{\todo[color=red, fancyline, #1]{\textbf{TODO:} #2}}
 
@@
 
+\listoftodos
+
 % Please keep the Author Summary between 150 and 200 words
@@
 Behaviour scoring is certainly not an exception to this trend.
-Indeed, many platforms (TODO citations) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
+Indeed, many platforms (\citationneeded{}) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
 
@@
 Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data and another one for Ethoscope data.
+Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data and another one for Ethoscope data\citationneeded{plos biol. 2017}.
 Although the structure of the raw results if very different, conceptually, loading data is very similar.
@@
 \section*{Acknowledgements}
-TODO:
+\TODO{Thank people here}
 

### 2018-02-13 manuscript.tex
todolist

@@
 
-TODO description of the dataset here
+\TODO{resut intro here}
 
@@
 
-* express the idea that shis is just simple example that could be scaled up
+* express the idea that this is just simple example that could be scaled up
 
@@
 
-Some conclusion here, TODO:
+\TODO{some concluding remarks}
 

### 2018-02-13 manuscript.Rnw
fix makefile and typo and bib

@@
 %% Include all macros below
-\newcommand{\citationneeded}[2][]{\todo[color=brown, fancyline, #1]{\textbf{Citation Needed:} #2}}
+\newcommand{\citationneeded}[2][]{\todo[color=blue, fancyline, #1]{\textbf{Citation Needed:} #2}}
 \newcommand{\TODO}[2][]{\todo[color=red, fancyline, #1]{\textbf{TODO:} #2}}
@@
 		Note that the names used in this for variables and metavariable in this example are only plausible cases that will likely differ in practice. 
-		B: Non exhaustive list of uses of a \texttt{behavr} table (refered as \texttt{dt}). 
+		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
 		In addition to operations on data, which are inherited from \texttt{data.table},
@@
 Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data and another one for Ethoscope data\citationneeded{plos biol. 2017}.
+Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
 Although the structure of the raw results if very different, conceptually, loading data is very similar.

### 2018-02-13 manuscript.tex
a bit more writing!

@@
 
-\TODO{resut intro here}
+In order to illustrate the workflow of \texttt{rethomics}, we decided to provide an example that is simple and reproducible, but which could be readily modified for more complicated cases.
+DAM2 being widely adopted, and fruit flies being a model of choice for high throughput studies,
+we decided to show how \texttt{rethomics} can be used to study circadian rhythm in fruit flies, using this type data.
 
-* experiment goal
+We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
+It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
+The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and TIM/+ (the appropriate control) in LL for six days.
+Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
+Two repetitions were performed resulting in a total of 58 animals.
+The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded on zenodo\citationneeded{ibid}.
 
-* design
-
-* source
-
-* express the idea that this is just simple example that could be scaled up
-
-First of all, we load the necessary libraries (see availability section for installation instructions):
+As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
+We start the analysis by loading the necessary \texttt{rethomics} packages (see availability section for installation instructions):
 \begin{knitrout}
@@
 
-\TODO{some concluding remarks}
-
-* high scalability
-
-* reproducibility
-
-* from raw data to publication-quality figures
-
-* flexibility
+This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
+We were able to comprehensively analyse classical circadian data with a few line of code.
+Since the interface is programmatic, it adapts really well to much larger data sets and provides a large degree of freedom and flexibility.
 
@@
 All packages are continuously integrated and unit tested on several version of \texttt{R} to minimise the risk of present and future issues.
+
+* How we have already used rethomics in two impactful papers
 
@@
 
-Markus, Patrick Kr{\"a}tschmer Rob
+Marcus Gosh, Patrick Kr{\"a}tschmer Rob
 

### 2018-02-13 results.Rnw
todolist
a bit more writing!

@@
 
-TODO description of the dataset here
+In order to illustrate the workflow of \texttt{rethomics}, we decided to provide an example that is simple and reproducible, but which could be readily modified for more complicated cases.
+DAM2 being widely adopted, and fruit flies being a model of choice for high throughput studies,
+we decided to show how \texttt{rethomics} can be used to study circadian rhythm in fruit flies, using this type data.
 
-* experiment goal
+We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
+It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
+The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and TIM/+ (the appropriate control) in LL for six days.
+Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
+Two repetitions were performed resulting in a total of 58 animals.
+The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded on zenodo\citationneeded{ibid}.
 
-* design
-
-* source
-
-* express the idea that shis is just simple example that could be scaled up
-
-First of all, we load the necessary libraries (see availability section for installation instructions):
+As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
+We start the analysis by loading the necessary \texttt{rethomics} packages (see availability section for installation instructions):
 <<message=FALSE>>=
@@
 
-Some conclusion here, TODO:
-
-* high scalability
-
-* reproducibility
-
-* from raw data to publication-quality figures
-
-* flexibility
+This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
+We were able to comprehensively analyse classical circadian data with a few line of code.
+Since the interface is programmatic, it adapts really well to much larger data sets and provides a large degree of freedom and flexibility.
 

### 2018-02-15 manuscript.Rnw
a bit more writing!
esteaban's comments included

@@
 However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
-We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offer utilities to:
-import, store, visualise and analyse behavioural data.
-In this article, we describe it and show an example of its application to the blooming field of sleep and circadian rhythm in fruit flies.
+We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offers utilities to:
+import, store, analyse and visualise behavioural data.
+In this article, we describe it and show an example of its application to the study of circadian rhythms and blooming field of sleep fruit flies.
 The \texttt{rethomics} framework is available and documented at \href{https://rethomics.github.io}{https://rethomics.github.io}.
@@
 
-Animal behaviours are complex phenotypical manifestations of the interaction between nervous systems and external or internal environments.
+The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal environments.
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
-Behaviour scoring is certainly not an exception to this trend.
+The scoring of behaviours is certainly not an exception to this trend.
 Indeed, many platforms (\citationneeded{}) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
@@
 The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
-Clearly, the multiplicity of model organisms, hypotheses and paradigms should be matched by a diverse range of recording tools.
-However, when it comes to the subsequent data analysis, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces with rigid functionalities that only import data from a single platform.
+The multiplicity of model organisms, hypotheses and paradigms makes the existence of diverse range of recording tools important.
+However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
+Instead, tools tend to consist of graphical interfaces that tend to import data from a single platform and provide somewhat rigid functionalities.
 There are, at least, three issues with this approach.
-First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programmatic interface can provide.
+First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programming interface can provide.
 Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
@@
 Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
-Typically, the behaviour of each individual is described by a long time series (possibly multivariate and heterogeneous).
+Typically, the behaviour of each individual is described by a long time series (possibly multivariate and irregular).
 Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating metadata and data of hundreds of individuals, each recorded for weeks, is not trivial.
+Efficiently combining and manipulating metadata and data of hundreds of individuals, each recorded for days or weeks, is not trivial.
 
 In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
-It offers an elegant computational solution to store, manipulate and visualise a large amount of data.
+We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted in the community of computational biologists with a vast number of open-source packages.
+\texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise  large amounts of data.
 We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
@@
 
-\texttt{rethomics} is implemented as a collection of small \texttt{R}\cite{r_core_team_r:_2017} packages related to one another (Fig~\ref{fig:fig-1}).
+\texttt{rethomics} is implemented as a collection of targeted \texttt{R} packages related to one another (Fig~\ref{fig:fig-1}).
 This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
@@
 are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment, genotype and so on) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}
-Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be manipulated and visualised regardless of the original input platform.
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatmentand genotype) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
+Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
 Results and plots integrate seamlessly into the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
@@
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:fig-2}), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address 
-two challenges that are inherent to manipulating ethomics results.
+We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to manipulating ethomics results.
 
@@
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is good practice to record as many metavariables as possible so they can later be used as covariates. 
+It is interesting to record multiple metavariables in so far as they so they can later be used as covariates. 
 Therefore, typically $p > 10$.
 
-\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata.
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}B).
 This approach guarantees that any data point can be mapped correctly to its parent metadata.
@@
 For instance, when one wants to update variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
+The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
 
@@
 		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual $i$.
-		Data and metadata are implicitly joined with the \texttt{id} field.
+		Data and metadata are implicitly joined on the \texttt{id} field.
 		Note that the names used in this for variables and metavariable in this example are only plausible cases that will likely differ in practice. 
@@
 Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read Drosophila Activity Monitor (DAM2) data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
-Although the structure of the raw results if very different, conceptually, loading data is very similar.
+Currently, we provide a package to read Drosophila Activity Monitor (DAM2)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
+Although the structure of the raw results is very different, conceptually, loading data is very similar.
 In all cases, the user is asked to generate a metadata table (one row per individual). 
@@
 
-In this respect, the metadata file is a standardised an comprehensive data frame describing an experiment.
-Using such a structure comes with multiple advantages.
-For instance, it helps collaboration and data exchange as all treatments and individuals are very explicit.
-Then, it promotes good experimental practices such as interspersion of treatments (indeed, without it, users are tempted to simplify their design by, for instance, confounding device/location and treatment).
+In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
+It explicitly lists all treatments and individuals, which facilitates interspersion of treatments (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
@@
 Typically, users are interested in understanding individual or population trends over time.
-To integrate visualisation in \texttt{rethomics},
-we implemented \texttt{ggetho}, a package extending the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016} by providing preprocessing tools as well as new layers and scales.
+To integrate visualisation in \texttt{rethomics}, we implemented \texttt{ggetho},
+a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016}.
+It provides preprocessing utilities as well as new layers and scales.
 Our tools make full use of the internal \texttt{behavr} structure to deliver efficient representations of temporal trends.
-It particularly applies to the visualisation of long experiments, with the ability to, for instance, annotate light and dark phases, wrap time over a circadian day, display ``double-plotted actograms'' and periodograms.
+It particularly applies to the visualisation of long experiments, with the ability to, for instance, display ``double-plotted actograms'', periodograms, annotate light and dark phases and wrap time over a given period.
 Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot2}.
@@
 The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute periodograms and find their peaks, score sleep from inactivity (\emph{e.g.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
+Together, they offer a suite of methods to compute autocorrelogram, $\chi{}^2$\citationneeded{} and Lomb-Scargle\citationneeded{}
+periodograms and find their peaks, score sleep from inactivity (\emph{i.e.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
 

### 2018-02-15 manuscript.tex
esteban's comments ++

@@
 \section*{Results}
+\TODO{@esteban rewrite this please ;)}
 
@@
 It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
-The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and TIM/+ (the appropriate control) in LL for six days.
+The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and NKCCOK/+ (the appropriate control) in LL for six days.
 Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
@@
 \subsection*{Preprocessing}
-We notice, from the metadata, that the two replicates do not have the same baseline time.
-We would like to express the time relative to the important event: the transition to \texttt{LL}. 
-To do so, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
+The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
+the transition from LD to LL. 
+Therefore, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
 This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
@@
 			Tile plot representing the fraction of time spent moving as a colour intensity.
-			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.}
+			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.
+			A: Uncurated raw data.
+			B: Data after the curation step. Time was trimmed and data from dead animals removed. 
+			Red `\textbf{+}' symbols show animals that were removed from the subsequent analysis as they had less than five complete days in LL.}
 	\label{fig:fig-3}
@@
 Our \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
-The updated version can be visualised in Fig~\ref{fig:fig-3}B.
 
@@
 \hlstd{dt} \hlkwb{<-} \hlstd{sleepr}\hlopt{::}\hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+In addition, we can trim our data to have the same number of days across experiments and individuals.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{dt} \hlkwb{<-} \hlstd{dt[t} \hlopt{%between%} \hlkwd{days}\hlstd{(}\hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{2}\hlstd{,} \hlnum{6}\hlstd{))]}
 \hlcom{# same as above}
@@
 
-For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
+For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have \emph{at least five days in LL}.
+An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.
+
 \begin{knitrout}
@@
 ##  3	variables
-##  1.40609e+05	measurements
+##  1.19546e+05	measurements
 ##  1	key (id)
@@
 ``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal. A representative sample of eight individuals is shown in Fig~\ref{fig:fig-4}A.
+In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal.
+A representative sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
 
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-Several methods are implemented in \texttt{zeitbebr}. In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
-A subset of eight animals is shown in Fig~\ref{fig:fig-4}B (see \nameref{S1-Fig}B for all individuals).
+Several methods are implemented in \texttt{zeitbebr} ($\chi{}^2$, Lomb-Scargle and autocorrelation).
+In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
+Periodograms dor the subset of eight animals used in Fig~\ref{fig:fig-4}A is shown in Fig~\ref{fig:fig-4}B. 
+See \nameref{S1-Fig}B for the visualisation of all individuals.
 
@@
 	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
-		A: Double plotted actograms showing activity over the experiment. Transition to LL happens at day 0.
+		A: Double plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
-		The blue annotation represents the first peak (if present) above the significance threshold (red line).
+		The blue cross represents the first peak (if present) above the significance threshold (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
-		A version of this figure with all animals is available \nameref{S1-Fig}.
+		See \nameref{S1-Fig} for all 52 animals.
 	}
@@
 \subsection*{Population statistics}
-Both double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
+
+As shown in the original study, double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
 arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
@@
 			      The solid lines and the shaded areas show population means and their 95\% bootstrap confidence interval, respectively.
-			B: Peak periodicity power and average.
-			      Values of the peak period for animals with a significant peak.
-			      Individual animals are shown by dots whose size represent power of the peak period.
-			      The error bars are 95\% bootstrap confidence interval on the population mean.
-			C: Frequencies of rhythmic animals.
+			B: Frequencies of rhythmic animals.
 			      Number of rhythmic animals (\emph{i.e.} with a significant peak) in each genotypes.
 			      Dark and clear fillings indicate rhythmic and arhythmic animals, respectively.
+      C: Peak periodicity power and average.
+			      Values of the peak period for animals with a significant peak (\emph{i.e.} rhythmic).
+			      Individual animals are shown by dots whose size represent relative power of the peak period.
+			      The error bars are 95\% bootstrap confidence interval on the population mean.
 			      }
@@
 
-To further quantify this difference, we can show the value of the peak for both groups (see Fig~\ref{fig:fig-5}B). First of all, we compute a summary per individual (\texttt{by=id}):
+To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each groups (see Fig~\ref{fig:fig-5}B).
+Then, we can compare the average value of the peak for the rhythmic animals (see Fig~\ref{fig:fig-5}C).
+First of all, we compute a summary per individual (\texttt{by=id}):
 
@@
                         \hlkwc{first_peak_period} \hlstd{= period[peak}\hlopt{==}\hlnum{1}\hlstd{],}
-                        \hlkwc{first_peak_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{],}
+                        \hlkwc{first_peak_rel_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{]} \hlopt{-} \hlstd{signif_threshold[peak}\hlopt{==}\hlnum{1}\hlstd{],}
                         \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak}\hlopt{==}\hlnum{1}\hlstd{)),}
                       \hlkwc{by}\hlstd{=id]}
-
 \hlcom{# rejoin metadata}
@@
 \hlcom{# standard ggplot}
-\hlstd{fig5B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
+\hlstd{fig5B} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
+                                \hlkwc{fill} \hlstd{= genotype,}
+                                \hlkwc{alpha} \hlstd{= is_rhythmic}
+                                \hlstd{))} \hlopt{+}
+              \hlkwd{geom_bar}\hlstd{(}\hlkwc{colour}\hlstd{=}\hlstr{"black"}\hlstd{)}
+\end{alltt}
+\end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlcom{# standard ggplot}
+\hlstd{fig5C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= first_peak_period,}
                                 \hlkwc{x} \hlstd{= genotype))} \hlopt{+}
@@
               \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour}\hlstd{= genotype,}
-                              \hlkwc{size}\hlstd{=first_peak_power),}
+                              \hlkwc{size}\hlstd{=first_peak_rel_power),}
                               \hlkwc{alpha}\hlstd{=}\hlnum{.67}\hlstd{)} \hlopt{+}
@@
 
-To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhythmic flies (see Fig~\ref{fig:fig-5}C):
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{alltt}
-\hlcom{# standard ggplot}
-\hlstd{fig5C} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(summary_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= genotype,}
-                                \hlkwc{fill} \hlstd{= genotype,}
-                                \hlkwc{alpha} \hlstd{= is_rhythmic}
-                                \hlstd{))} \hlopt{+}
-              \hlkwd{geom_bar}\hlstd{(}\hlkwc{colour}\hlstd{=}\hlstr{"black"}\hlstd{)}
-\end{alltt}
-\end{kframe}
-\end{knitrout}
-
 \texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):
+To address the same question, we fit a binomial generalised linear model:
+
 \begin{knitrout}
@@
 
+The result shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $6.27 \times 10^{-06}$):
+
 Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
@@
             \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
-            \hlkwc{n} \hlstd{= .N}
+            \hlkwc{n} \hlstd{= .N,}
+            \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)}\hlopt{/}\hlstd{.N}
             \hlstd{),}
@@
 \begin{verbatim}
-##      genotype mean_period sd_period n_rhythmic  n
-## 1:   NKCCOX/+    25.92500  4.133098          4 22
-## 2: Tim/NKCCOX    26.36071  1.834513         28 30
+##      genotype mean_period sd_period n_rhythmic  n percent_rhythmic
+## 1:   NKCCOX/+    25.22500  3.598495          4 22         18.18182
+## 2: Tim/NKCCOX    26.21429  2.363568         28 30         93.33333
 \end{verbatim}
@@
 We were able to comprehensively analyse classical circadian data with a few line of code.
-Since the interface is programmatic, it adapts really well to much larger data sets and provides a large degree of freedom and flexibility.
+This workflow applies  particularly really well to much larger data sets and provides a large degree of flexibility taht will allow
+biologists to tune their analysis to their specific questions.
 

### 2018-02-15 results.Rnw
esteban's comments ++

@@
 dt <- sleepr::curate_dead_animals(dt, moving)
+@
+
+In addition, we can trim our data to have the same number of days across experiments and individuals.
+<<cache=TRUE>>=
+dt <- dt[t %between% days(c(-2, 6))]
 # same as above
@@
 
-For the purpose of this example, we keep only individuals that have \emph{at least five days in LL}.
+For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have \emph{at least five days in LL}.
+An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.
+
+<<cache=TRUE, echo=FALSE>>=
+# draw point for invalid animals
+valid_dt <- dt[, .(valid = max(t) > days(5)), by = id]
+valid_ids <- valid_dt[valid == T, id]
+fig3B <- fig3B + geom_text(data = dt[!id %in% valid_ids,.(label), meta = T], aes(x=days(6), y=label), inherit.aes = F, colour="red", label="+", fontface = "bold")
+@
+
 <<cache=TRUE>>=
@@
                                   labels=new_labs
-                                  )
+                                  ) + stat_ld_annotations(ld_colours = c("white", "grey"))
 @
@@
 	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
-		A: Double plotted actograms showing activity over the experiment. Transition to LL happens at day 0.
+		A: Double plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
-		The blue annotation represents the first peak (if present) above the significance threshold (red line).
+		The blue cross represents the first peak (if present) above the significance threshold (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
-		A version of this figure with all animals is available \nameref{S1-Fig}.
+		See \nameref{S1-Fig} for all \Sexpr{nrow(dt[meta=TRUE])} animals.
 	}
@@
                         first_peak_period = period[peak==1],
-                        first_peak_power = power[peak==1],
+                        first_peak_rel_power = power[peak==1] - signif_threshold[peak==1],
                         is_rhythmic = any(peak==1)),
                       by=id]
-
 # rejoin metadata
@@
 
+<<cache=TRUE, warning=FALSE, results="hide">>=
+# standard ggplot
+fig5B <- ggplot(summary_dt, aes(x = genotype,
+                                fill = genotype,
+                                alpha = is_rhythmic
+                                )) +
+              geom_bar(colour="black")
+
+@
+
 <<>>=
 # standard ggplot
-fig5B <- ggplot(summary_dt, aes(y = first_peak_period,
+fig5C <- ggplot(summary_dt, aes(y = first_peak_period,
                                 x = genotype)) + 
@@
               geom_jitter(aes(colour= genotype, 
-                              size=first_peak_power),
+                              size=first_peak_rel_power),
                               alpha=.67) +
@@
 
-To convey that Tim/NKCCOX mutants are more rhythmic, we could represent the proportion of rhythmic flies (see Fig~\ref{fig:fig-5}C):
-<<cache=TRUE, warning=FALSE, results="hide">>=
-# standard ggplot
-fig5C <- ggplot(summary_dt, aes(x = genotype,
-                                fill = genotype,
-                                alpha = is_rhythmic
-                                )) +
-              geom_bar(colour="black")
-
-@
-
 \texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model, which shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $< 10^{-5}$):
+To address the same question, we fit a binomial generalised linear model:
+
 <<>>=
@@
 @
+
+<<echo=FALSE>>=
+str <- sapply(strsplit(format(summary(fit)$coefficients[2,4], scientific=TRUE, digits = 3), split="e"), function(x) paste0("$", x[1], " \\times 10^{", x[2], "}$"))
+@
+
+The result shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value \Sexpr{str}):
 
@@
             n_rhythmic = sum(is_rhythmic),
-            n = .N
+            n = .N,
+            percent_rhythmic = 100 * sum(is_rhythmic)/.N
             ),
@@
 library(cowplot)
+label_size <- 24 # "A", "B" ...
 my_theme <- theme(axis.title.y=element_blank(),
@@
                         axis.line = element_blank(),
+                        axis.text.y = element_text(size=8),
                         #axis.ticks.y=element_blank(),
@@
                           labels=c("A", "", "B"),
-                          rel_widths = c(1, 2, 2),
-                          label_size = 32,
+                          rel_widths = c(1, 4, 4),
+                          label_size = label_size,
                           axis="b",
@@
 
-pdf("../fig/fig-3.pdf",w=12,h=12)
+pdf("../fig/fig-3.pdf",w=12,h=8)
 print(fig3_final)
@@
                           labels=c( "A", "B"),
-                          label_size = 32,
+                          label_size = label_size,
                           align="v",
@@
 				axis="b",
-				label_size = 32,
+				label_size = label_size,
 				hjust=0, vjust=1.01
@@
 fig5 <- cowplot::plot_grid(fig5A + theme(legend.position="none"),
-                           fig5B + theme(legend.position="none") + scale_x_discrete(name="Genotype"),
-                           fig5C + theme(legend.position="none") + scale_y_continuous(name="Number") +
-                             scale_x_discrete(name="Genotype"),
+                           fig5B + theme(legend.position="none") + scale_y_continuous(name="Number") + scale_x_discrete(name="Genotype"),
+                           fig5C + theme(legend.position="none") + scale_x_discrete(name="Genotype"),
                            nrow=1,
-                           label_size = 32,
+                           label_size = label_size,
                            align = "hv",
@@
     geom_bar(aes(x=genotype, alpha= ifelse(is_rhythmic, "Rhythmic", "Arhythmic"), fill=genotype), inherit.aes = F) + 
-  geom_jitter(aes(is_rhythmic, first_peak_period, size=first_peak_power), inherit.aes = F) +
-   scale_fill_discrete(name="Genotype")+scale_size(name= "Peak power") +  scale_alpha_discrete(name= "") +
+  geom_jitter(aes(is_rhythmic, first_peak_period, size=first_peak_rel_power), inherit.aes = F) +
+   scale_fill_discrete(name="Genotype")+scale_size(name= "Peak relative power") +  scale_alpha_discrete(name= "") +
   theme(legend.position = "bottom")  + 
@@
 fig5_final <- cowplot::plot_grid(fig5, legend, nrow = 2, rel_heights = c(15,1))
-pdf("../fig/fig-5.pdf",w=12,h=6)
+pdf("../fig/fig-5.pdf",w=12,h=4)
 print(fig5_final)

### 2018-02-15 manuscript.Rnw
ref++

@@
 The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute autocorrelogram, $\chi{}^2$\citationneeded{} and Lomb-Scargle\citationneeded{}
+Together, they offer a suite of methods to compute autocorrelogram, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999}
 periodograms and find their peaks, score sleep from inactivity (\emph{i.e.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).

### 2018-02-15 manuscript.tex
ref++

@@
 
-We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
+We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
 It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.

### 2018-02-16 manuscript.Rnw
typos, figures and citations (a bit of everything)

@@
 Quentin Geissmann\textsuperscript{1*},
-Luis Garcia Rodriguez\textsuperscript{2},
+Luis G Rodriguez\textsuperscript{2},
 Esteban J Beckwith\textsuperscript{1},
@@
 \\
-\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country %TODO @luis
+\textbf{2} Institute for Neuro- and Behavioral Biology, Westf{\"a}lische Wilhelms University, 48149 M{\"u}nster, Germany
 \\
@@
 The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
-The multiplicity of model organisms, hypotheses and paradigms makes the existence of diverse range of recording tools important.
+The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
 However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces that tend to import data from a single platform and provide somewhat rigid functionalities.
+Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities.
 There are, at least, three issues with this approach.
@@
 Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating metadata and data of hundreds of individuals, each recorded for days or weeks, is not trivial.
+Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, is not trivial.
 
 In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
-We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted in the community of computational biologists with a vast number of open-source packages.
+We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
+It is also complemented with a vast ecosystem of open-source packages.
 \texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise large amounts of data.
@@
 \begin{figure}[!h]
-	%\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
+%	\includegraphics[width=1\textwidth]{fig/fig-1.pdf}
 	\caption{{\bf The \texttt{rethomics} workflow.}
@@
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to manipulating ethomics results.
+We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to handling ethomics results.
 
@@
 \begin{figure}[!h]
-	%\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
+% 	\includegraphics[width=1\textwidth]{fig/fig-2.pdf}
 	\caption{{\bf \texttt{behavr} table.}
@@
 		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time-point).
-		It is formed of $q$ variables and is expected to have a very large number of reads $k$ for each individual $i$.
+		It is formed of $q$ variables and is expected to have a very large number of reads, $k$, for each individual $i$.
 		Data and metadata are implicitly joined on the \texttt{id} field.
-		Note that the names used in this for variables and metavariable in this example are only plausible cases that will likely differ in practice. 
+		Note that the names used in this for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
 		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
@@
 		we provide utilities designed specifically to act on both metadata and data.  
+		Commented examples are prefixed by \texttt{>}.
 	}
@@
 Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read Drosophila Activity Monitor (DAM2)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
+Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor (DAM)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
 Although the structure of the raw results is very different, conceptually, loading data is very similar.
@@
 In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
-It explicitly lists all treatments and individuals, which facilitates interspersion of treatments (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
+It explicitly lists all treatments and individuals, which facilitates interspersion of conditions (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.

### 2018-02-16 manuscript.tex
typos, figures and citations (a bit of everything)

@@
 
-We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
+We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
 It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
@@
 Two repetitions were performed resulting in a total of 58 animals.
-The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded on zenodo\citationneeded{ibid}.
+The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
 
@@
 \begin{alltt}
-\hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam2_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,}\hlstr{"."}\hlstd{)}    \hlcom{# linking}
+\hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam2_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,} \hlstr{"."}\hlstd{)}   \hlcom{# linking}
 \hlcom{# print(metadata)                                     # check metadata}
@@
 \begin{alltt}
-\hlstd{dt[, label} \hlkwb{:=} \hlkwd{interaction}\hlstd{(}\hlnum{1}\hlopt{:}\hlstd{.N, genotype),} \hlkwc{meta}\hlstd{=T]}
+\hlstd{dt[, label} \hlkwb{:=} \hlkwd{interaction}\hlstd{(}\hlnum{1}\hlopt{:}\hlstd{.N, genotype),} \hlkwc{meta} \hlstd{= T]}
 \hlkwd{print}\hlstd{(dt)}
@@
 \begin{figure}[!h]
-	%\includegraphics[width=1\textwidth]{fig/fig-3.pdf}
+%    \includegraphics[width=1\textwidth]{fig/fig-3.pdf}
 	\caption{{\bf Experiment quality control.}
-			Tile plot representing the fraction of time spent moving as a colour intensity.
+			Tile plot showing the fraction of time spent moving as a colour intensity.
 			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.
@@
   \hlcom{# add layers to draw annotations to show L and D phases}
-  \hlcom{# as white and black, respectivelly}
+  \hlcom{# as white and black, respectively}
   \hlcom{# the first layer is for the baseline (until t = 0)}
@@
 \begin{alltt}
+\hlcom{# filter dt between -2d and 6d}
 \hlstd{dt} \hlkwb{<-} \hlstd{dt[t} \hlopt{%between%} \hlkwd{days}\hlstd{(}\hlkwd{c}\hlstd{(}\hlopt{-}\hlnum{2}\hlstd{,} \hlnum{6}\hlstd{))]}
@@
             \hlkwd{stat_bar_tile_etho}\hlstd{()} \hlopt{+}
-            \hlcom{# divide plot by individual}
+            \hlcom{# split plot by individual}
             \hlkwd{facet_wrap}\hlstd{(} \hlopt{~} \hlstd{label,} \hlkwc{ncol} \hlstd{=} \hlnum{4}\hlstd{)} \hlopt{+}
             \hlcom{# rename the y axis}
-            \hlkwd{scale_y_discrete}\hlstd{(}\hlkwc{name}\hlstd{=}\hlstr{"Day"}\hlstd{)}
+            \hlkwd{scale_y_discrete}\hlstd{(}\hlkwc{name} \hlstd{=} \hlstr{"Day"}\hlstd{)}
 \end{alltt}
@@
 In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
-Periodograms dor the subset of eight animals used in Fig~\ref{fig:fig-4}A is shown in Fig~\ref{fig:fig-4}B. 
+Periodograms for the subset of eight animals used in Fig~\ref{fig:fig-4}A is shown in Fig~\ref{fig:fig-4}B. 
 See \nameref{S1-Fig}B for the visualisation of all individuals.
@@
 \begin{alltt}
+\hlcom{# only the LL data}
 \hlstd{dt_ll} \hlkwb{<-} \hlstd{dt[t} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{1}\hlstd{)]}
+\hlcom{# compute chi sqr periodogram }
 \hlstd{per_dt} \hlkwb{<-} \hlkwd{periodogram}\hlstd{(moving,}
                         \hlstd{dt_ll,}
-                        \hlkwc{resample_rate} \hlstd{=} \hlnum{1}\hlopt{/}\hlkwd{mins}\hlstd{(}\hlnum{10}\hlstd{),}
+                        \hlkwc{resample_rate} \hlstd{=} \hlnum{1} \hlopt{/} \hlkwd{mins}\hlstd{(}\hlnum{10}\hlstd{),}
                         \hlkwc{FUN}\hlstd{=chi_sq_periodogram)}
@@
 \hlcom{# we also show a subset of this figure in supplementary materials}
-\hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak}\hlstd{=peak))} \hlopt{+}
+\hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak} \hlstd{= peak))} \hlopt{+}
                   \hlcom{# periododogram drawn as a line}
@@
                   \hlcom{# the significance line in red}
-                  \hlkwd{geom_line}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= signif_threshold),} \hlkwc{colour}\hlstd{=}\hlstr{"red"}\hlstd{)} \hlopt{+}
+                  \hlkwd{geom_line}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= signif_threshold),} \hlkwc{colour} \hlstd{=} \hlstr{"red"}\hlstd{)} \hlopt{+}
                   \hlcom{# point and text at the peak}
@@
 \begin{figure}[!h]
-	%\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
+%	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
 	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
-		A: Double plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
+		A: Double-plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
@@
 \begin{alltt}
+\hlcom{# display periodogram }
 \hlstd{fig5A} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power} \hlopt{-} \hlstd{signif_threshold,}
@@
           \hlcom{# rename x and y axis }
-          \hlkwd{scale_y_continuous}\hlstd{(}\hlkwc{name}\hlstd{=}\hlstr{"Relative power"}\hlstd{)} \hlopt{+}
+          \hlkwd{scale_y_continuous}\hlstd{(}\hlkwc{name} \hlstd{=} \hlstr{"Relative power"}\hlstd{)} \hlopt{+}
           \hlkwd{scale_x_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
@@
 \begin{figure}[!h]
-	%\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
+%  	\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
 	\caption{{\bf Population statistics on circadian phenotype.}
@@
 
-To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each groups (see Fig~\ref{fig:fig-5}B).
+To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each group (see Fig~\ref{fig:fig-5}B).
 Then, we can compare the average value of the peak for the rhythmic animals (see Fig~\ref{fig:fig-5}C).
@@
 \begin{alltt}
-\hlstd{summary_dt} \hlkwb{<-} \hlstd{per_dt[,}\hlkwd{.}\hlstd{(}
-                        \hlkwc{first_peak_period} \hlstd{= period[peak}\hlopt{==}\hlnum{1}\hlstd{],}
-                        \hlkwc{first_peak_rel_power} \hlstd{= power[peak}\hlopt{==}\hlnum{1}\hlstd{]} \hlopt{-} \hlstd{signif_threshold[peak}\hlopt{==}\hlnum{1}\hlstd{],}
-                        \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak}\hlopt{==}\hlnum{1}\hlstd{)),}
+\hlstd{summary_dt} \hlkwb{<-}
+    \hlstd{per_dt[,}
+          \hlkwd{.}\hlstd{(}
+           \hlkwc{first_peak_period} \hlstd{= period[peak} \hlopt{==} \hlnum{1}\hlstd{],}
+           \hlstd{\{} \hlcom{# \{\} can be used for tmp variables}
+             \hlstd{signif} \hlkwb{=} \hlstd{signif_threshold[peak} \hlopt{==} \hlnum{1}\hlstd{]}
+             \hlstd{power} \hlkwb{=} \hlstd{power[peak} \hlopt{==} \hlnum{1}\hlstd{]}
+             \hlstd{first_peak_rel_power} \hlkwb{=}  \hlstd{power} \hlopt{-} \hlstd{signif}
+            \hlstd{\},}
+           \hlkwc{is_rhythmic} \hlstd{=} \hlkwd{any}\hlstd{(peak} \hlopt{==} \hlnum{1}\hlstd{)}
+           \hlstd{),}
            \hlkwc{by}\hlstd{=id]}
+
 \hlcom{# rejoin metadata}
@@
               \hlkwd{stat_summary}\hlstd{(}\hlkwc{fun.data} \hlstd{= mean_cl_boot,} \hlkwc{geom} \hlstd{=} \hlstr{"errorbar"}\hlstd{)} \hlopt{+}
-              \hlcom{# shows all individuals as points. }
+              \hlcom{# shows all individuals as points}
               \hlcom{# the size of the point expresses the power of the peak}
-              \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour}\hlstd{= genotype,}
-                              \hlkwc{size}\hlstd{=first_peak_rel_power),}
-                              \hlkwc{alpha}\hlstd{=}\hlnum{.67}\hlstd{)} \hlopt{+}
+              \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour} \hlstd{= genotype,}
+                              \hlkwc{size} \hlstd{= first_peak_rel_power),}
+                              \hlkwc{alpha} \hlstd{=} \hlnum{0.67}\hlstd{)} \hlopt{+}
               \hlcom{# We would like to convert time in hour}
@@
 
-\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper tin he analysis of the extracted variables.
+\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in the analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
@@
 \begin{alltt}
-\hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family}\hlstd{=}\hlstr{"binomial"}\hlstd{)}
+\hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family} \hlstd{=} \hlstr{"binomial"}\hlstd{)}
 
@@
             \hlkwc{n} \hlstd{= .N,}
-            \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)}\hlopt{/}\hlstd{.N}
+            \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)} \hlopt{/} \hlstd{.N}
             \hlstd{),}
@@
 
-This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
-We were able to comprehensively analyse classical circadian data with a few line of code.
-This workflow applies  particularly really well to much larger data sets and provides a large degree of flexibility taht will allow
-biologists to tune their analysis to their specific questions.
+This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
+We were able to comprehensively analyse the data from a circadian experiment with a few line of code.
+This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in FUN(X[[i]], ...): object 'first\_peak\_rel\_power' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in FUN(X[[i]], ...): object 'first\_peak\_rel\_power' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in cowplot::plot\_grid(fig5, legend, nrow = 2, rel\_heights = c(15, : object 'fig5' not found}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in print(fig5\_final): object 'fig5\_final' not found}}\end{kframe}
+\end{knitrout}
 

### 2018-02-16 results.Rnw
typos, figures and citations (a bit of everything)

--- 2018-02-15 results.Rnw
@@
 
-We gathered a subset of the data from a recent publication \citationneeded{}, which was gracefully made publicly available by the authors\citationneeded{zenodo}.
+We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
 It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
@@
 <<cache=TRUE>>=
-metadata <- link_dam2_metadata("metadata.csv",".")    # linking
+metadata <- link_dam2_metadata("metadata.csv", ".")   # linking
 # print(metadata)                                     # check metadata
@@
 <<message=FALSE, results="hide", cache=TRUE>>=
-dt[, label := interaction(1:.N, genotype), meta=T]
+dt[, label := interaction(1:.N, genotype), meta = T]
 print(dt)
@@
   # add layers to draw annotations to show L and D phases
-  # as white and black, respectivelly
+  # as white and black, respectively
   # the first layer is for the baseline (until t = 0)
@@
 <<cache=TRUE>>=
+# filter dt between -2d and 6d
 dt <- dt[t %between% days(c(-2, 6))]
@@
 valid_ids <- valid_dt[valid == T, id]
-fig3B <- fig3B + geom_text(data = dt[!id %in% valid_ids,.(label), meta = T], aes(x=days(6), y=label), inherit.aes = F, colour="red", label="+", fontface = "bold")
+fig3B <- fig3B + geom_text(data = dt[!id %in% valid_ids,.(label), meta = T], aes(x = days(6), y = label), inherit.aes = F, colour = "red", label = "+", fontface = "bold")
 @
@@
             stat_bar_tile_etho() + 
-            # divide plot by individual
+            # split plot by individual
             facet_wrap( ~ label, ncol = 4) +
             # rename the y axis
-            scale_y_discrete(name="Day")
+            scale_y_discrete(name = "Day")
 @
@@
 	facet_wrap( ~ label, ncol = 4) +
-	scale_y_discrete(name="Day") +
-  theme(strip.text= element_text(size = 11))
+	scale_y_discrete(name = "Day") +
+  theme(strip.text = element_text(size = 11))
 
@@
 
-fig4A <- fig4A + scale_y_discrete(name="Day",
-                                  labels=new_labs
+fig4A <- fig4A + scale_y_discrete(name = "Day",
+                                  labels = new_labs
                                   ) + stat_ld_annotations(ld_colours = c("white", "grey"))
@@
 <<cache=TRUE>>=
+# only the LL data
 dt_ll <- dt[t > days(1)]
+# compute chi sqr periodogram 
 per_dt <- periodogram(moving, 
                     	dt_ll, 
-                    	resample_rate = 1/mins(10),
+                    	resample_rate = 1 / mins(10),
                     	FUN=chi_sq_periodogram)
@@
 # we also show a subset of this figure in supplementary materials
-figS1B <- ggperio(per_dt, aes(y = power, peak=peak)) + 
+figS1B <- ggperio(per_dt, aes(y = power, peak = peak)) + 
                   # periododogram drawn as a line
@@
                   # the significance line in red
-                  geom_line(aes(y = signif_threshold), colour="red") + 
+                  geom_line(aes(y = signif_threshold), colour = "red") + 
                   # point and text at the peak
@@
 fig4B <- ggperio(per_dt[xmv(label) %in% fig_4_picked_labels],
-                      aes(y = power, peak=peak)) + 
+                      aes(y = power, peak = peak)) + 
             # periododogram drawn as a line
             geom_line() +
-            bg_layer+
+            bg_layer +
             # the significance line in red
-            geom_line(aes(y = signif_threshold), colour="red") + 
+            geom_line(aes(y = signif_threshold), colour = "red") + 
             # point and text at the peak
@@
             facet_wrap( ~ label, ncol = 4) + 
-            scale_y_continuous(name="Power") +
-            theme(strip.text= element_text(size = 11))
+            scale_y_continuous(name = "Power") +
+            theme(strip.text = element_text(size = 11))
 
@@
 <<cache=TRUE, warning=FALSE,message=FALSE, result="hide">>=
+# display periodogram 
 fig5A <- ggperio(per_dt, aes(y = power - signif_threshold,
@@
           # rename x and y axis 
-          scale_y_continuous(name="Relative power") + 
+          scale_y_continuous(name = "Relative power") + 
           scale_x_hours("Period")
@@
 <<cache=TRUE, warning=FALSE, results="hide">>=
+summary_dt <- 
+    per_dt[, 
+          .(
+           first_peak_period = period[peak == 1],
+           { # {} can be used for tmp variables
+             signif = signif_threshold[peak == 1]
+             power = power[peak == 1]
+             first_peak_rel_power =  power - signif
+            },
+           is_rhythmic = any(peak == 1)
+           ),
+           by=id]
 
-summary_dt <- per_dt[,.(
-                        first_peak_period = period[peak==1],
-                        first_peak_rel_power = power[peak==1] - signif_threshold[peak==1],
-                        is_rhythmic = any(peak==1)),
-                      by=id]
 # rejoin metadata
@@
               stat_summary(fun.data = mean_cl_boot, geom = "errorbar") +
-              # shows all individuals as points. 
+              # shows all individuals as points
               # the size of the point expresses the power of the peak
-              geom_jitter(aes(colour= genotype, 
-                              size=first_peak_rel_power),
-                              alpha=.67) +
+              geom_jitter(aes(colour = genotype, 
+                              size = first_peak_rel_power),
+                              alpha = 0.67) +
               # We would like to convert time in hour
@@
 <<>>=
-fit <- glm(is_rhythmic ~ genotype, summary_dt, family="binomial")
+fit <- glm(is_rhythmic ~ genotype, summary_dt, family = "binomial")
 
@@
             n = .N,
-            percent_rhythmic = 100 * sum(is_rhythmic)/.N
+            percent_rhythmic = 100 * sum(is_rhythmic) / .N
             ),
@@
 
-This examples shows how \texttt{rethomics} can be used from the raw data to making publication quality figures and statistics.
-We were able to comprehensively analyse classical circadian data with a few line of code.
-This workflow applies  particularly really well to much larger data sets and provides a large degree of flexibility taht will allow
-biologists to tune their analysis to their specific questions.
+This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
+We were able to comprehensively analyse the data from a circadian experiment with a few line of code.
+This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
 
@@
   guides(fill = guide_legend(order=1),
-         size = guide_legend(order=2),
-         alpha = guide_legend(order=3))
+         alpha = guide_legend(order=2),
+         size = guide_legend(order=3))
 

### 2018-02-18 manuscript.Rnw
another update

@@
 The scoring of behaviours is certainly not an exception to this trend.
-Indeed, many platforms (\citationneeded{}) have been developed in order to allow biologists to continuously record behaviours such as activity, position and feeding of multiple animals over long durations (days or weeks).
+Indeed, many platforms (\citationneeded{}) have been developed in order to allow biologists to continuously record behaviours such as activity, position\cite{geissmann_ethoscopes:_2017} and feeding\cite{itskov_automated_2014} of multiple animals over long durations (days or weeks).
 
@@
 However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities.
+Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities \cite{gilestro_pysolo:_2009,schmid_new_2011-1}.
 There are, at least, three issues with this approach.
@@
 Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor (DAM)\citationneeded{} data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
+Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor System (Trikinetics Inc.) data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
 Although the structure of the raw results is very different, conceptually, loading data is very similar.
@@
 
-* How we have already used rethomics in two impactful papers
-
-* Other inputs
-
-* Position analysis
-
-* GUI
-
-* ...
+Several users, in different research groups, have already adopted and are contributing to the future development framework.
+Several new packages in the \texttt{rethomics} framework are currently envisaged. 
+They include utilities to input new behaviour tracking methods and analyse position and multi-animal interaction.
 
@@
 \section*{Acknowledgements}
-\TODO{Thank people here}
-
-Han Kim
-
-Maite
-
-Hannah, Alice, Diana
-
-Marcus Gosh, Patrick Kr{\"a}tschmer Rob
+We would like to thank people who have directly or indirectly contributed to the this manuscript.
+In particular, Han Kim, for his invaluable comments on the early versions of \texttt{rethomics} and his dedicated contribution to the tutorials.
+Maite Ogueta, for making the results data available.
+\TODO{@esteban names of the BA users}
+Alice French, Hannah Jones and Diana Bicazan for her comments as early users.
+Marcus Gosh, for his feed back on the manuscript.
+Brenna Williams, for her help to support multi-beam DAMS.
+Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework.
+Robert Dickinson, for his support.
 

### 2018-02-18 manuscript.tex
another update

@@
            \hlkwc{first_peak_period} \hlstd{= period[peak} \hlopt{==} \hlnum{1}\hlstd{],}
-           \hlstd{\{} \hlcom{# \{\} can be used for tmp variables}
+            \hlcom{# \{\} can be used for tmp variables}
+           \hlkwc{first_peak_rel_power} \hlstd{= \{}
              \hlstd{signif} \hlkwb{=} \hlstd{signif_threshold[peak} \hlopt{==} \hlnum{1}\hlstd{]}
              \hlstd{power} \hlkwb{=} \hlstd{power[peak} \hlopt{==} \hlnum{1}\hlstd{]}
-             \hlstd{first_peak_rel_power} \hlkwb{=}  \hlstd{power} \hlopt{-} \hlstd{signif}
+             \hlstd{power} \hlopt{-} \hlstd{signif}
             \hlstd{\},}
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in FUN(X[[i]], ...): object 'first\_peak\_rel\_power' not found}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in FUN(X[[i]], ...): object 'first\_peak\_rel\_power' not found}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in cowplot::plot\_grid(fig5, legend, nrow = 2, rel\_heights = c(15, : object 'fig5' not found}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in print(fig5\_final): object 'fig5\_final' not found}}\end{kframe}
-\end{knitrout}
-
 % We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 

### 2018-02-18 results.Rnw
another update

@@
            first_peak_period = period[peak == 1],
-           { # {} can be used for tmp variables
+            # {} can be used for tmp variables
+           first_peak_rel_power = {
              signif = signif_threshold[peak == 1]
              power = power[peak == 1]
-             first_peak_rel_power =  power - signif
+             power - signif
             },

### 2018-02-18 manuscript.Rnw
include referewnces and estebans comments

@@
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
-The scoring of behaviours is certainly not an exception to this trend.
-Indeed, many platforms (\citationneeded{}) have been developed in order to allow biologists to continuously record behaviours such as activity, position\cite{geissmann_ethoscopes:_2017} and feeding\cite{itskov_automated_2014} of multiple animals over long durations (days or weeks).
+The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
+Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic:_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker:_2014} animals over long durations (days or weeks).
 
@@
 However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities \cite{gilestro_pysolo:_2009,schmid_new_2011-1}.
+Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities \cite{gilestro_pysolo:_2009,schmid_new_2011}.
 There are, at least, three issues with this approach.
@@
 are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatmentand genotype) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
+At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment and genotype) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
@@
 Several new packages in the \texttt{rethomics} framework are currently envisaged. 
-They include utilities to input new behaviour tracking methods and analyse position and multi-animal interaction.
+They include utilities to input new behaviour tracking methods and analyse position and multi-animal interactions.
 
@@
 \TODO{@esteban names of the BA users}
-Alice French, Hannah Jones and Diana Bicazan for her comments as early users.
+Alice French, Hannah Jones and Diana Bicazan for their comments as early users.
 Marcus Gosh, for his feed back on the manuscript.

### 2018-02-18 manuscript.tex
include referewnces and estebans comments

@@
 \section*{Results}
-\TODO{@esteban rewrite this please ;)}
+In order to illustrate the workflow of \texttt{rethomics}, we provide a straightforward and simplified example which could be readily modified for more intricate cases.
 
-In order to illustrate the workflow of \texttt{rethomics}, we decided to provide an example that is simple and reproducible, but which could be readily modified for more complicated cases.
-DAM2 being widely adopted, and fruit flies being a model of choice for high throughput studies,
-we decided to show how \texttt{rethomics} can be used to study circadian rhythm in fruit flies, using this type data.
+The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
+Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
+We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
+Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
+In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+and that the endogenous period in LL is longer than 24 hours.
 
-We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
-It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
-The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and NKCCOK/+ (the appropriate control) in LL for six days.
-Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
-Two repetitions were performed resulting in a total of 58 animals.
+We took two genotypes employed in the study, one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in all clock neurons (TIM/NKCC\textsuperscript{ox}).
+In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six days.
 The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
+We start the analysis by loading the necessary rethomics packages (see availability section for installation instructions):
 
-As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
-We start the analysis by loading the necessary \texttt{rethomics} packages (see availability section for installation instructions):
 \begin{knitrout}
@@
 
-As shown in the original study, double-plotted actograms and periodograms suggest that NKCCOX/+ flies are mostly 
-arhythmic in LL whilst Tim/NKCCOX appear to have a consistent, long-period rhythm.
+As shown in the original study, double-plotted actograms and periodograms suggest that NKCC\textsuperscript{ox}/+ flies are mostly 
+arhythmic in LL whilst Tim/NKCC\textsuperscript{ox} appear to have a consistent, long-period rhythm.
 To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
@@
 
-The result shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value $6.27 \times 10^{-06}$):
+The result shows a strong positive effect of genotype Tim/NKCC\textsuperscript{ox} on the probability of being rhythmic ($p$-value $6.27 \times 10^{-06}$):
 

### 2018-02-18 results.Rnw
include referewnces and estebans comments

@@
 \section*{Results}
-\TODO{@esteban rewrite this please ;)}
+In order to illustrate the workflow of \texttt{rethomics}, we provide a straightforward and simplified example which could be readily modified for more intricate cases.
 
-In order to illustrate the workflow of \texttt{rethomics}, we decided to provide an example that is simple and reproducible, but which could be readily modified for more complicated cases.
-DAM2 being widely adopted, and fruit flies being a model of choice for high throughput studies,
-we decided to show how \texttt{rethomics} can be used to study circadian rhythm in fruit flies, using this type data.
+The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
+Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
+We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
+Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
+In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+and that the endogenous period in LL is longer than 24 hours.
 
-We gathered a subset of the data from a recent publication \cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
-It shows that flies, which normally become arrhythmic in a Light-Light (LL) regime, are rhythmic (with an period greater than 24h\TODO{phrasing}) when the gene NKCCox is expressed in clock neurons.
-The authors have recorded the activity of both TIM/NKCCOK (NKCCox is expressed in their clock neurons) and NKCCOK/+ (the appropriate control) in LL for six days.
-Prior to LL, the animals were recorded for a baseline three or four days in normal LD, 12h:12h, conditions.
-Two repetitions were performed resulting in a total of 58 animals.
+We took two genotypes employed in the study, one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in all clock neurons (TIM/NKCC\textsuperscript{ox}).
+In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six days.
 The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
+We start the analysis by loading the necessary rethomics packages (see availability section for installation instructions):
 
-As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
-We start the analysis by loading the necessary \texttt{rethomics} packages (see availability section for installation instructions):
 <<message=FALSE>>=
@@
 
-The result shows a strong positive effect of genotype Tim/NKCCOX on the probability of being rhythmic ($p$-value \Sexpr{str}):
+The result shows a strong positive effect of genotype Tim/NKCC\textsuperscript{ox} on the probability of being rhythmic ($p$-value \Sexpr{str}):
 

### 2018-02-19 manuscript.Rnw
comments on manuscript
Some comments on the first part of manuscript.

@@
 Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
-However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
-We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offers utilities to:
+However, there is little effort (I would not say little effort, I will rephrase it, making a genereic framework is not easy or 
+what people is focus on) in providing a generic framework to specifically analyse multiple and long behavioural time series.
+(some conector with previos sentence, for that reason, that is why...) We developed the \texttt{rethomics} framework, 
+a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
 import, store, analyse and visualise behavioural data.
@@
 
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal environments.
+The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal (input signals) environments.
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
@@
 
-The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
+The availability of large amounts of data paves the way for in-depth quantitative analyses.
 The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
@@
 Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, is not trivial.
+Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, 
+is not trivial  (why is not trivial, explain, for example: due to large datasets, memory problems,...).
 
@@
 It is also complemented with a vast ecosystem of open-source packages.
-\texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise large amounts of data.
-We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
+\texttt{rethomics} offers an elegant (I will remove subjective opinions like elegant) computational solution to store, manipulate and visualise large amounts of data.
+We see it as a tool to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.
@@
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is interesting to record multiple metavariables in so far as they so they can later be used as covariates. 
+It is interesting to record multiple metavariables in so they can later be used as covariates. 
 Therefore, typically $p > 10$.
@@
 It is also important that metadata and data can interoperate.
-For instance, when one wants to update variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
+For instance, when one wants to update a variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 \subsection*{Data import}
-Data import package translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
+Data import packages translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
 Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor System (Trikinetics Inc.) data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.

### 2018-02-20 results.Rnw
Minimum comments.
unsure

@@
 <<message=FALSE>>=
-library(damr)      # input DAM2 data
+library(damr)      # input DAM data
 library(zeitgebr)  # periodogram computation
@@
 <<cache=TRUE>>=
-metadata <- link_dam2_metadata("metadata.csv", ".")   # linking
+metadata <- link_dam_metadata("metadata.csv", ".")   # linking
 # print(metadata)                                     # check metadata
-dt <- load_dam2(metadata)                             # loading
+dt <- load_dam(metadata)                             # loading
 summary(dt)                                           # quick summary

### 2018-02-20 manuscript.Rnw

@@
 Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
-However, there is little effort (I would not say little effort, I will rephrase it, making a genereic framework is not easy or 
-what people is focus on) in providing a generic framework to specifically analyse multiple and long behavioural time series.
-(some conector with previos sentence, for that reason, that is why...) We developed the \texttt{rethomics} framework, 
+However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
+To address this limitation, we developed the \texttt{rethomics} framework, 
 a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
@@
 \section*{Introduction}
-
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal (input signals) environments.
+The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal environments.
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
@@
 Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, 
-is not trivial  (why is not trivial, explain, for example: due to large datasets, memory problems,...).
+is not trivial \TODO{explain why it is not trivial}.
 
@@
 It is also complemented with a vast ecosystem of open-source packages.
-\texttt{rethomics} offers an elegant (I will remove subjective opinions like elegant) computational solution to store, manipulate and visualise large amounts of data.
-We see it as a tool to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
+\texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of data.
+We propose it as a tool to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.
@@
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is interesting to record multiple metavariables in so they can later be used as covariates. 
+It is interesting to record multiple metavariables since they can later be used as covariates. 
 Therefore, typically $p > 10$.

### 2018-02-20 manuscript.Rnw
include some comments form marcus

@@
 Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
-However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
+However, there is little effort in providing a generic toolbox to specifically analyse multiple and long behavioural time series.
 To address this limitation, we developed the \texttt{rethomics} framework, 
@@
 %\section*{Author summary}
-%Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non %sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque %pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
-\linenumbers
+% TODO revert for sumbmisson
+% \linenumbers
+\TODO{revert line numbers before submission}
+% TODO also embed bib
 
@@
 
-The availability of large amounts of data paves the way for in-depth quantitative analyses.
+The availability of large amounts of data paves the way for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles
+and ultimately a better understanding of biology\cite{brown_study_2017}.
 The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
@@
 Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework.
-Robert Dickinson, for his support.
+\TODO{@giorgio do you think it is ok to thank Rob?} Robert Dickinson, for his support.
 

### 2018-02-20 manuscript.tex
include some comments form marcus

@@
 Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
-However, there is little effort in providing a generic framework to specifically analyse multiple and long behavioural time series.
-We developed the \texttt{rethomics} framework, a suite of \texttt{R} packages that altogether offers utilities to:
+However, there is little effort in providing a generic toolbox to specifically analyse multiple and long behavioural time series.
+To address this limitation, we developed the \texttt{rethomics} framework, 
+a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
 import, store, analyse and visualise behavioural data.
@@
 \section*{Introduction}
-
 The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal environments.
@@
 
-The availability of large amounts of data is very exciting as it paves the way for in-depth quantitative analyses.
+The availability of large amounts of data paves the way for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles
+and ultimately a better understanding of biology\cite{brown_study_2017}.
 The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
@@
 Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, is not trivial.
+Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, 
+is not trivial \TODO{explain why it is not trivial}.
 
@@
 It is also complemented with a vast ecosystem of open-source packages.
-\texttt{rethomics} offers an elegant computational solution to store, manipulate and visualise large amounts of data.
-We expect it to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
+\texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of data.
+We propose it as a tool to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.
@@
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is interesting to record multiple metavariables in so far as they so they can later be used as covariates. 
+It is interesting to record multiple metavariables since they can later be used as covariates. 
 Therefore, typically $p > 10$.
@@
 Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
-We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, which was gracefully made publicly available by the authors\cite{ogueta_ll_2018}.
+We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, which was graciously made publicly available by the authors\cite{ogueta_ll_2018}.
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
@@
 \begin{alltt}
-\hlkwd{library}\hlstd{(damr)}      \hlcom{# input DAM2 data}
+\hlkwd{library}\hlstd{(damr)}      \hlcom{# input DAM data}
 \hlkwd{library}\hlstd{(zeitgebr)}  \hlcom{# periodogram computation}
@@
 \begin{alltt}
-\hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam2_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,} \hlstr{"."}\hlstd{)}   \hlcom{# linking}
+\hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,} \hlstr{"."}\hlstd{)}   \hlcom{# linking}
 \hlcom{# print(metadata)                                     # check metadata}
-\hlstd{dt} \hlkwb{<-} \hlkwd{load_dam2}\hlstd{(metadata)}                             \hlcom{# loading}
+\hlstd{dt} \hlkwb{<-} \hlkwd{load_dam}\hlstd{(metadata)}                             \hlcom{# loading}
 \hlkwd{summary}\hlstd{(dt)}                                           \hlcom{# quick summary}
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-Several methods are implemented in \texttt{zeitbebr} ($\chi{}^2$, Lomb-Scargle and autocorrelation).
+Several methods are implemented in \texttt{zeitbebr}: $\chi{}^2$, Lomb-Scargle, autocorrelation 
+and Fourier.
 In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.

### 2018-03-31 manuscript.Rnw
rewrite intro

@@
 \section*{Abstract}
-Ethomics, a quantitative and high-throughput approach to animal behaviour, is a new and exciting field.
+%Ethomics, a quantitative and high-throughput approach to animal behaviour, is a new and exciting field.
 The recent development of automatised methods that can score various behaviours on a large number of animals
 provides biologists with an unprecedented set of tools to decipher these complex phenotypes. 
-Analysing ethomics data comes with many challenges that are largely shared across acquisition platform and paradigms.
-However, there is little effort in providing a generic toolbox to specifically analyse multiple and long behavioural time series.
-To address this limitation, we developed the \texttt{rethomics} framework, 
-a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
-import, store, analyse and visualise behavioural data.
-In this article, we describe it and show an example of its application to the study of circadian rhythms and blooming field of sleep fruit flies.
-The \texttt{rethomics} framework is available and documented at \href{https://rethomics.github.io}{https://rethomics.github.io}.
+Analysing such data comes with several challenges that are largely shared across acquisition platform and paradigms.
+%However, there is little effort in providing a generic toolbox to specifically analyse multiple and long behavioural time series.
+%To address this limitation, we developed the \texttt{rethomics} framework, 
+%We present the \texttt{rethomics} framework, 
+%a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
+%import, store, analyse and visualise behavioural data.
+In the article herein, we present \texttt{rethomics}, a set of \texttt{R} packages that unifies analysis of behavioural dataset in an efficient and flexible manner.
+%We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
+%It is also complemented with a vast ecosystem of open-source packages.
+\texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of behavioural data.
+We propose it as a tool to fill the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available \href{https://rethomics.github.io}{https://rethomics.github.io}.
+
+%In this article, we describe it and show an example of its application to the study of circadian rhythms and blooming field of sleep fruit flies.
+%The \texttt{rethomics} framework is available and documented at \href{https://rethomics.github.io}{https://rethomics.github.io}.
 
@@
 \section*{Introduction}
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system and external or internal environments.
-In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased.
+The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, internal state and environment.
+In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased\cite{stephens_big_2015}.
 The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
@@
 The availability of large amounts of data paves the way for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles
-and ultimately a better understanding of biology\cite{brown_study_2017}.
-The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse range of recording tools important.
-However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
-Instead, tools tend to consist of graphical interfaces that import data from a single platform and provide somewhat rigid functionalities \cite{gilestro_pysolo:_2009,schmid_new_2011}.
-There are, at least, three issues with this approach.
-First of all, state-of-the-art analysis and visualisation require a level of reproducibility, flexibility and scalability that only a programming interface can provide.
-Secondly, it favours replicated work as developers need to create their independent solution to similar problems.
-Lastly, it links analysis and visualisation to the target acquisition tool, which makes it very difficult to share cross-tool utilities and concepts.
+and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
+The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse set of recording tools important.
+However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a palette of building blocks in a pipeline.
 
-Thankfully, behavioural data is conceptually largely agnostic of the acquisition platform and paradigm. 	
-Typically, the behaviour of each individual is described by a long time series (possibly multivariate and irregular).
-Importantly, individuals are labelled with arbitrary metadata defined by the experimenter (\emph{e.g.} sex, treatment and genotype).
-Efficiently combining and manipulating metadata and data from hundreds of individuals, each recorded for days or weeks, 
-is not trivial \TODO{explain why it is not trivial}.
+Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
+In addition such tools can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
+For example, the field of bioinformatics benefits from sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
+Since many aspect of behaviour analysis are are becoming increasingly linked to data sciences, the development of common tools and data structures would be an interesting resource to the community.
+Behavioural experiments result in sets of long time series (sometimes multivariate and irregular), the data, but also involve a formal description of the treatment applied to each individual.
+Such implicit structure is consistent between acquisition platforms and experimental paradigms, which broaden the scope for developing of utilities an shared principles.
 
-In the article herein, we describe \texttt{rethomics}, a framework that unifies analysis of behavioural dataset in an efficient and flexible manner.
-We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
-It is also complemented with a vast ecosystem of open-source packages.
-\texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of data.
-We propose it as a tool to fill the gap between behavioural biology and data sciences, thus promoting collaboration between computational and behavioural scientists.
-\texttt{rethomics} comes with a extensive documentation and a set of both practical and theoretical demos and tutorials.
+In the article herein, we describe the \texttt{rethomics} platform, a collection of packages offering a solution to import, store, manipulate and visualise large amounts of behavioural data.
+We also present two examples of application: the analysis of circadian rhythm in fruit flies.
 
 \section*{Design and Implementation}
+\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, since it is widely taught and adopted by computational biologists,
+as a collection of packages related to one another (Fig~\ref{fig:fig-1}).
 
-\texttt{rethomics} is implemented as a collection of targeted \texttt{R} packages related to one another (Fig~\ref{fig:fig-1}).
-This paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
+This architecture paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
 In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)

### 2018-03-31 manuscript.tex
rewrite intro

@@
 \hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,} \hlstr{"."}\hlstd{)}   \hlcom{# linking}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in link\_dam\_metadata("{}metadata.csv"{}, "{}."{}): could not find function "{}link\_dam\_metadata"{}}}\begin{alltt}
 \hlcom{# print(metadata)                                     # check metadata}
 \hlstd{dt} \hlkwb{<-} \hlkwd{load_dam}\hlstd{(metadata)}                             \hlcom{# loading}
+\end{alltt}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in load\_dam(metadata): could not find function "{}load\_dam"{}}}\begin{alltt}
 \hlkwd{summary}\hlstd{(dt)}                                           \hlcom{# quick summary}
 \end{alltt}
-\begin{verbatim}
-## behavr table with:
-##  58	individuals
-##  8	metavariables
-##  2	variables
-##  1.58722e+05	measurements
-##  1	key (id)
-\end{verbatim}
-\end{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in object[[i]]: object of type 'closure' is not subsettable}}\end{kframe}
 \end{knitrout}

### 2018-04-04 manuscript.Rnw
fog 6 and a bit off proofreading

@@
 We propose it as a tool to fill the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
-Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available \href{https://rethomics.github.io}{https://rethomics.github.io}.
+Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available \href{https://rethomics.github.io}{https://rethomics.github.io}).
 
@@
 \section*{Introduction}
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, internal state and environment.
+The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, its internal state and its environment.
 In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased\cite{stephens_big_2015}.
@@
 and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse set of recording tools important.
-However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a palette of building blocks in a pipeline.
+The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse palette of recording tools important.
+However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
 
 Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
-In addition such tools can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
+In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
 For example, the field of bioinformatics benefits from sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
-Since many aspect of behaviour analysis are are becoming increasingly linked to data sciences, the development of common tools and data structures would be an interesting resource to the community.
-Behavioural experiments result in sets of long time series (sometimes multivariate and irregular), the data, but also involve a formal description of the treatment applied to each individual.
-Such implicit structure is consistent between acquisition platforms and experimental paradigms, which broaden the scope for developing of utilities an shared principles.
+Since many aspect of behaviour analysis are becoming increasingly linked to data sciences, the development of common tools and data structures would be valuable to the community.
 
-In the article herein, we describe the \texttt{rethomics} platform, a collection of packages offering a solution to import, store, manipulate and visualise large amounts of behavioural data.
-We also present two examples of application: the analysis of circadian rhythm in fruit flies.
+At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for an analogous ecosystem to arise.
+However, some low-level conceptual consistencies and methodological challenges are common between experiments.
+For instance, the results (\emph{i.e.} the `data')  feature a set of long time series (sometimes multivariate and irregular), but also contain a formal description of the treatment applied to each individual, the `metadata'.
+Storing and accessing data and metadata efficiently involves the implementation of a nested data structure which, in principle,
+can be shared between acquisition platforms and experimental paradigms.
+% it broaden the scope for developing of utilities an shared principles.
+
+In the article herein, we describe the \texttt{rethomics} platform, a collection of packages offering a solution to import, store, manipulate and visualise large amounts of behavioural results
+in a data mining and statistical framework.
+We also present two examples of application of this tool to the analysis of circadian rhythm in fruit flies, a widely studied behaviour.
 
 \section*{Design and Implementation}
-\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, since it is widely taught and adopted by computational biologists,
-as a collection of packages related to one another (Fig~\ref{fig:fig-1}).
-
-This architecture paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
-In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
-are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment and genotype) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
+\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, which is widely taught and adopted by computational biologists,
+as a collection of packages (Fig~\ref{fig:fig-1}).
+This architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
+In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package.
+At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
-Results and plots integrate seamlessly into the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
+Visualisation and other results are standard objects, and can therefore be further analysed inside the wide \texttt{R} package ecosystem.
 
@@
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to handling ethomics results.
 
-Firstly, there could be very long (typically $k_i > 10^8, \forall i \in [1,n]$), multivariate (often, $q > 10$), time series for each individual.
-For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second, over multiple days.
-Therefore, the data structure must be computationally efficient -- both in term of memory footprint and processing speed.
+Ethomics results can easily scale and data structure therefore gains from being computationally efficient – both in term of memory footprint and processing speed.
+For instance, there could be very long time series, sampled several times per second, over multiple days, for each individual. %(typically, $k_i > 10^8, \forall i \in [1,n]$) 
+In addition, time series can be multivariate, encoding coordinates, orientation, dimensions, activity, colour intensity and so on.
+Furthermore, experiments may  feature a large number of individuals. %(typically $n > 100$).
+Each individual is also associated with some metadata: a set of `metavariables' that describe experimental conditions.
+For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
+A large set of metavariables is an important asset since they can later be used as covariates.
 
-Secondly, a large number of individuals are often studied (typically $n > 100$).
-Each individual ($i$) is associated with metadata: a set of $p$ ``metavariables'' that describe experimental conditions.
-For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is interesting to record multiple metavariables since they can later be used as covariates. 
-Therefore, typically $p > 10$.
-
-\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}B).
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}A~and~B).
 This approach guarantees that any data point can be mapped correctly to its parent metadata.
-It also allows implicit update of metadata when data is altered.
+Furthermore, it allows implicit update of metadata when data is altered.
 For instance, when is data filtered, only the remaining individuals should be in the new metadata. 
 It is also important that metadata and data can interoperate.
-For instance, when one wants to update a variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
+For example, when one wants to update a variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=`male'$).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 		Its columns, the $p$ metavariables, are one of two kinds: either required -- and defined by the acquisition platform (\emph{i.e.} used to fetch the data) -- or user-defined (\emph{i.e.} arbitrary).
-		In the data, each row is a ``read'' (\emph{i.e.} information about one individual at one time-point).
+		In the data, each row is a `read' (\emph{i.e.} information about one individual at one time-point).
 		It is formed of $q$ variables and is expected to have a very large number of reads, $k$, for each individual $i$.
@@
 		we provide utilities designed specifically to act on both metadata and data.  
-		Commented examples are prefixed by \texttt{>}.
+		Commented examples are prefixed by `\texttt{>}'.
 	}
@@
 \subsection*{Visualisation}
-Long time series often need to be preprocessed before visualisation.
-Typically, users are interested in understanding individual or population trends over time.
+
 To integrate visualisation in \texttt{rethomics}, we implemented \texttt{ggetho},
 a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016}.
-It provides preprocessing utilities as well as new layers and scales.
-Our tools make full use of the internal \texttt{behavr} structure to deliver efficient representations of temporal trends.
-It particularly applies to the visualisation of long experiments, with the ability to, for instance, display ``double-plotted actograms'', periodograms, annotate light and dark phases and wrap time over a given period.
+\texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise of temporal trends.
+We implemented a set of new `layers' and `scales' that particularly applies to the visualisation of long experiments, with the ability to, for instance, display `double-plotted actograms', periodograms, annotate light and dark phases and wrap time over a given period.
 Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot2}. 
-
-\subsection*{Circadian and sleep analysis}
-The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute autocorrelogram, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999}
-periodograms and find their peaks, score sleep from inactivity (\emph{i.e.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
+For instance, \texttt{ggplot2} operations such as faceting, transforming axes and adding new layers will function natively with \texttt{ggetho}.
 

### 2018-04-04 cwt/script.R
the code for the cwt

@@ -1,0 +1,201 @@
+rm(list=ls());gc()
+options(nwarnings = 1000)
+library(scopr)
+library(ggetho)
+library(sleepr)
+library(gtools)
+library(TSclust)
+
+FEMALE_MALE_PALETTE <- c("#be2828ff", "#282896ff")
+CONTROL_SD_PALETTE <- c( "#969696ff", "#3caa3cff")
+
+METADATA <- "metadata.csv"
+CACHE <- "./cache/"
+#~ RESULT_DIR <- "./raw_results/"
+RESULT_DIR <- "/data/ethoscope_results"
+
+REMOTE_DIR <- "ftp://nas.lab.gilest.ro/auto_generated_data/ethoscope_results/"
+
+met <- fread(METADATA)
+met <- met[status == "OK" & sdi == 0]
+
+met <- link_ethoscope_metadata(met, result_dir = RESULT_DIR)
+                                      
+                                                                            
+dt <- load_ethoscope(met,
+					   max_time=days(7),
+					   reference_hour=9.0,
+					   cache = CACHE,
+					   FUN = sleep_annotation,
+					   ncores=1)
+
+
+
+
+curate_data <- function(data){
+  #data[, t := t - days(xmv(baseline_days))]
+  #data <- data[is_interpolated == F]
+	# first we remove animals that do not have enought data points
+	valid_animals <- data[,.(t_range = max(t) - min(t)), by=id][t_range >= days(5)]$id
+	data <- data[t > days(1) & 
+				 t < days(6) &
+				 id %in% valid_animals]
+	data[, t := t - days(1)]
+	
+	data[,x_rel:=ifelse(xmv(region_id) > 10, 1-x, x)]
+	data[,x_rel:=ifelse(xmv(inverted), 1-x_rel, x_rel)]
+	norm_x <- function(x){
+		min_x <- quantile(na.omit(x),probs=0.01)
+		x <- x - min_x
+		max_x <- quantile(na.omit(x),probs=0.99)
+		x / max_x
+		}
+	data[, x_rel := norm_x(x_rel), by=id]
+}
+
+
+ggetho(dt, aes(t, z=asleep)) + 
+			stat_tile_etho()
+
+dt <- curate_data(dt)
+
[…]

### 2018-04-04 manuscript.Rnw
another read through the trext

@@
 \texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of behavioural data.
-We propose it as a tool to fill the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
-Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available \href{https://rethomics.github.io}{https://rethomics.github.io}).
+We propose it as a tool to breach the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
 
@@
 and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse palette of recording tools important.
+The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques important.
 However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
@@
 Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
-In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
-For example, the field of bioinformatics benefits from sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
-Since many aspect of behaviour analysis are becoming increasingly linked to data sciences, the development of common tools and data structures would be valuable to the community.
+In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
+The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
+Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
 
-At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for an analogous ecosystem to arise.
+At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for a similar community to arise.
 However, some low-level conceptual consistencies and methodological challenges are common between experiments.
@@
 
-In the article herein, we describe the \texttt{rethomics} platform, a collection of packages offering a solution to import, store, manipulate and visualise large amounts of behavioural results
-in a data mining and statistical framework.
-We also present two examples of application of this tool to the analysis of circadian rhythm in fruit flies, a widely studied behaviour.
+In the article herein, we describe the \texttt{rethomics} platform, an effort to promote the interaction between behavioural biology and data science.
+\texttt{rethomics} is implemented as a collection of packages, altogether offering a solution to import, store, manipulate and visualise large amounts of behavioural results.
+We also present two practical examples of its application to the analysis of circadian rhythm in fruit flies, a widely studied behaviour.
 
@@
 It is also important that metadata and data can interoperate.
-For example, when one wants to update a variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=`male'$).
+For example, when one wants to update a variable according to the value of a metavariable (say, alter the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
-It explicitly lists all treatments and individuals, which facilitates interspersion of conditions (indeed, without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
+It explicitly lists all treatments and individuals, which facilitates interspersion of conditions (without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
@@
 We would like to thank people who have directly or indirectly contributed to the this manuscript.
-In particular, Han Kim, for his invaluable comments on the early versions of \texttt{rethomics} and his dedicated contribution to the tutorials.
-Maite Ogueta, for making the results data available.
+In particular, Han Kim, for his invaluable comments on the early versions of \texttt{rethomics} and his dedicated contribution to the tutorials;
+Maite Ogueta, for making the results data available;
 \TODO{@esteban names of the BA users}
-Alice French, Hannah Jones and Diana Bicazan for their comments as early users.
-Marcus Gosh, for his feed back on the manuscript.
-Brenna Williams, for her help to support multi-beam DAMS.
-Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework.
-\TODO{@giorgio do you think it is ok to thank Rob?} Robert Dickinson, for his support.
+Alice French, Hannah Jones and Diana Bicazan for their comments as early users;
+Marcus Gosh, for his feed back on the manuscript;
+Brenna Williams, for her help to support multi-beam DAMS;
+Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework;
+and Robert Dickinson, for his support.
 

### 2018-04-04 manuscript.tex
another read through the trext

@@
 \texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of behavioural data.
-We propose it as a tool to fill the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
-Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available \href{https://rethomics.github.io}{https://rethomics.github.io}.
+We propose it as a tool to breach the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
 
@@
 and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms makes the existence of a diverse set of recording tools important.
-However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a palette of building blocks in a pipeline.
+The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques important.
+However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
 
 Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
-In addition such tools can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
-For example, the field of bioinformatics benefits from sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
-Since many aspect of behaviour analysis are are becoming increasingly linked to data sciences, the development of common tools and data structures would be an interesting resource to the community.
-Behavioural experiments result in sets of long time series (sometimes multivariate and irregular), the data, but also involve a formal description of the treatment applied to each individual.
-Such implicit structure is consistent between acquisition platforms and experimental paradigms, which broaden the scope for developing of utilities an shared principles.
+In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
+The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
+Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
 
-In the article herein, we describe the \texttt{rethomics} platform, a collection of packages offering a solution to import, store, manipulate and visualise large amounts of behavioural data.
-We also present two examples of application: the analysis of circadian rhythm in fruit flies.
+At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for a similar community to arise.
+However, some low-level conceptual consistencies and methodological challenges are common between experiments.
+For instance, the results (\emph{i.e.} the `data')  feature a set of long time series (sometimes multivariate and irregular), but also contain a formal description of the treatment applied to each individual, the `metadata'.
+Storing and accessing data and metadata efficiently involves the implementation of a nested data structure which, in principle,
+can be shared between acquisition platforms and experimental paradigms.
+% it broaden the scope for developing of utilities an shared principles.
+
+In the article herein, we describe the \texttt{rethomics} platform, an effort to promote the interaction between behavioural biology and data science.
+\texttt{rethomics} is implemented as a collection of packages, altogether offering a solution to import, store, manipulate and visualise large amounts of behavioural results.
+We also present two practical examples of its application to the analysis of circadian rhythm in fruit flies, a widely studied behaviour.
 
 \section*{Design and Implementation}
-\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, since it is widely taught and adopted by computational biologists,
-as a collection of packages related to one another (Fig~\ref{fig:fig-1}).
-
-This architecture paradigm follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
-In it, the different tasks of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation)
-are explicitly handled by different packages.
-At the core of \texttt{rethomics}, the \texttt{behavr} package offers a very flexible and efficient solution to store large amounts data (\emph{e.g.} position and activity) as well as metadata (\emph{e.g.} treatment and genotype) in a single \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
+\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, which is widely taught and adopted by computational biologists,
+as a collection of packages (Fig~\ref{fig:fig-1}).
+This architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
+In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package.
+At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
-Results and plots integrate seamlessly into the \texttt{R} ecosystem, hence providing users with state-of-the-art visualisation and statistical tools.
+Visualisation and other results are standard objects, and can therefore be further analysed inside the wide \texttt{R} package ecosystem.
 
@@
 \subsection*{Internal data structure}
-We created \texttt{behavr} (Fig~\ref{fig:fig-2}A), a new data structure, based on the widely adopted \texttt{data.table} object, in order to address two challenges that are inherent to handling ethomics results.
 
-Firstly, there could be very long (typically $k_i > 10^8, \forall i \in [1,n]$), multivariate (often, $q > 10$), time series for each individual.
-For instance, each series could represent variables that encode coordinates, orientation, dimensions, activity, colour intensity and so on, sampled several times per second, over multiple days.
-Therefore, the data structure must be computationally efficient -- both in term of memory footprint and processing speed.
+Ethomics results can easily scale and data structure therefore gains from being computationally efficient – both in term of memory footprint and processing speed.
+For instance, there could be very long time series, sampled several times per second, over multiple days, for each individual. %(typically, $k_i > 10^8, \forall i \in [1,n]$) 
+In addition, time series can be multivariate, encoding coordinates, orientation, dimensions, activity, colour intensity and so on.
+Furthermore, experiments may  feature a large number of individuals. %(typically $n > 100$).
+Each individual is also associated with some metadata: a set of `metavariables' that describe experimental conditions.
+For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
+A large set of metavariables is an important asset since they can later be used as covariates.
 
-Secondly, a large number of individuals are often studied (typically $n > 100$).
-Each individual ($i$) is associated with metadata: a set of $p$ ``metavariables'' that describe experimental conditions.
-For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-It is interesting to record multiple metavariables since they can later be used as covariates. 
-Therefore, typically $p > 10$.
-
-\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}B).
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}A~and~B).
 This approach guarantees that any data point can be mapped correctly to its parent metadata.
-It also allows implicit update of metadata when data is altered.
+Furthermore, it allows implicit update of metadata when data is altered.
 For instance, when is data filtered, only the remaining individuals should be in the new metadata. 
 It is also important that metadata and data can interoperate.
-For instance, when one wants to update a variable according to the value of a metavariable (say, alter the variable $x$ only for animals with the metavariable $sex=``male"$).
+For example, when one wants to update a variable according to the value of a metavariable (say, alter the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 \subsection*{Visualisation}
-Long time series often need to be preprocessed before visualisation.
-Typically, users are interested in understanding individual or population trends over time.
+
 To integrate visualisation in \texttt{rethomics}, we implemented \texttt{ggetho},
 a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016}.
-It provides preprocessing utilities as well as new layers and scales.
-Our tools make full use of the internal \texttt{behavr} structure to deliver efficient representations of temporal trends.
-It particularly applies to the visualisation of long experiments, with the ability to, for instance, display ``double-plotted actograms'', periodograms, annotate light and dark phases and wrap time over a given period.
+\texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise of temporal trends.
+We implemented a set of new `layers' and `scales' that particularly applies to the visualisation of long experiments, with the ability to, for instance, display `double-plotted actograms', periodograms, annotate light and dark phases and wrap time over a given period.
 Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot2}. 
-
-\subsection*{Circadian and sleep analysis}
-The packages \texttt{zeitgebr} and \texttt{sleepr} provide tools to analyse circadian behaviours and sleep, respectively.
-Together, they offer a suite of methods to compute autocorrelogram, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999}
-periodograms and find their peaks, score sleep from inactivity (\emph{i.e.} using the ``five-minute rule''), and characterise the architecture of sleep bouts (\emph{e.g.} number, length and latency).
+For instance, \texttt{ggplot2} operations such as faceting, transforming axes and adding new layers will function natively with \texttt{ggetho}.
 
 \section*{Results}
-In order to illustrate the workflow of \texttt{rethomics}, we provide a straightforward and simplified example which could be readily modified for more intricate cases.
+%In order to illustrate the workflow of \texttt{rethomics}, we provide a simplified and reproducible example, which could be readily modified to suit more intricate cases.
+
+In order to illustrate the usefulness of \texttt{rethomics}, we provide two use cases examples.
+The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
+The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with an advanced case of a multi-scale analysis of a periodic behaviour using continuous wavelet transform and ethoscope data \cite{geissmann_ethoscopes:_2017}.
+
+\subsection*{Drosophila Activity Monitor}
+It uses the package \texttt{zeitgebr} which implements a suite of methods to compute autocorrelogram,
+ $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms, and find their peaks.
 
@@
 Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
-We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, which was graciously made publicly available by the authors\cite{ogueta_ll_2018}.
+We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
@@
 \hlstd{metadata} \hlkwb{<-} \hlkwd{link_dam_metadata}\hlstd{(}\hlstr{"metadata.csv"}\hlstd{,} \hlstr{"."}\hlstd{)}   \hlcom{# linking}
-\end{alltt}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in link\_dam\_metadata("{}metadata.csv"{}, "{}."{}): could not find function "{}link\_dam\_metadata"{}}}\begin{alltt}
 \hlcom{# print(metadata)                                     # check metadata}
 \hlstd{dt} \hlkwb{<-} \hlkwd{load_dam}\hlstd{(metadata)}                             \hlcom{# loading}
-\end{alltt}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in load\_dam(metadata): could not find function "{}load\_dam"{}}}\begin{alltt}
 \hlkwd{summary}\hlstd{(dt)}                                           \hlcom{# quick summary}
 \end{alltt}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in object[[i]]: object of type 'closure' is not subsettable}}\end{kframe}
+\begin{verbatim}
+## behavr table with:
+##  58	individuals
+##  8	metavariables
+##  2	variables
+##  1.58722e+05	measurements
+##  1	key (id)
+\end{verbatim}
+\end{kframe}
 \end{knitrout}
 
-\subsection*{Preprocessing}
+\subsubsection*{Preprocessing}
 The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
@@
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
-Importantly, we keep \texttt{id}, which is more rigorous and universal.
+Importantly, we also keep \texttt{id} as an \emph{unambiguous} unique identifier.
+Indeed, two animals in different experiments may have the same label, but different \texttt{id}s.
+in addition, if the metadata changes -- for instance by the addition or removal of individuals -- the label is likely to change, not the \texttt{id},
+which could lead to confusion.
 
@@
 
-\subsection*{Curation}
+\subsubsection*{Curation}
 It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
@@
 The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
-Our \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
+The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
 
@@
 
-Note that as a result, we now have 52 ``valid''individuals.
+Note that as a result, we now have 52 `valid' individuals.
 
-\subsection*{Double-plotted actograms}
-``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
+\subsubsection*{Double-plotted actograms}
+`Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
 In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal.
@@
 
-\subsection*{Periodograms}
+\subsubsection*{Periodograms}
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
@@
 
-\subsection*{Population statistics}
+\subsubsection*{Population statistics}
 
@@
 
+\subsection*{Multi-scale analysis of position}
+
+One of the challenges of behaviour analysis is the `nesting' of events happening over different time scales.
+In other words, a behavioural variable can be modulated by the circadian rhythm, but also by co-occurring ultradian and infradian rhythms.
+For instance, an animal could have rhythmic bursts of activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable, activity.
+In this example, both rhythms happen at time scales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate in the same analysis.
+Being able to keep frequency information over multiple scales is however important in some cases.
+In particular, when interested in the frequency modulation of one rhythm by another --
+that is, if the periodicity of a high frequency rhythm itself can be driven by a lower frequency one.
+%For instance, the frequency of rhythmic activity bursts could itself be regulated by another rhythm such as the circadian rhythm.
+
+The problem of understanding time series at different scales in not uncommon in fields such as economics\cite{aguiar-conraria_business_2011}, climate sciences\cite{lau_climate_1995} and ecology\cite{cazelles_wavelet_2008} where variables are governed by multiple underlying rhythms (\emph{e.g.} tidal, daily, yearly and multi-yearly).
+One approach is to study a variable of interest in the time/period domain using, for instance, continuous wavelet transform (CWT)\cite{grossmann_decomposition_1984}.
+
+To illustrate how this analysis can be adapted to the study of behaviour, we studied the position of 40 males and 40 virgins females in their glass tubes.
+The male data was collected from our previous study \cite{geissmann_ethoscopes:_2017} (controls in figure~5M-P).
+We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled it at 0.1 Hz.
+Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
+Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.
+
+\begin{figure}[!h]
+	%  	\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
+	\caption{{\bf Wavelet analysis of positional data.}
+		A: Raw position data for a representative male (top) and virgin female (bottom) \emph{Drosophila} over five days, in black. 
+		The thick coloured lines show the local average position.
+		The green rectangles in the background shows the two time windows selected for B and C.
+		B: Close up of A, showing position over one hour, in the beginning of the L phase ($t \in [24, 25]h$).
+		C: Close up of A, showing position over one hour, in the middle of the L phase ($t \in [30, 31]h$).
+		D: Continuous wavelet transform spectrogram for the two representative animals.
+		E: Average spectrogram across 40 males and 40 females.
+		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average on the period domain).
+	}
+	\label{fig:fig-6}
+\end{figure}
+
+In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp:_2018}.
+We then averaged the result of the five consecutive days in the time/period domain over one circadian day (Fig~\ref{fig:fig-6}D~and~E).
+
+As expected, we observe strong, high period (12~h and 24~h), rhythms of the position.
+In addition, a large amount of signal is detected for low period (around 60~s) events.
+This correspond to the position of animals walking along (back and forward) their tubes in a very paced manner.
+
+Interestingly, in females, this low period pace appears to be frequency modulated during the L phase, suggesting a slower walking speed around ZT6~h.
+In contrast, males shows only a high frequency rhythm around the phase transitions (L to D and D to L).
+Surprisingly, the peak of high frequency rhythm imply a faster pace in males (approximately 60~s) than in females (approximately 120~s) --
+indicating that, when active, males walk faster than females.
+
+This non-exhaustive proof of principle illustrates how analysis of behavioural data can be taken further by adapting the wide range of numerical tools already available in the \texttt{R} ecosystem.
+
 % We present an small example examining the circadian behaviour of 128 fruit flies in recorded in a DAM2 -- a paradigm very widely adopted. 

### 2018-04-04 results.Rnw
another read through the trext

@@
 \section*{Results}
-In order to illustrate the workflow of \texttt{rethomics}, we provide a straightforward and simplified example which could be readily modified for more intricate cases.
+%In order to illustrate the workflow of \texttt{rethomics}, we provide a simplified and reproducible example, which could be readily modified to suit more intricate cases.
+
+In order to illustrate the usefulness of \texttt{rethomics}, we provide two use cases examples.
+The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
+The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with an advanced case of a multi-scale analysis of a periodic behaviour using continuous wavelet transform and ethoscope data \cite{geissmann_ethoscopes:_2017}.
+
+\subsection*{Drosophila Activity Monitor}
+It uses the package \texttt{zeitgebr} which implements a suite of methods to compute autocorrelogram,
+ $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms, and find their peaks.
 
@@
 
-\subsection*{Preprocessing}
+\subsubsection*{Preprocessing}
 The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
@@
 
-Note that as a result, we now have \Sexpr{nrow(dt[meta=T])} ``valid''individuals.
+Note that as a result, we now have \Sexpr{nrow(dt[meta=T])} `valid' individuals.
 
-\subsection*{Double-plotted actograms}
-``Double-plotted actograms'' are a common visualisation of periodicity and rhythmicity in circadian experiments.
+\subsubsection*{Double-plotted actograms}
+`Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
 In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal.
@@
 This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
+
+\subsection*{Multi-scale analysis of position}
+
+One of the challenges of behaviour analysis is the `nesting' of events happening over different time scales.
+In other words, a behavioural variable can be modulated by the circadian rhythm, but also by co-occurring ultradian and infradian rhythms.
+For instance, an animal could have rhythmic bursts of activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable, activity.
+In this example, both rhythms happen at time scales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate in the same analysis.
+Being able to keep frequency information over multiple scales is however important in some cases.
+In particular, when interested in the frequency modulation of one rhythm by another --
+that is, if the periodicity of a high frequency rhythm itself can be driven by a lower frequency one.
+%For instance, the frequency of rhythmic activity bursts could itself be regulated by another rhythm such as the circadian rhythm.
+
+The problem of understanding time series at different scales in not uncommon in fields such as economics\cite{aguiar-conraria_business_2011}, climate sciences\cite{lau_climate_1995} and ecology\cite{cazelles_wavelet_2008} where variables are governed by multiple underlying rhythms (\emph{e.g.} tidal, daily, yearly and multi-yearly).
+One approach is to study a variable of interest in the time/period domain using, for instance, continuous wavelet transform (CWT)\cite{grossmann_decomposition_1984}.
+
+To illustrate how this analysis can be adapted to the study of behaviour, we studied the position of 40 males and 40 virgins females in their glass tubes.
+The male data was collected from our previous study \cite{geissmann_ethoscopes:_2017} (controls in figure~5M-P).
+We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled it at 0.1 Hz.
+Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
+Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.
+
+\begin{figure}[!h]
+	%  	\includegraphics[width=1\textwidth]{fig/fig-5.pdf}
+	\caption{{\bf Wavelet analysis of positional data.}
+		A: Raw position data for a representative male (top) and virgin female (bottom) \emph{Drosophila} over five days, in black. 
+		The thick coloured lines show the local average position.
+		The green rectangles in the background shows the two time windows selected for B and C.
+		B: Close up of A, showing position over one hour, in the beginning of the L phase ($t \in [24, 25]h$).
+		C: Close up of A, showing position over one hour, in the middle of the L phase ($t \in [30, 31]h$).
+		D: Continuous wavelet transform spectrogram for the two representative animals.
+		E: Average spectrogram across 40 males and 40 females.
+		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average on the period domain).
+	}
+	\label{fig:fig-6}
+\end{figure}
+
+In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp:_2018}.
+We then averaged the result of the five consecutive days in the time/period domain over one circadian day (Fig~\ref{fig:fig-6}D~and~E).
+
+As expected, we observe strong, high period (12~h and 24~h), rhythms of the position.
+In addition, a large amount of signal is detected for low period (around 60~s) events.
+This correspond to the position of animals walking along (back and forward) their tubes in a very paced manner.
+
+Interestingly, in females, this low period pace appears to be frequency modulated during the L phase, suggesting a slower walking speed around ZT6~h.
+In contrast, males shows only a high frequency rhythm around the phase transitions (L to D and D to L).
+Surprisingly, the peak of high frequency rhythm imply a faster pace in males (approximately 60~s) than in females (approximately 120~s) --
+indicating that, when active, males walk faster than females.
+
+This non-exhaustive proof of principle illustrates how analysis of behavioural data can be taken further by adapting the wide range of numerical tools already available in the \texttt{R} ecosystem.
 

### 2018-04-05 manuscript.Rnw
include esteban's feed back. still need to work on fig6

@@
 The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
-Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic:_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker:_2014} animals over long durations (days or weeks).
+Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic:_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker:_2014,robie_mapping_2017} animals over long durations (days or weeks).
 
@@
 and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques important.
+The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques.
 However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
@@
 \texttt{rethomics} is implemented as a collection of packages, altogether offering a solution to import, store, manipulate and visualise large amounts of behavioural results.
-We also present two practical examples of its application to the analysis of circadian rhythm in fruit flies, a widely studied behaviour.
+We also present two practical examples of its application to the analysis of behavioural rhythm in fruit flies, a widely studied behaviour.
 
@@
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
-Visualisation and other results are standard objects, and can therefore be further analysed inside the wide \texttt{R} package ecosystem.
+Numerical results and plots are standard objects that can therefore be further analysed inside the wide \texttt{R} package ecosystem.
 
@@
 \texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}A~and~B).
-This approach guarantees that any data point can be mapped correctly to its parent metadata.
+This approach guarantees that any data point can be mapped correctly to its parent metadata thanks to a shared key (\texttt{id}).
 Furthermore, it allows implicit update of metadata when data is altered.
-For instance, when is data filtered, only the remaining individuals should be in the new metadata. 
+For instance, when data is filtered, only the remaining individuals should be in the new metadata. 
 It is also important that metadata and data can interoperate.
@@
 		Data and metadata are implicitly joined on the \texttt{id} field.
-		Note that the names used in this for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
+		Note that the names used for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
 		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
@@
 In all cases, the user is asked to generate a metadata table (one row per individual). 
-In it, there will be both mandatory and optional columns.
+In it, there will be both mandatory and optional columns (Fig~\ref{fig:fig-2}A).
 The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
@@
 In this respect, the metadata file is a standardised and comprehensive data frame describing an experiment.
-It explicitly lists all treatments and individuals, which facilitates interspersion of conditions (without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
+It explicitly lists all treatments and individuals, which facilitates interspersion of conditions.
+%(without it, users are tempted to simplify their experimental design by, for instance, confounding device/location and treatment).
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
@@
 a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016}.
-\texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise of temporal trends.
+\texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise temporal trends.
 We implemented a set of new `layers' and `scales' that particularly applies to the visualisation of long experiments, with the ability to, for instance, display `double-plotted actograms', periodograms, annotate light and dark phases and wrap time over a given period.
@@
 Several new packages in the \texttt{rethomics} framework are currently envisaged. 
-They include utilities to input new behaviour tracking methods and analyse position and multi-animal interactions.
+They include utilities to input new behaviour tracking methods and analyse multi-animal interactions.
 
@@
 In particular, Han Kim, for his invaluable comments on the early versions of \texttt{rethomics} and his dedicated contribution to the tutorials;
-Maite Ogueta, for making the results data available;
-\TODO{@esteban names of the BA users}
-Alice French, Hannah Jones and Diana Bicazan for their comments as early users;
+Maite Ogueta, for making the DAMS results data available;
+Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
 Marcus Gosh, for his feed back on the manuscript;

### 2018-04-05 manuscript.tex
include esteban's feed back. still need to work on fig6

@@
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
-The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with an advanced case of a multi-scale analysis of a periodic behaviour using continuous wavelet transform and ethoscope data \cite{geissmann_ethoscopes:_2017}.
+The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with the advanced case of a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes:_2017}.
 
@@
 We took two genotypes employed in the study, one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in all clock neurons (TIM/NKCC\textsuperscript{ox}).
-In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six days.
+In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
 The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
-As a prerequisite, we downloaded the data, extracted the zip archive in our working directory.
-We start the analysis by loading the necessary rethomics packages (see availability section for installation instructions):
+As a prerequisite, we downloaded the data and extracted the zip archive in our working directory.
+We start the analysis by loading the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
 
@@
 Importantly, we also keep \texttt{id} as an \emph{unambiguous} unique identifier.
-Indeed, two animals in different experiments may have the same label, but different \texttt{id}s.
-in addition, if the metadata changes -- for instance by the addition or removal of individuals -- the label is likely to change, not the \texttt{id},
-which could lead to confusion.
+Indeed, two animals in separate experiments may have the same label, but different \texttt{id}s.
+in addition, if the metadata changes -- for instance by the addition or removal of individuals -- the label is likely to change, not the \texttt{id}.
+%,which could lead to confusion.
 
@@
 \hlcom{# remove data after death}
-\hlstd{dt} \hlkwb{<-} \hlstd{sleepr}\hlopt{::}\hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
+\hlstd{dt} \hlkwb{<-} \hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
 \end{alltt}
@@
 `Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show the double-plotted actograms of each animal.
-A representative sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
+In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
+A selected sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
 
@@
 %	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
-	\caption{{\bf Visualisation of the periodicity in activity of eight representative animals.}
+	\caption{{\bf Visualisation of the periodicity in activity of eight selected animals.}
 		A: Double-plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
@@
           \hlkwd{.}\hlstd{(}
-            \hlkwc{mean_period} \hlstd{=} \hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
-            \hlkwc{sd_period} \hlstd{=} \hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
+            \hlkwc{mean_period} \hlstd{=} \hlkwd{round}\hlstd{(}\hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),} \hlnum{2}\hlstd{),}
+            \hlkwc{sd_period} \hlstd{=} \hlkwd{round}\hlstd{(}\hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),} \hlnum{2}\hlstd{),}
+            \hlkwc{percent_rhythmic} \hlstd{=} \hlkwd{round}\hlstd{(}\hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)} \hlopt{/} \hlstd{.N,} \hlnum{2}\hlstd{),}
             \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
-            \hlkwc{n} \hlstd{= .N,}
-            \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)} \hlopt{/} \hlstd{.N}
+            \hlkwc{n} \hlstd{= .N}
             \hlstd{),}
@@
 \begin{verbatim}
-##      genotype mean_period sd_period n_rhythmic  n percent_rhythmic
-## 1:   NKCCOX/+    25.22500  3.598495          4 22         18.18182
-## 2: Tim/NKCCOX    26.21429  2.363568         28 30         93.33333
+##      genotype mean_period sd_period percent_rhythmic n_rhythmic  n
+## 1:   NKCCOX/+       25.23      3.60            18.18          4 22
+## 2: Tim/NKCCOX       26.21      2.36            93.33         28 30
 \end{verbatim}
@@
 This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
-We were able to comprehensively analyse the data from a circadian experiment with a few line of code.
+We were able to comprehensively analyse the data from a circadian experiment with a few lines of code.
 This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
@@
 In other words, a behavioural variable can be modulated by the circadian rhythm, but also by co-occurring ultradian and infradian rhythms.
-For instance, an animal could have rhythmic bursts of activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable, activity.
+For instance, an animal could have rhythmic bursts of locomotor activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable.
 In this example, both rhythms happen at time scales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate in the same analysis.
@@
 In particular, when interested in the frequency modulation of one rhythm by another --
-that is, if the periodicity of a high frequency rhythm itself can be driven by a lower frequency one.
+that is, if the periodicity of a high frequency rhythm itself can be function of a lower frequency one.
 %For instance, the frequency of rhythmic activity bursts could itself be regulated by another rhythm such as the circadian rhythm.
@@
 One approach is to study a variable of interest in the time/period domain using, for instance, continuous wavelet transform (CWT)\cite{grossmann_decomposition_1984}.
+In the context of chronobiology, CWT has been suggested as a tool to investigate ultradian rhythms \cite{leise_wavelet_2013}.
 
-To illustrate how this analysis can be adapted to the study of behaviour, we studied the position of 40 males and 40 virgins females in their glass tubes.
-The male data was collected from our previous study \cite{geissmann_ethoscopes:_2017} (controls in figure~5M-P).
-We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled it at 0.1 Hz.
+To illustrate how \texttt{rethomic} integrates with other packages to render type of less traditional analysis possible,
+we performed a wavelet analysis of the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
+We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1 Hz.
 Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
@@
 	\caption{{\bf Wavelet analysis of positional data.}
-		A: Raw position data for a representative male (top) and virgin female (bottom) \emph{Drosophila} over five days, in black. 
-		The thick coloured lines show the local average position.
+		A: Raw position data for a representative female (top) and male (bottom) \emph{Drosophila} over five days, in black. 
+		The thick coloured lines show the average position every two.
 		The green rectangles in the background shows the two time windows selected for B and C.
-		B: Close up of A, showing position over one hour, in the beginning of the L phase ($t \in [24, 25]h$).
-		C: Close up of A, showing position over one hour, in the middle of the L phase ($t \in [30, 31]h$).
+		B: Close up of A, showing position over one hour, in the beginning of the L phase of day 1.
+		C: Close up of A, showing position over one hour, in the middle of the L phase of day 1.
 		D: Continuous wavelet transform spectrogram for the two representative animals.
 		E: Average spectrogram across 40 males and 40 females.
-		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average on the period domain).
+		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average across all time).
+		The male data was collected and described in our previous study \cite{geissmann_ethoscopes:_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
 	}
@@
 In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp:_2018}.
-We then averaged the result of the five consecutive days in the time/period domain over one circadian day (Fig~\ref{fig:fig-6}D~and~E).
+We then averaged the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
+and for the population(Fig~\ref{fig:fig-6}E).
 
-As expected, we observe strong, high period (12~h and 24~h), rhythms of the position.
+As suggested by the slow oscilations of the mean position (Fig~\ref{fig:fig-6}A), we observe strong, high period (12~h and 24~h), rhythms.
 In addition, a large amount of signal is detected for low period (around 60~s) events.

### 2018-04-05 results.Rnw
include esteban's feed back. still need to work on fig6

@@
 # remove data after death
-dt <- sleepr::curate_dead_animals(dt, moving)
+dt <- curate_dead_animals(dt, moving)
 @
@@
           .(
-            mean_period = mean(first_peak_period, na.rm = T) / hours(1),
-            sd_period = sd(first_peak_period, na.rm = T) / hours(1),
+            mean_period = round(mean(first_peak_period, na.rm = T) / hours(1), 2),
+            sd_period = round(sd(first_peak_period, na.rm = T) / hours(1), 2),
+            percent_rhythmic = round(100 * sum(is_rhythmic) / .N, 2),
             n_rhythmic = sum(is_rhythmic),
-            n = .N,
-            percent_rhythmic = 100 * sum(is_rhythmic) / .N
+            n = .N
             ),

### 2018-04-06 manuscript.Rnw
fix a few more typose

@@
 The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
-Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic:_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker:_2014,robie_mapping_2017} animals over long durations (days or weeks).
+Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017} animals over long durations (days or weeks).
 
@@
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
-The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled to build pipelines\cite{leipzig_review_2017}.
+The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled into pipelines\cite{leipzig_review_2017}.
 Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
@@
 At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for a similar community to arise.
-However, some low-level conceptual consistencies and methodological challenges are common between experiments.
+However, some low-level conceptual consistencies and methodological challenges are common across experiments.
 For instance, the results (\emph{i.e.} the `data')  feature a set of long time series (sometimes multivariate and irregular), but also contain a formal description of the treatment applied to each individual, the `metadata'.
@@
 \texttt{rethomics} is implemented as a collection of packages, altogether offering a solution to import, store, manipulate and visualise large amounts of behavioural results.
-We also present two practical examples of its application to the analysis of behavioural rhythm in fruit flies, a widely studied behaviour.
+We also present two practical examples of its application to the analysis of behavioural rhythm in fruit flies, a widely studied subject.
 
 \section*{Design and Implementation}
-\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r:_2017}, which is widely taught and adopted by computational biologists,
+\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}, which is widely taught and adopted by computational biologists,
 as a collection of packages (Fig~\ref{fig:fig-1}).
-This architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse:_2017}, which results in increased testability and maintainability.
+This architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability and maintainability.
 In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package.
-At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table:_2017}.
+At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
@@
 For instance, there could be very long time series, sampled several times per second, over multiple days, for each individual. %(typically, $k_i > 10^8, \forall i \in [1,n]$) 
-In addition, time series can be multivariate, encoding coordinates, orientation, dimensions, activity, colour intensity and so on.
+In addition, time series can be multivariate -- encoding coordinates, orientation, dimensions, activity, colour intensity and so on.
 Furthermore, experiments may  feature a large number of individuals. %(typically $n > 100$).
@@
 Data import packages translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor System (Trikinetics Inc.) data and another one for Ethoscope data\cite{geissmann_ethoscopes:_2017}.
+Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor System (Trikinetics Inc.) data and another one for Ethoscope data\cite{geissmann_ethoscopes_2017}.
 Although the structure of the raw results is very different, conceptually, loading data is very similar.
@@
 To integrate visualisation in \texttt{rethomics}, we implemented \texttt{ggetho},
-a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2:_2016}.
+a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2_2016}.
 \texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise temporal trends.
@@
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
-All packages are continuously integrated and unit tested on several version of \texttt{R} to minimise the risk of present and future issues.
+All packages are continuously integrated and unit tested on several versions of \texttt{R} to minimise the risk of present and future issues.
 

### 2018-04-06 manuscript.tex
fix a few more typose

@@
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
-The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with the advanced case of a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes:_2017}.
+The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with the advanced case of a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
 
@@
 Being able to keep frequency information over multiple scales is however important in some cases.
-In particular, when interested in the frequency modulation of one rhythm by another --
+In particular, when interested in the frequency modulation of a rhythm by another --
 that is, if the periodicity of a high frequency rhythm itself can be function of a lower frequency one.
@@
 
-To illustrate how \texttt{rethomic} integrates with other packages to render type of less traditional analysis possible,
+To illustrate how \texttt{rethomic} integrates with other packages and render such non-mainstream analysis possible,
 we performed a wavelet analysis of the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
-We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1 Hz.
+We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
 Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
@@
 		A: Raw position data for a representative female (top) and male (bottom) \emph{Drosophila} over five days, in black. 
-		The thick coloured lines show the average position every two.
+		The thick coloured lines show the average position every two hours.
 		The green rectangles in the background shows the two time windows selected for B and C.
@@
 		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average across all time).
-		The male data was collected and described in our previous study \cite{geissmann_ethoscopes:_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
+		The male data was collected and described in our previous study \cite{geissmann_ethoscopes_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
 	}
@@
 
-In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp:_2018}.
+In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018}.
 We then averaged the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)

### 2018-04-09 manuscript.Rnw
tara's comments

@@
 % TODO revert for sumbmisson
-% \linenumbers
-\TODO{revert line numbers before submission}
+\linenumbers
+%\TODO{revert line numbers before submission}
 % TODO also embed bib
@@
 The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques.
-However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
+However, when it comes to the subsequent processing of the results, there is no unified programmatic framework that could be used as a set of building blocks in a pipeline.
 
-Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
+Scripting interfaces are the standard in data sciences and statistics since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
@@
 Furthermore, it streamlines the inclusion and analysis of further replicates in the same workflow.
-Indeed, additional replicates can simply be added as new rows -- and the ID of the replicate later used, if needed, as a covariate.
+Indeed, additional replicates can simply be added as new rows -- and the \texttt{id} of the replicate later used, if needed, as a covariate.
 
@@
 To integrate visualisation in \texttt{rethomics}, we implemented \texttt{ggetho},
-a package that offers new tools that extent the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2_2016}.
+a package that offers new tools that extend the widely adopted \texttt{ggplot2}\cite{wickham_ggplot2_2016}.
 \texttt{ggetho} makes full use of the internal \texttt{behavr} structure to summarise temporal trends.
@@
 Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
-Marcus Gosh, for his feed back on the manuscript;
+Marcus Gosh and Tara Kane for their feed back on the manuscript;
 Brenna Williams, for her help to support multi-beam DAMS;

### 2018-04-09 results.Rnw
tara's comments

@@
 The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
+It proceeds by detecting the first time an animal is immobile for more than 1~\% of the time (the default) for at least \texttt{time\_window} seconds and discard any subsequent data.
 
@@
 # remove data after death
-dt <- curate_dead_animals(dt, moving)
+dt <- curate_dead_animals(dt, moving, time_window = days(1.5))
 @
@@
 
-\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in the analysis of the extracted variables.
+\texttt{R} provides one of the richest statistical toolboxes available, which allows users to go deeper in the analysis of the extracted variables.
 One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
@@
 In addition, a large amount of signal is detected for low period (around 60~s) events.
-This correspond to the position of animals walking along (back and forward) their tubes in a very paced manner.
+This corresponds to the position of animals walking along (back and forward) their tubes in a very paced manner.
 

### 2018-04-13 results.Rnw
round final DAM results on two steps

@@
 <<warning=FALSE>>=
-result_dt <- summary_dt[,
+result_dt <- 
+  summary_dt[,
     .(
-            mean_period = round(mean(first_peak_period, na.rm = T) / hours(1), 2),
-            sd_period = round(sd(first_peak_period, na.rm = T) / hours(1), 2),
-            percent_rhythmic = round(100 * sum(is_rhythmic) / .N, 2),
+       mean_period = mean(first_peak_period, na.rm = T) / hours(1),
+       sd_period = sd(first_peak_period, na.rm = T) / hours(1),
+       percent_rhythmic = 100 * sum(is_rhythmic) / .N,
        n_rhythmic = sum(is_rhythmic),
@@
   ]
-result_dt
+
+# to round all numeric columns  to two digits
+result_dt[, lapply(.SD, function(x) 
+				if(is.numeric(x)) round(x, 2) else x
+			)]
 @

### 2018-04-13 manuscript.Rnw
ibid
simple corrections

@@
 %Ethomics, a quantitative and high-throughput approach to animal behaviour, is a new and exciting field.
-The recent development of automatised methods that can score various behaviours on a large number of animals
+The recent development of automatised methods to score various behaviours on a large number of animals
 provides biologists with an unprecedented set of tools to decipher these complex phenotypes. 
 Analysing such data comes with several challenges that are largely shared across acquisition platform and paradigms.
-%However, there is little effort in providing a generic toolbox to specifically analyse multiple and long behavioural time series.
-%To address this limitation, we developed the \texttt{rethomics} framework, 
-%We present the \texttt{rethomics} framework, 
-%a suite of \texttt{R} packages that altogether offers utilities to work through all data manipulation stage:
-%import, store, analyse and visualise behavioural data.
-In the article herein, we present \texttt{rethomics}, a set of \texttt{R} packages that unifies analysis of behavioural dataset in an efficient and flexible manner.
+Here, we present \texttt{rethomics}, a set of \texttt{R} packages that unifies analysis of behavioural dataset in an efficient and flexible manner.
 %We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
 %It is also complemented with a vast ecosystem of open-source packages.
-\texttt{rethomics} offers a computational solution to store, manipulate and visualise large amounts of behavioural data.
+\texttt{rethomics} offers a computational solution to storing, manipulating and visualising large amounts of behavioural data.
 We propose it as a tool to breach the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
-Our software comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
+\texttt{rethomics} comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
 
@@
 % TODO revert for sumbmisson
-\linenumbers
-%\TODO{revert line numbers before submission}
+%\linenumbers
+\TODO{revert line numbers before submission}
 % TODO also embed bib
@@
 
-In the article herein, we describe the \texttt{rethomics} platform, an effort to promote the interaction between behavioural biology and data science.
-\texttt{rethomics} is implemented as a collection of packages, altogether offering a solution to import, store, manipulate and visualise large amounts of behavioural results.
+Here, we describe the \texttt{rethomics} platform, an effort to promote the interaction between behavioural biologists and data scientists.
+\texttt{rethomics} is implemented as a collection of interconnected packages, offering solutions to importing, storing, manipulating and visualising large amounts of behavioural results.
 We also present two practical examples of its application to the analysis of behavioural rhythm in fruit flies, a widely studied subject.
@@
 as a collection of packages (Fig~\ref{fig:fig-1}).
-This architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability and maintainability.
-In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package.
+Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
+In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
 At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
@@
 
-Ethomics results can easily scale and data structure therefore gains from being computationally efficient – both in term of memory footprint and processing speed.
+Ethomics results can easily scale and data structure therefore gains from being computationally efficient -- both in term of memory footprint and processing speed.
 For instance, there could be very long time series, sampled several times per second, over multiple days, for each individual. %(typically, $k_i > 10^8, \forall i \in [1,n]$) 
@@
 For instance, when data is filtered, only the remaining individuals should be in the new metadata. 
-It is also important that metadata and data can interoperate.
-For example, when one wants to update a variable according to the value of a metavariable (say, alter the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
+It is also important that metadata and data can interoperate --
+for example, when updating a variable according to the value of a metavariable (say, alter the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 \subsection*{Data import}
-Data import packages translate results from a recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide a package to read single or multi-beam Drosophila Activity Monitor System (Trikinetics Inc.) data and another one for Ethoscope data\cite{geissmann_ethoscopes_2017}.
+Data import packages translate results from a specific recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
+Currently, we provide two packages: one to import results from single and multi-beam Drosophila Activity Monitor Systems (Trikinetics Inc.) 
+and another one for Ethoscopes \cite{geissmann_ethoscopes_2017}.
 Although the structure of the raw results is very different, conceptually, loading data is very similar.
-In all cases, the user is asked to generate a metadata table (one row per individual). 
-In it, there will be both mandatory and optional columns (Fig~\ref{fig:fig-2}A).
+In all cases, users must provide a metadata table, with one row per individual,
+and featuring both mandatory and optional columns (Fig~\ref{fig:fig-2}A).
 The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
@@
 Brenna Williams, for her help to support multi-beam DAMS;
-Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework;
-and Robert Dickinson, for his support.
+Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework.
 

### 2018-04-13 results.Rnw
simple corrections

@@
 
-In order to illustrate the usefulness of \texttt{rethomics}, we provide two use cases examples.
+In order to illustrate the usefulness of \texttt{rethomics}, we provide two carefully annotated examples.
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
@@
 \subsection*{Drosophila Activity Monitor}
-It uses the package \texttt{zeitgebr} which implements a suite of methods to compute autocorrelogram,
- $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms, and find their peaks.
+The \texttt{zeitgebr} package implements a comprehensive suite of methods to analyse circadian rhythms, 
+including the computation of autocorrelograms, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms,
+and peak detection.
 
@@
 
-We took two genotypes employed in the study, one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in all clock neurons (TIM/NKCC\textsuperscript{ox}).
-In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
-The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
-As a prerequisite, we downloaded the data and extracted the zip archive in our working directory.
-We start the analysis by loading the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
+Here, we guide the reader through the analysis of two of the genotypes employed in that study;
+one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in clock neurons (TIM/NKCC\textsuperscript{ox}).
+In particular, we outline the necessary steps to analyse two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
+The \texttt{metadata.csv} file as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+
+We start by downloading the data and extract the zip archive in our working directory.
+Then we load the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
 
@@
 \subsubsection*{Preprocessing}
-The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
+The two original replicates do not have the same baseline duration.
+To include them in the same analysis, we decided to express time relative to the experimental perturbation:
 the transition from LD to LL. 
@@
 In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
-It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:
+It is \texttt{TRUE} when \texttt{activity} is greater than zero and \texttt{FALSE} otherwise:
 
@@
 @
-
-To simplify visualisation, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+The \texttt{id} is a long and exhaustive string of character, which incidentally 
+makes it difficult to read and display as a label on a plot.
+To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
-Importantly, we also keep \texttt{id} as an \emph{unambiguous} unique identifier.
+Importantly, we also retain \texttt{id} as an \emph{unambiguous} unique identifier.
 Indeed, two animals in separate experiments may have the same label, but different \texttt{id}s.
@@
 \subsubsection*{Curation}
-It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
+It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly.
+For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
 
@@
 
-The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
+The activity of dead or escaped animals is falsely scored as long series of zeros,
+which may be erroneously interpreted as inactivity (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
 The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
@@
 
-In addition, we can trim our data to have the same number of days across experiments and individuals.
+In addition, we can trim the data to the same number of days across experiments and individuals.
 <<cache=TRUE>>=
@@
 
-For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have \emph{at least five days in LL}.
+For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have lived for \emph{at least five days in LL}.
 An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.
@@
 \subsubsection*{Double-plotted actograms}
-`Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
+`Double-plotted actograms' are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
 In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
@@
 	\caption{{\bf Visualisation of the periodicity in activity of eight selected animals.}
-		A: Double-plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
+		A: Double-plotted actograms showing all activity during experiment.
+		Time is defined relative to the transition from LD to LL (at day 0).
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
@@
 
-As shown in the original study, double-plotted actograms and periodograms suggest that NKCC\textsuperscript{ox}/+ flies are mostly 
-arhythmic in LL whilst Tim/NKCC\textsuperscript{ox} appear to have a consistent, long-period rhythm.
-To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
+As shown in the original study\cite{buhl_quasimodo_2016}, double-plotted actograms and periodograms suggest that NKCC\textsuperscript{ox}/+ flies are mostly 
+arhythmic in LL whilst Tim/NKCC\textsuperscript{ox} appear to have a long-period rhythm.
+To visualise this difference at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
 
@@
 
-To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each group (see Fig~\ref{fig:fig-5}B).
+To further quantify this difference, we opt to show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each group (see Fig~\ref{fig:fig-5}B).
 Then, we can compare the average value of the peak for the rhythmic animals (see Fig~\ref{fig:fig-5}C).
@@
 \texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics. 
-It can therefore be used directly by \texttt{ggplot} and other tools:
+It can therefore be used directly by \texttt{ggplot}:
 
@@
 
-\texttt{R} provides one of the richest statistical toolboxes available, which allows users to go deeper in the analysis of the extracted variables.
-One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model:
+\texttt{R} provides one of the richest statistical toolboxes available.
+Using base \texttt{R} we could perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes, or,
+like in this case, fit a binomial generalised linear model:
 
@@
 fit <- glm(is_rhythmic ~ genotype, summary_dt, family = "binomial")
-
 summary(fit)$coefficients
@@
 
-This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
-We were able to comprehensively analyse the data from a circadian experiment with a few lines of code.
-This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
+The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state of the art statistics.
+We were able to comprehensively analyse the data from a circadian experiment with a few lines of code,
+presenting a workflow that applies equally well to much larger datasets.
 
@@
 We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
-Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
+Our variable of interest is the position of animals in their tube (from the food end, $Position = 0$,  to the cotton end, $Position = 1$).
 Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.
@@
 
-As suggested by the slow oscilations of the mean position (Fig~\ref{fig:fig-6}A), we observe strong, high period (12~h and 24~h), rhythms.
-In addition, a large amount of signal is detected for low period (around 60~s) events.
-This corresponds to the position of animals walking along (back and forward) their tubes in a very paced manner.
+As suggested by the slow oscillations of the mean position (Fig~\ref{fig:fig-6}A), we observe peaks in power corresponding to high-period (12~h and 24~h) rhythms.
+In addition, a large amount of signal is detected for low period (around 60~s) events -- 
+likely corresponding to the position of animals walking along (back and forward) their tubes in a very paced manner.
 

### 2018-04-13 results.Rnw
more small changes in phrasing2

@@
 \subsubsection*{Preprocessing}
-The two original replicates do not have the same baseline duration.
-To include them in the same analysis, we decided to express time relative to the experimental perturbation:
-the transition from LD to LL. 
-Therefore, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
+Since the two original replicates do not have the same baseline duration and we want to analyse them together,
+we align their respective times to the experimental perturbation: the transition from LD to LL ($t = 0$). 
+This is achieved by subtracting the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
 This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
@@
 \subsubsection*{Curation}
-It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly.
+It is important to visualise an overview of how each individual behaved and, if necessary, amend the metadata accordingly.
 For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
@@
 
-The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state of the art statistics.
+The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state-of-the-art statistics.
 We were able to comprehensively analyse the data from a circadian experiment with a few lines of code,

### 2018-04-15 manuscript.Rnw
intro bullet point a first part

@@
 \section*{Introduction}
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, its internal state and its environment.
-In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased\cite{stephens_big_2015}.
-The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
-Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017} animals over long durations (days or weeks).
+%The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, its internal state and its environment.
 
-The availability of large amounts of data paves the way for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles
-and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques.
-However, when it comes to the subsequent processing of the results, there is no unified programmatic framework that could be used as a set of building blocks in a pipeline.
+The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been quiestion of prime interest.
+In the 1960s researchers started to identify genetic determinant of behaviours.
+OUr ability to understand genetics has been enpowered by new high throuput methods.
+The scoring of behaviour as well is undergoing a transition towards high througput (ethomics, CE).
 
-Scripting interfaces are the standard in data sciences and statistics since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
+%In the last few decades, our ability to sequence vast quantities of genetic and phenoypic data has tremendously increased\cite{stephens_big_2015}.
+%The scoring of various behavioural variables is certainly not an exception to this trend.
+%High-throughput approaches to ethology has been proposed as a new discipline, prompting the concepts of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
+
+%In the 1960's, Seymounr Benzer had shown:  behavioural genetics
+The availability of large amounts of data, in combination with the use of methods borrowed from the data sciences, allow for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles and ultimately a better understanding of their underlying biology\cite{brown_ethology_2018, berman_measuring_2018}.
+
+%
+
+Although general questions regarding the environmental, evolutionary, neural and genetic determinants of behaviours are shared within the community,
+the multiplicity of model organisms, hypotheses and paradigms has led the existence of a very diverse palette of specific recording techniques.
+Some tools were developed to, for instance:
+continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
+score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
+and study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
+Whilst most recording platforms are unrelated to each other, there are also some attempts to build general purpose tools that can be adapted by researchers to
+suit their specific goals \cite{branson_high-throughput_2009,kabra_jaaba_2013,lopes_bonsai_2015,geissmann_ethoscopes_2017}.
+However, when it comes to the subsequent analysis of the generated results, there is still no unified programmatic framework that could be used as a set of building blocks in a pipeline.
+
+The fields of structural biology and  bioinformatics are good examples of communities that have taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled into pipelines\cite{leipzig_review_2017}.
+In these research areas, which are closely linked to data sciences and statistics, scripting interfaces are the standard since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
-The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled into pipelines\cite{leipzig_review_2017}.
 Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.

### 2018-04-15 manuscript.tex
intro bullet point a first part

@@
 % TODO revert for sumbmisson
-% \linenumbers
+%\linenumbers
 \TODO{revert line numbers before submission}
@@
 \section*{Introduction}
-The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, its internal state and its environment.
-In the last few decades, our ability to record vast quantities of various phenotypical data has tremendously increased\cite{stephens_big_2015}.
-The scoring of behaviours is certainly not an exception to this trend \cite{reiser_ethomics_2009}.
-Indeed, many platforms have been developed in order to allow biologists to continuously record behaviours such as activity\cite{faville_how_2015}, position\cite{pelkowski_novel_2011} and feeding\cite{itskov_automated_2014,ro_flic_2014} of single or multiple\cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017} animals over long durations (days or weeks).
+%The behaviour of an animal is a complex phenotypical manifestation of the interaction between its nervous system, its internal state and its environment.
 
-The availability of large amounts of data paves the way for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles
-and ultimately a better understanding of their underlying biology\cite{brown_study_2017,berman_measuring_2018}.
-The multiplicity of model organisms, hypotheses and paradigms justifies the existence of a diverse palette of recording techniques.
-However, when it comes to the subsequent processing of the results, there is no unified, programmatic, framework that could be used as a set of building blocks in a pipeline.
+The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been quiestion of prime interest.
+In the 1960s researchers started to identify genetic determinant of behaviours.
+OUr ability to understand genetics has been enpowered by new high throuput methods.
+The scoring of behaviour as well is undergoing a transition towards high througput (ethomics, CE).
 
-Scripting interfaces are the standard in data sciences and statistics since they help delivering reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
+%In the last few decades, our ability to sequence vast quantities of genetic and phenoypic data has tremendously increased\cite{stephens_big_2015}.
+%The scoring of various behavioural variables is certainly not an exception to this trend.
+%High-throughput approaches to ethology has been proposed as a new discipline, prompting the concepts of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
+
+%In the 1960's, Seymounr Benzer had shown:  behavioural genetics
+The availability of large amounts of data, in combination with the use of methods borrowed from the data sciences, allow for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles and ultimately a better understanding of their underlying biology\cite{brown_ethology_2018, berman_measuring_2018}.
+
+%
+
+Although general questions regarding the environmental, evolutionary, neural and genetic determinants of behaviours are shared within the community,
+the multiplicity of model organisms, hypotheses and paradigms has led the existence of a very diverse palette of specific recording techniques.
+Some tools were developed to, for instance:
+continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
+score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
+and study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
+Whilst most recording platforms are unrelated to each other, there are also some attempts to build general purpose tools that can be adapted by researchers to
+suit their specific goals \cite{branson_high-throughput_2009,kabra_jaaba_2013,lopes_bonsai_2015,geissmann_ethoscopes_2017}.
+However, when it comes to the subsequent analysis of the generated results, there is still no unified programmatic framework that could be used as a set of building blocks in a pipeline.
+
+The fields of structural biology and  bioinformatics are good examples of communities that have taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled into pipelines\cite{leipzig_review_2017}.
+In these research areas, which are closely linked to data sciences and statistics, scripting interfaces are the standard since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
-The field of bioinformatics is a good example of a community that has taken advantage of sharing standard files formats, modular command line tools\cite{roumpeka_review_2017} and software packages\cite{huber_orchestrating_2015} that can be assembled into pipelines\cite{leipzig_review_2017}.
 Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
@@
 \subsubsection*{Preprocessing}
-The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
-the transition from LD to LL. 
-Therefore, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
+Since the two original replicates do not have the same baseline duration and we want to analyse them together,
+we align their respective times to the experimental perturbation: the transition from LD to LL ($t = 0$). 
+This is achieved by subtracting the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
 This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
 In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
-It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:
+It is \texttt{TRUE} when \texttt{activity} is greater than zero and \texttt{FALSE} otherwise:
 
@@
 \end{knitrout}
-
-To simplify visualisation, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+The \texttt{id} is a long and exhaustive string of character, which incidentally 
+makes it difficult to read and display as a label on a plot.
+To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
-Importantly, we also keep \texttt{id} as an \emph{unambiguous} unique identifier.
+Importantly, we also retain \texttt{id} as an \emph{unambiguous} unique identifier.
 Indeed, two animals in separate experiments may have the same label, but different \texttt{id}s.
@@
 \subsubsection*{Curation}
-It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
+It is important to visualise an overview of how each individual behaved and, if necessary, amend the metadata accordingly.
+For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
 
@@
 
-The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
+The activity of dead or escaped animals is falsely scored as long series of zeros,
+which may be erroneously interpreted as inactivity (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
 The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
+It proceeds by detecting the first time an animal is immobile for more than 1~\% of the time (the default) for at least \texttt{time\_window} seconds and discard any subsequent data.
 
@@
 \hlcom{# remove data after death}
-\hlstd{dt} \hlkwb{<-} \hlkwd{curate_dead_animals}\hlstd{(dt, moving)}
+\hlstd{dt} \hlkwb{<-} \hlkwd{curate_dead_animals}\hlstd{(dt, moving,} \hlkwc{time_window} \hlstd{=} \hlkwd{days}\hlstd{(}\hlnum{1.5}\hlstd{))}
 \end{alltt}
@@
 
-In addition, we can trim our data to have the same number of days across experiments and individuals.
+In addition, we can trim the data to the same number of days across experiments and individuals.
 \begin{knitrout}
@@
 ## behavr table with:
-##  52	individuals
+##  53	individuals
 ##  9	metavariables
 ##  3	variables
-##  1.19546e+05	measurements
+##  1.2184e+05	measurements
 ##  1	key (id)
@@
 
-Note that as a result, we now have 52 `valid' individuals.
+Note that as a result, we now have 53 `valid' individuals.
 
 \subsubsection*{Double-plotted actograms}
-`Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
+`Double-plotted actograms' are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
 In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
@@
 		Titles on top of each facet refer to the label allocated to each individual.
-		See \nameref{S1-Fig} for all 52 animals.
+		See \nameref{S1-Fig} for all 53 animals.
 	}
@@
 
-\texttt{R} provides one of the richest statistical toolbox available, which allows users to go deeper in the analysis of the extracted variables.
-One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
-To address the same question, we fit a binomial generalised linear model:
+\texttt{R} provides one of the richest statistical toolboxes available.
+Using base \texttt{R} we could perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes, or,
+like in this case, fit a binomial generalised linear model:
 
@@
 \hlstd{fit} \hlkwb{<-} \hlkwd{glm}\hlstd{(is_rhythmic} \hlopt{~} \hlstd{genotype, summary_dt,} \hlkwc{family} \hlstd{=} \hlstr{"binomial"}\hlstd{)}
-
 \hlkwd{summary}\hlstd{(fit)}\hlopt{$}\hlstd{coefficients}
@@
 ## (Intercept)        -1.504077  0.5527708 -2.720978 6.508902e-03
-## genotypeTim/NKCCOX  4.143135  0.9172057  4.517127 6.268433e-06
+## genotypeTim/NKCCOX  4.178226  0.9165333  4.558728 5.146439e-06
 \end{verbatim}
@@
 
-The result shows a strong positive effect of genotype Tim/NKCC\textsuperscript{ox} on the probability of being rhythmic ($p$-value $6.27 \times 10^{-06}$):
+The result shows a strong positive effect of genotype Tim/NKCC\textsuperscript{ox} on the probability of being rhythmic ($p$-value $5.15 \times 10^{-06}$):
 
@@
 \begin{alltt}
-\hlstd{result_dt} \hlkwb{<-} \hlstd{summary_dt[,}
+\hlstd{result_dt} \hlkwb{<-}
+  \hlstd{summary_dt[,}
     \hlkwd{.}\hlstd{(}
-            \hlkwc{mean_period} \hlstd{=} \hlkwd{round}\hlstd{(}\hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),} \hlnum{2}\hlstd{),}
-            \hlkwc{sd_period} \hlstd{=} \hlkwd{round}\hlstd{(}\hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),} \hlnum{2}\hlstd{),}
-            \hlkwc{percent_rhythmic} \hlstd{=} \hlkwd{round}\hlstd{(}\hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)} \hlopt{/} \hlstd{.N,} \hlnum{2}\hlstd{),}
+       \hlkwc{mean_period} \hlstd{=} \hlkwd{mean}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
+       \hlkwc{sd_period} \hlstd{=} \hlkwd{sd}\hlstd{(first_peak_period,} \hlkwc{na.rm} \hlstd{= T)} \hlopt{/} \hlkwd{hours}\hlstd{(}\hlnum{1}\hlstd{),}
+       \hlkwc{percent_rhythmic} \hlstd{=} \hlnum{100} \hlopt{*} \hlkwd{sum}\hlstd{(is_rhythmic)} \hlopt{/} \hlstd{.N,}
        \hlkwc{n_rhythmic} \hlstd{=} \hlkwd{sum}\hlstd{(is_rhythmic),}
@@
   \hlstd{]}
-\hlstd{result_dt}
+
+\hlcom{# to round all numeric columns  to two digits}
+\hlstd{result_dt[,} \hlkwd{lapply}\hlstd{(.SD,} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)}
+                                \hlkwa{if}\hlstd{(}\hlkwd{is.numeric}\hlstd{(x))} \hlkwd{round}\hlstd{(x,} \hlnum{2}\hlstd{)} \hlkwa{else} \hlstd{x}
+                        \hlstd{)]}
 \end{alltt}
@@
 ## 1:   NKCCOX/+       25.23      3.60            18.18          4 22
-## 2: Tim/NKCCOX       26.21      2.36            93.33         28 30
+## 2: Tim/NKCCOX       26.22      2.32            93.55         29 31
 \end{verbatim}
@@
 
-This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
-We were able to comprehensively analyse the data from a circadian experiment with a few lines of code.
-This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
+The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state-of-the-art statistics.
+We were able to comprehensively analyse the data from a circadian experiment with a few lines of code,
+presenting a workflow that applies equally well to much larger datasets.
 
@@
 
-As suggested by the slow oscilations of the mean position (Fig~\ref{fig:fig-6}A), we observe strong, high period (12~h and 24~h), rhythms.
-In addition, a large amount of signal is detected for low period (around 60~s) events.
-This correspond to the position of animals walking along (back and forward) their tubes in a very paced manner.
+As suggested by the slow oscillations of the mean position (Fig~\ref{fig:fig-6}A), we observe peaks in power corresponding to high-period (12~h and 24~h) rhythms.
+In addition, a large amount of signal is detected for low period (around 60~s) events -- 
+likely corresponding to the position of animals walking along (back and forward) their tubes in a very paced manner.
 
@@
 Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
-Marcus Gosh, for his feed back on the manuscript;
+Marcus Gosh and Tara Kane for their feed back on the manuscript;
 Brenna Williams, for her help to support multi-beam DAMS;
-Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework;
-and Robert Dickinson, for his support.
+Patrick Kr{\"a}tschmer, for his time discussing the conceptual framework.
 

### 2018-04-15 manuscript.Rnw
intro bullet point a first part2

@@
 
-The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been quiestion of prime interest.
-In the 1960s researchers started to identify genetic determinant of behaviours.
-OUr ability to understand genetics has been enpowered by new high throuput methods.
+The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been question of prime interest.
+In the 1960-70s genetics and behaviour inerpenetrated as researchers started to identify genetic determinant of behaviours.
+OUr ability to understand genetics and neurobiology has been enpowered by new high throughput methods and data analysis.
 The scoring of behaviour as well is undergoing a transition towards high througput (ethomics, CE).

### 2018-04-16 manuscript.Rnw
first part of intro ++

--- 2018-04-15 manuscript.Rnw
@@
 Analysing such data comes with several challenges that are largely shared across acquisition platform and paradigms.
-Here, we present \texttt{rethomics}, a set of \texttt{R} packages that unifies analysis of behavioural dataset in an efficient and flexible manner.
+Here, we present \texttt{rethomics}, a set of \texttt{R} packages that unifies the analysis of behavioural datasets in an efficient and flexible manner.
 %We implemented it in \texttt{R}\cite{r_core_team_r:_2017} since it is widely taught and adopted by computational biologists.
@@
 % TODO revert for sumbmisson
-%\linenumbers
+\linenumbers
 \TODO{revert line numbers before submission}
@@
 
-The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been quiestion of prime interest.
-In the 1960s researchers started to identify genetic determinant of behaviours.
-OUr ability to understand genetics has been enpowered by new high throuput methods.
-The scoring of behaviour as well is undergoing a transition towards high througput (ethomics, CE).
+The biological significance and determinism of animal behaviours have long been a subject of prime scientific interest.
+In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant,
+which breached the gap between genetics and the study of behaviour\citationneeded{}.
+In the last few decades, our ability to acquire and analyse vast amount of biological data has tremendously increased\cite{stephens_big_2015},
+which is deeply transforming both genetics\citationneeded{} and neurobiology\citationneeded{}.
+%The rise of high-throughput methods and computational approaches, have particularly transformed genetics and neurobiology.
+In fact, ethology itself is undergoing its own transition towards data sciences, which has prompted the terms of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
+It is now accepted that the study of behaviour can also benefit from quantitative sciences such as mathematics, physics and computational linguistics \cite{brown_ethology_2018, berman_measuring_2018}.
 
+%The availability of large amounts of data, in combination with the use of methods borrowed from the data sciences, allow for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles and ultimately a better understanding of their underlying biology\cite{brown_ethology_2018, berman_measuring_2018}.
 %In the last few decades, our ability to sequence vast quantities of genetic and phenoypic data has tremendously increased\cite{stephens_big_2015}.
@@
 %In the 1960's, Seymounr Benzer had shown:  behavioural genetics
-The availability of large amounts of data, in combination with the use of methods borrowed from the data sciences, allow for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles and ultimately a better understanding of their underlying biology\cite{brown_ethology_2018, berman_measuring_2018}.
-
 %
@@
 Although general questions regarding the environmental, evolutionary, neural and genetic determinants of behaviours are shared within the community,
-the multiplicity of model organisms, hypotheses and paradigms has led the existence of a very diverse palette of specific recording techniques.
-Some tools were developed to, for instance:
-continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
-score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
-and study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
+the multiplicity of model organisms, hypotheses and paradigms has led to the existence of a very diverse palette of specific recording techniques.
+Some tools were developed, for instance:
+to continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
+to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
+and to study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
 Whilst most recording platforms are unrelated to each other, there are also some attempts to build general purpose tools that can be adapted by researchers to
@@
 In these research areas, which are closely linked to data sciences and statistics, scripting interfaces are the standard since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
-In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
+In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
 Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
@@
 In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
-At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
+At the core of \texttt{rethomics} lies the \texttt{behavr} data structure, an object used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.

### 2018-04-16 results.Rnw
first part of intro ++

@@
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
-The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} with the advanced case of a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
+The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} to perform a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
 
-\subsection*{Drosophila Activity Monitor}
+\subsection*{Canonical circadian analysis in \emph{Drosophila}}
 The \texttt{zeitgebr} package implements a comprehensive suite of methods to analyse circadian rhythms, 
@@
 
-The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
-Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
-We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+%The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
+%Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
+%We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+%Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
+%In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+%and that the endogenous period in LL is longer than 24 hours.
+
+The study of the rhythmical activity of fruit flies has played a crucial role in the development of circadian biology\citationneeded{}.
+To date, most of the behavioural data used in the field is acquired through the Drosophila Activity Monitor System (DAMS).
+The package \texttt{damr} was developed to import DAMS results in the \texttt{rethomics} framework, which we envision will be a common use case. 
+To illustrate the ability of \texttt{rethomics} to analyse pre-existing results,
+we gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
@@
 Indeed, two animals in separate experiments may have the same label, but different \texttt{id}s.
-in addition, if the metadata changes -- for instance by the addition or removal of individuals -- the label is likely to change, not the \texttt{id}.
+In addition, if the metadata changes -- for instance by the addition or removal of individuals -- the label is likely to change, not the \texttt{id}.
 %,which could lead to confusion.

### 2018-04-17 manuscript.tex
another few touches

@@
 \bibliographystyle{plos2015}
+%\bibliographystyle{apalike}
 
@@
 \texttt{rethomics} offers a computational solution to storing, manipulating and visualising large amounts of behavioural data.
-We propose it as a tool to breach the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+We propose it as a tool to bridged the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
@@
 
-The significANCE AND determinism OF ANIMAL BEHAVIOUR has long been quiestion of prime interest.
-In the 1960s researchers started to identify genetic determinant of behaviours.
-OUr ability to understand genetics has been enpowered by new high throuput methods.
-The scoring of behaviour as well is undergoing a transition towards high througput (ethomics, CE).
+The biological significance and determinism of animal behaviours have long been a subject of prime scientific interest.
+In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant, which connected genetics and behaviour\cite{sokolowski_drosophila_2001}.
+In the last few decades, our ability to acquire and analyse vast amount of biological data has tremendously increased\cite{stephens_big_2015},
+which has been deeply transforming both genetics\cite{schatz_biological_2015} and neuroscience\cite{sejnowski_putting_2014}.
+%The rise of high-throughput methods and computational approaches, have particularly transformed genetics and neurobiology.
+In fact, ethology itself is undergoing its own transition towards data sciences, which has prompted the terms of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
+It is now accepted that the study of behaviour can also benefit from quantitative sciences such as machine learning, physics and computational linguistics \cite{brown_ethology_2018, berman_measuring_2018}.
 
+%The availability of large amounts of data, in combination with the use of methods borrowed from the data sciences, allow for in-depth quantitative analyses which, in turn, leads to the characterisation of new principles and ultimately a better understanding of their underlying biology\cite{brown_ethology_2018, berman_measuring_2018}.
 %In the last few decades, our ability to sequence vast quantities of genetic and phenoypic data has tremendously increased\cite{stephens_big_2015}.
@@
 Although general questions regarding the environmental, evolutionary, neural and genetic determinants of behaviours are shared within the community,
-the multiplicity of model organisms, hypotheses and paradigms has led the existence of a very diverse palette of specific recording techniques.
-Some tools were developed to, for instance:
-continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
-score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
-and study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
+the multiplicity of model organisms, hypotheses and paradigms has led to the existence of a very diverse palette of specific recording techniques.
+Some tools were developed, for instance:
+to continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
+to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014} and courtship\cite{tsai_image_2012};
+and to study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
 Whilst most recording platforms are unrelated to each other, there are also some attempts to build general purpose tools that can be adapted by researchers to
@@
 In these research areas, which are closely linked to data sciences and statistics, scripting interfaces are the standard since they help to deliver reproducible results \cite{peng_reproducible_2011,stodden_toward_2013}.
-In addition, they can be used on remote resources such as computer clusters, which makes them more scalable in the context of `big data'\cite{hashem_rise_2015}.
-Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
+In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
+Since many aspect of behaviour analysis are also becoming increasingly dependent on data sciences, the development of such common tools and data structures would be very valuable.
 
@@
 In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
-At the core of \texttt{rethomics}, lies the \texttt{behavr} package, used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
+At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
@@
 
-The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
-Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
-We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+%The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
+%Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
+%We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+%Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
+%In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+%and that the endogenous period in LL is longer than 24 hours.
+
+The study of the rhythmical activity of fruit flies has played a crucial role in the development of circadian biology\cite{dubowy_circadian_2017}.
+To date, most of the behavioural data used in the field is acquired through the Drosophila Activity Monitor System (DAMS).
+The package \texttt{damr} was developed to import DAMS results in the \texttt{rethomics} framework, which we envision will be a common use case. 
+To illustrate the ability of \texttt{rethomics} to analyse pre-existing results,
+we gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
@@
 We start by downloading the data and extract the zip archive in our working directory.
-Then we load the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
+Then, we load the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
 
@@
 			Tile plot showing the fraction of time spent moving as a colour intensity.
-			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes.
+			Each individual is represented by a row and time, on the x-axis, is binned in 30 minutes consecutive epochs.
 			A: Uncurated raw data.
 			B: Data after the curation step. Time was trimmed and data from dead animals removed. 
-			Red `\textbf{+}' symbols show animals that were removed from the subsequent analysis as they had less than five complete days in LL.}
+			The red `\textbf{+}' symbols show animals that were removed from the subsequent analysis as they did not survive five complete days in LL.}
 	\label{fig:fig-3}
@@
 The activity of dead or escaped animals is falsely scored as long series of zeros,
-which may be erroneously interpreted as inactivity (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
+which may be erroneously interpreted as inactivity (see, for instance, individual labelled \texttt{30} and \texttt{18} in Fig~\ref{fig:fig-3}A).
 The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
-It proceeds by detecting the first time an animal is immobile for more than 1~\% of the time (the default) for at least \texttt{time\_window} seconds and discard any subsequent data.
+It proceeds by detecting the first time an animal is immobile for more than 99~\% of the time (the default) for at least \texttt{time\_window} seconds and then discard any subsequent data.
 
@@
 \begin{alltt}
-\hlcom{# we also show a subset of this figure in 4A}
+\hlcom{# we also show a subset of this figure in Fig 4A}
 \hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
-            \hlcom{# bars n the z axis could }
             \hlcom{# one could also use stat_tile_etho}
@@
 \hlstd{dt_ll} \hlkwb{<-} \hlstd{dt[t} \hlopt{>} \hlkwd{days}\hlstd{(}\hlnum{1}\hlstd{)]}
-\hlcom{# compute chi sqr periodogram }
+\hlcom{# compute chi square periodograms }
 \hlstd{per_dt} \hlkwb{<-} \hlkwd{periodogram}\hlstd{(moving,}
@@
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
-		The blue cross represents the first peak (if present) above the significance threshold (red line).
+		The blue cross represents the highest peak (if present) above the significance threshold (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
@@
 \begin{alltt}
-\hlstd{summary_dt} \hlkwb{<-}
-    \hlstd{per_dt[,}
+\hlstd{summary_dt} \hlkwb{<-} \hlstd{per_dt[,}
                 \hlkwd{.}\hlstd{(}
@@
               \hlcom{# shows all individuals as points}
-              \hlcom{# the size of the point expresses the power of the peak}
+              \hlcom{# the size of the point represents the power of the peak}
               \hlkwd{geom_jitter}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{colour} \hlstd{= genotype,}
@@
                               \hlkwc{alpha} \hlstd{=} \hlnum{0.67}\hlstd{)} \hlopt{+}
-              \hlcom{# We would like to convert time in hour}
+              \hlcom{# We would like to show time in hours}
               \hlkwd{scale_y_hours}\hlstd{(}\hlstr{"Period"}\hlstd{)}
@@
 \hlcom{# to round all numeric columns  to two digits}
-\hlstd{result_dt[,} \hlkwd{lapply}\hlstd{(.SD,} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)}
-                                \hlkwa{if}\hlstd{(}\hlkwd{is.numeric}\hlstd{(x))} \hlkwd{round}\hlstd{(x,} \hlnum{2}\hlstd{)} \hlkwa{else} \hlstd{x}
-                        \hlstd{)]}
+\hlstd{result_dt[,} \hlkwd{lapply}\hlstd{(.SD,}
+                   \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwa{if}\hlstd{(}\hlkwd{is.numeric}\hlstd{(x))} \hlkwd{round}\hlstd{(x,} \hlnum{2}\hlstd{)} \hlkwa{else} \hlstd{x}
+                   \hlstd{)}
+           \hlstd{]}
 \end{alltt}
@@
 As suggested by the slow oscillations of the mean position (Fig~\ref{fig:fig-6}A), we observe peaks in power corresponding to high-period (12~h and 24~h) rhythms.
-In addition, a large amount of signal is detected for low period (around 60~s) events -- 
+In addition, a large amount of signal is detected for low-period (around 60~s) events -- 
 likely corresponding to the position of animals walking along (back and forward) their tubes in a very paced manner.

### 2018-04-17 results.Rnw
another few touches

@@
 
-The study of the rhythmical activity of fruit flies has played a crucial role in the development of circadian biology\citationneeded{}.
+The study of the rhythmical activity of fruit flies has played a crucial role in the development of circadian biology\cite{dubowy_circadian_2017}.
 To date, most of the behavioural data used in the field is acquired through the Drosophila Activity Monitor System (DAMS).
@@
 <<cache=TRUE, warning=FALSE, result="hide", message=FALSE>>=
-# we also show a subset of this figure in 4A
+# we also show a subset of this figure in Fig 4A
 figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
-            # bars n the z axis could 
             # one could also use stat_tile_etho
@@
 dt_ll <- dt[t > days(1)]
-# compute chi sqr periodogram 
+# compute chi square periodograms 
 per_dt <- periodogram(moving, 
@@
 <<cache=TRUE, warning=FALSE, results="hide">>=
-summary_dt <- 
-    per_dt[, 
+summary_dt <- per_dt[, 
                 .(
@@
               # shows all individuals as points
-              # the size of the point expresses the power of the peak
+              # the size of the point represents the power of the peak
               geom_jitter(aes(colour = genotype, 
@@
                               alpha = 0.67) +
-              # We would like to convert time in hour
+              # We would like to show time in hours
               scale_y_hours("Period")
@@
 # to round all numeric columns  to two digits
-result_dt[, lapply(.SD, function(x) 
-				if(is.numeric(x)) round(x, 2) else x
-			)]
+result_dt[, lapply(.SD,
+                   function(x) if(is.numeric(x)) round(x, 2) else x
+                   )
+           ]
 @

### 2018-04-17 results.Rnw

--- 2018-04-13 results.Rnw
@@
 
-In order to illustrate the usefulness of \texttt{rethomics}, we provide two carefully annotated examples.
+In order to illustrate the usefulness of \texttt{rethomics}, we provide two use cases examples.
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
@@
 \subsection*{Drosophila Activity Monitor}
-The \texttt{zeitgebr} package implements a comprehensive suite of methods to analyse circadian rhythms, 
-including the computation of autocorrelograms, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms,
-and peak detection.
+It uses the package \texttt{zeitgebr} which implements a suite of methods to compute autocorrelogram,
+ $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms, and find their peaks.
 
-The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
+The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded {This is not atemporal, in 10 years will still say recently} with the Nobel Prize in Physiology or Medicine.
 Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
@@
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
-In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in LL,
 and that the endogenous period in LL is longer than 24 hours.
 
-Here, we guide the reader through the analysis of two of the genotypes employed in that study;
-one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in clock neurons (TIM/NKCC\textsuperscript{ox}).
-In particular, we outline the necessary steps to analyse two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
-The \texttt{metadata.csv} file as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
-
-We start by downloading the data and extract the zip archive in our working directory.
-Then we load the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
+We took two genotypes employed in the study, one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in all clock neurons (TIM/NKCC\textsuperscript{ox}).
+In particular, we re-analysed two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
+The \texttt{metadata.csv} file, which describes exhaustively all individuals, as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+As a prerequisite, we downloaded the data and extracted the zip archive in our working directory.
+We start the analysis by loading the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
 
@@
 \subsubsection*{Preprocessing}
-Since the two original replicates do not have the same baseline duration and we want to analyse them together,
-we align their respective times to the experimental perturbation: the transition from LD to LL ($t = 0$). 
-This is achieved by subtracting the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
+The two replicates do not have the same baseline time and we would like to express the time relative to the important event:
+the transition from LD to LL. 
+Therefore, we subtract the \texttt{baseline\_days} metavariable from the \texttt{t} variable.
 This gives us an opportunity to illustrate the use \texttt{xmv()}, which expands metavariables as variables.
 In addition, we use the \texttt{data.table} syntax to create, in place, a \texttt{moving} variable.
-It is \texttt{TRUE} when \texttt{activity} is greater than zero and \texttt{FALSE} otherwise:
+It is \texttt{TRUE} when and only when \texttt{activity} is greater than zero:
 
@@
 @
-The \texttt{id} is a long and exhaustive string of character, which incidentally 
-makes it difficult to read and display as a label on a plot.
-To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+
+To simplify visualisation, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
-Importantly, we also retain \texttt{id} as an \emph{unambiguous} unique identifier.
+Importantly, we also keep \texttt{id} as an \emph{unambiguous} unique identifier.
 Indeed, two animals in separate experiments may have the same label, but different \texttt{id}s.
@@
 \subsubsection*{Curation}
-It is important to visualise an overview of how each individual behaved and, if necessary, amend the metadata accordingly.
-For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
+It is important to visualise an overview of how each individual behaved and, if necessary, alter the data accordingly. For this, we generate a tile plot (Fig~\ref{fig:fig-3}A).
 
@@
 
-The activity of dead or escaped animals is falsely scored as long series of zeros,
-which may be erroneously interpreted as inactivity (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
+The activity of dead or escaped animals is falsely scored as long series of zeros (see, for instance, individual 30 and 18 in Fig~\ref{fig:fig-3}A).
 The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
@@
 
-In addition, we can trim the data to the same number of days across experiments and individuals.
+In addition, we can trim our data to have the same number of days across experiments and individuals.
 <<cache=TRUE>>=
@@
 
-For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have lived for \emph{at least five days in LL}.
+For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have \emph{at least five days in LL}.
 An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.
@@
 \subsubsection*{Double-plotted actograms}
-`Double-plotted actograms' are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
+`Double-plotted actograms' are a common visualisation of periodicity and rhythmicity in circadian experiments.
 In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
@@
 	\caption{{\bf Visualisation of the periodicity in activity of eight selected animals.}
-		A: Double-plotted actograms showing all activity during experiment.
-		Time is defined relative to the transition from LD to LL (at day 0).
+		A: Double-plotted actograms showing activity over the experiment. Transition from LD to LL happens at day 0.
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
@@
 
-As shown in the original study\cite{buhl_quasimodo_2016}, double-plotted actograms and periodograms suggest that NKCC\textsuperscript{ox}/+ flies are mostly 
-arhythmic in LL whilst Tim/NKCC\textsuperscript{ox} appear to have a long-period rhythm.
-To visualise this difference at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
+As shown in the original study, double-plotted actograms and periodograms suggest that NKCC\textsuperscript{ox}/+ flies are mostly 
+arhythmic in LL whilst Tim/NKCC\textsuperscript{ox} appear to have a consistent, long-period rhythm.
+To visualise this at the population scale, we can plot an average periodogram (see Fig~\ref{fig:fig-5}A):
 
@@
 
-To further quantify this difference, we opt to show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each group (see Fig~\ref{fig:fig-5}B).
+To further quantify this difference, we can show the number of rhythmic animals -- \emph{i.e.} individuals for which a peak was found -- in each group (see Fig~\ref{fig:fig-5}B).
 Then, we can compare the average value of the peak for the rhythmic animals (see Fig~\ref{fig:fig-5}C).
@@
 \texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics.
-It can therefore be used directly by \texttt{ggplot}:
+It can therefore be used directly by \texttt{ggplot} and other tools:
 
@@
 
-\texttt{R} provides one of the richest statistical toolboxes available.
-Using base \texttt{R} we could perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes, or,
-like in this case, fit a binomial generalised linear model:
+\texttt{R} provides one of the richest statistical toolboxes available, which allows users to go deeper in the analysis of the extracted variables.
+One could, for instance, perform a $\chi{}^2$ test on the number of rhythmic \emph{vs} arhythmic flies in both genotypes.
+To address the same question, we fit a binomial generalised linear model:
 
@@
 fit <- glm(is_rhythmic ~ genotype, summary_dt, family = "binomial")
+
 summary(fit)$coefficients
@@
 <<warning=FALSE>>=
-result_dt <- 
-  summary_dt[,
+result_dt <- summary_dt[,
           .(
-       mean_period = mean(first_peak_period, na.rm = T) / hours(1),
-       sd_period = sd(first_peak_period, na.rm = T) / hours(1),
-       percent_rhythmic = 100 * sum(is_rhythmic) / .N,
+            mean_period = round(mean(first_peak_period, na.rm = T) / hours(1), 2),
+            sd_period = round(sd(first_peak_period, na.rm = T) / hours(1), 2),
+            percent_rhythmic = round(100 * sum(is_rhythmic) / .N, 2),
             n_rhythmic = sum(is_rhythmic),
@@
           ]
-
-# to round all numeric columns  to two digits
-result_dt[, lapply(.SD, function(x) 
-				if(is.numeric(x)) round(x, 2) else x
-			)]
+result_dt
 @
 
-The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state-of-the-art statistics.
-We were able to comprehensively analyse the data from a circadian experiment with a few lines of code,
-presenting a workflow that applies equally well to much larger datasets.
+This example shows how \texttt{rethomics} can be used from the raw data to making publication-quality figures and statistics.
+We were able to comprehensively analyse the data from a circadian experiment with a few lines of code.
+This workflow applies particularly to much larger data sets and provides a large degree of flexibility that will allow biologists to tune their analysis to their specific questions.
 
@@
 We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
-Our variable of interest is the position of animals in their tube (from the food end, $Position = 0$,  to the cotton end, $Position = 1$).
+Our variable of interest is the position of animals in their tube (from the food end, 0,  to the cotton end, 1).
 Fig~\ref{fig:fig-6}A-C shows the raw position data at two different scales for two representative animals.
@@
 
-As suggested by the slow oscillations of the mean position (Fig~\ref{fig:fig-6}A), we observe peaks in power corresponding to high-period (12~h and 24~h) rhythms.
-In addition, a large amount of signal is detected for low period (around 60~s) events -- 
-likely corresponding to the position of animals walking along (back and forward) their tubes in a very paced manner.
+As suggested by the slow oscilations of the mean position (Fig~\ref{fig:fig-6}A), we observe strong, high period (12~h and 24~h), rhythms.
+In addition, a large amount of signal is detected for low period (around 60~s) events.
+This corresponds to the position of animals walking along (back and forward) their tubes in a very paced manner.
 

### 2018-04-17 manuscript.Rnw

@@
 \texttt{rethomics} offers a computational solution to storing, manipulating and visualising large amounts of behavioural data.
-We propose it as a tool to breach the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+We propose it as a tool to bridge the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).
@@
 The biological significance and determinism of animal behaviours have long been a subject of prime scientific interest.
-In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant,
-which breached the gap between genetics and the study of behaviour\citationneeded{}.
+In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant, which connected genetics and behaviour\cite{sokolowski_drosophila_2001}.
 In the last few decades, our ability to acquire and analyse vast amount of biological data has tremendously increased\cite{stephens_big_2015},
-which is deeply transforming both genetics\citationneeded{} and neurobiology\citationneeded{}.
+which has been deeply transforming both genetics\cite{schatz_biological_2015} and neuroscience\cite{sejnowski_putting_2014}.
 %The rise of high-throughput methods and computational approaches, have particularly transformed genetics and neurobiology.
 In fact, ethology itself is undergoing its own transition towards data sciences, which has prompted the terms of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
-It is now accepted that the study of behaviour can also benefit from quantitative sciences such as mathematics, physics and computational linguistics \cite{brown_ethology_2018, berman_measuring_2018}.
+It is now accepted that the study of behaviour can also benefit from quantitative sciences such as machine learning, physics and computational linguistics \cite{brown_ethology_2018, berman_measuring_2018}.
 
@@
 to continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
-to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression \citationneeded{}, and courtship\cite{tsai_image_2012};
+to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014} and courtship\cite{tsai_image_2012};
 and to study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
@@
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
-Since many aspect of behaviour analysis are also becoming increasingly linked to data sciences, the development of such common tools and data structures would be very valuable.
+Since many aspect of behaviour analysis are also becoming increasingly dependent on data sciences, the development of such common tools and data structures would be very valuable.
 
@@
 In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
-At the core of \texttt{rethomics} lies the \texttt{behavr} data structure, an object used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
+At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.

### 2018-04-17 results.Rnw

--- 2018-04-17 results.Rnw
@@
 
-%The study of circadian rhythm, employing fruit flies as an animal model, is a well established research field recently awarded with the Nobel Prize in Physiology or Medicine.
-%Importantly, DAM2 (the second, and most popular, version of the Drosophila Activity Monitors System) is the most widely adopted behaviour recording platform in this research field.
-%We gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
-%Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
-%In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
-%and that the endogenous period in LL is longer than 24 hours.
-
 The study of the rhythmical activity of fruit flies has played a crucial role in the development of circadian biology\cite{dubowy_circadian_2017}.
@@
 we gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+
 Wild type flies are highly rhythmic in Light-Dark (LD) cycles and become arrhythmic in constant light (LL).
-In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in both LD and LL,
+In their study, the authors gain understanding of the function of the molecular clock by showing that overexpression of the gene NKCC makes the flies rhythmic in LL,
 and that the endogenous period in LL is longer than 24 hours.

### 2018-04-17 manuscript.tex
all files merged for submission

@@
 \texttt{rethomics} offers a computational solution to storing, manipulating and visualising large amounts of behavioural data.
-We propose it as a tool to bridged the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
+We propose it as a tool to bridge the gap between behavioural biology and data sciences, thus connecting computational and behavioural scientists.
 \texttt{rethomics} comes with a extensive documentation as well as a set of both practical and theoretical tutorials (available at \href{https://rethomics.github.io}{https://rethomics.github.io}).

### 2018-04-18 manuscript.Rnw
typose + esteban last fixes

@@
 % TODO revert for sumbmisson
-\linenumbers
+%\linenumbers
 \TODO{revert line numbers before submission}
@@
 In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant, which connected genetics and behaviour\cite{sokolowski_drosophila_2001}.
-In the last few decades, our ability to acquire and analyse vast amount of biological data has tremendously increased\cite{stephens_big_2015},
+In the last few decades, our ability to acquire and analyse vast amounts of biological data has tremendously increased\cite{stephens_big_2015},
 which has been deeply transforming both genetics\cite{schatz_biological_2015} and neuroscience\cite{sejnowski_putting_2014}.
@@
 to continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
-to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014} and courtship\cite{tsai_image_2012};
+to score more complex ones such as feeding\cite{itskov_automated_2014,ro_flic_2014}, aggression\cite{dankert_automated_2009} and courtship\cite{tsai_image_2012};
 and to study the behaviour of multiple interacting animals \cite{swierczek_high-throughput_2011,perez-escudero_idtracker_2014,robie_mapping_2017}.
@@
 In addition, they can be used on remote resources such as computer clusters, which makes them more scalable to the context of `big data'\cite{hashem_rise_2015}.
-Since many aspect of behaviour analysis are also becoming increasingly dependent on data sciences, the development of such common tools and data structures would be very valuable.
+Since many aspects of behaviour analysis are also becoming increasingly dependent on data sciences, the development of such common tools and data structures would be very valuable.
 
-At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and time scale -- for a similar community to arise.
+At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and timescale -- for a similar community to arise.
 However, some low-level conceptual consistencies and methodological challenges are common across experiments.
@@
 In this model, each task of the analysis workflow (\emph{i.e.} data import, manipulation and visualisation) is handled by a different package, and new ones can be designed to suit specific needs.
-At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
+At the core of \texttt{rethomics} lies the \texttt{behavr} table, a structure used to store large amounts of data (\emph{e.g.} position and activity) and metadata (\emph{e.g.} treatment and genotype) in a unique \texttt{data.table}-derived object\cite{dowle_data.table_2017}.
 Any input package will import experimental data as a \texttt{behavr} table which can, in turn, be analysed and visualised regardless of the original input platform.
-Numerical results and plots are standard objects that can therefore be further analysed inside the wide \texttt{R} package ecosystem.
+Numerical results and plots are standard objects that can, therefore, be further analysed inside the wide \texttt{R} package ecosystem.
 
@@
 		In the data, each row is a `read' (\emph{i.e.} information about one individual at one time-point).
-		It is formed of $q$ variables and is expected to have a very large number of reads, $k$, for each individual $i$.
+		It is formed of $q$ variables and is expected to have a very large number of reads, $k$, for each individual $i\in [1,n]$.
 		Data and metadata are implicitly joined on the \texttt{id} field.
 		Note that the names used for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
-		B: Non exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
+		B: Non-exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
 		In addition to operations on data, which are inherited from \texttt{data.table},
@@
 Importantly, \texttt{ggetho} is fully compatible with \texttt{ggplot2}. 
-For instance, \texttt{ggplot2} operations such as faceting, transforming axes and adding new layers will function natively with \texttt{ggetho}.
+For instance, \texttt{ggplot2} operations such as faceting, transforming axes and adding new layers functions natively with \texttt{ggetho}.
 
@@
 \href{https://github.com/rethomics}{https://github.com/rethomics/}.
-Extensive installation instructions as well as reproducible demos and tutorials are available at
+Extensive installation instructions, as well as reproducible demos and tutorials, are available at
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
@@
 Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
-Marcus Gosh and Tara Kane for their feed back on the manuscript;
+Marcus Gosh and Tara Kane for their feedback on the manuscript;
 Brenna Williams, for her help to support multi-beam DAMS;

### 2018-04-18 manuscript.tex
typose + esteban last fixes

@@
 
-In order to illustrate the usefulness of \texttt{rethomics}, we provide two carefully annotated examples.
+In order to illustrate the usefulness of \texttt{rethomics}, we provide two examples.
 The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
-The second one shows how \texttt{rethomics} integrates with the rest of \texttt{R} to perform a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
+The second one shows how \texttt{rethomics} integrates within \texttt{R} to perform a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
 
@@
 including the computation of autocorrelograms, $\chi{}^2$\cite{sokolove_chi_1978} and Lomb-Scargle\cite{ruf_lomb-scargle_1999} periodograms,
-and peak detection.
+and automatic peak detection.
 
@@
 Here, we guide the reader through the analysis of two of the genotypes employed in that study;
-one control group (NKCC\textsuperscript{ox}/+) and one where NKCC\textsuperscript{ox} is overexpressed in clock neurons (TIM/NKCC\textsuperscript{ox}).
+one control group (NKCC\textsuperscript{ox}/+, which is arrhythmic under LL) and one where NKCC\textsuperscript{ox} is overexpressed in clock neurons (TIM/NKCC\textsuperscript{ox}, which is rhythmic under LL).
 In particular, we outline the necessary steps to analyse two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
-The \texttt{metadata.csv} file as well as all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+The \texttt{metadata.csv} file and all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
 
@@
 
-The activity of dead or escaped animals is falsely scored as long series of zeros,
+The activity of dead or escaped animals is scored as long series of zeros,
 which may be erroneously interpreted as inactivity (see, for instance, individual labelled \texttt{30} and \texttt{18} in Fig~\ref{fig:fig-3}A).
@@
 %	\includegraphics[width=1\textwidth]{fig/fig-4.pdf}
-	\caption{{\bf Visualisation of the periodicity in activity of eight selected animals.}
+	\caption{{\bf Visualisation of the periodicity in the activity of eight selected animals.}
 		A: Double-plotted actograms showing all activity during experiment.
@@
 			      Values of the peak period for animals with a significant peak (\emph{i.e.} rhythmic).
-			      Individual animals are shown by dots whose size represent relative power of the peak period.
+			      Individual animals are shown by dots whose size represent the relative power of the peak period.
 			      The error bars are 95\% bootstrap confidence interval on the population mean.
@@
 
-One of the challenges of behaviour analysis is the `nesting' of events happening over different time scales.
+One of the challenges of behaviour analysis is the `nesting' of events happening over different timescales.
 In other words, a behavioural variable can be modulated by the circadian rhythm, but also by co-occurring ultradian and infradian rhythms.
 For instance, an animal could have rhythmic bursts of locomotor activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable.
-In this example, both rhythms happen at time scales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate in the same analysis.
+In this example, both rhythms happen at timescales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate into the same analysis.
 Being able to keep frequency information over multiple scales is however important in some cases.
 In particular, when interested in the frequency modulation of a rhythm by another --
-that is, if the periodicity of a high frequency rhythm itself can be function of a lower frequency one.
+that is, if the periodicity of a high-frequency rhythm itself can be a function of a lower frequency one.
 %For instance, the frequency of rhythmic activity bursts could itself be regulated by another rhythm such as the circadian rhythm.
 
-The problem of understanding time series at different scales in not uncommon in fields such as economics\cite{aguiar-conraria_business_2011}, climate sciences\cite{lau_climate_1995} and ecology\cite{cazelles_wavelet_2008} where variables are governed by multiple underlying rhythms (\emph{e.g.} tidal, daily, yearly and multi-yearly).
+The problem of understanding time series at different scales is not uncommon in fields such as economics\cite{aguiar-conraria_business_2011}, climate sciences\cite{lau_climate_1995} and ecology\cite{cazelles_wavelet_2008} where variables are governed by multiple underlying rhythms (\emph{e.g.} tidal, daily, yearly and multi-yearly).
 One approach is to study a variable of interest in the time/period domain using, for instance, continuous wavelet transform (CWT)\cite{grossmann_decomposition_1984}.
@@
 		The thick coloured lines show the average position every two hours.
-		The green rectangles in the background shows the two time windows selected for B and C.
-		B: Close up of A, showing position over one hour, in the beginning of the L phase of day 1.
-		C: Close up of A, showing position over one hour, in the middle of the L phase of day 1.
+		The green rectangles in the background show the two time windows selected for B and C.
+		B: Close up of A, showing position over one hour, starting at the beginning of the L phase of day 1.
+		C: Close up of A, showing position over one hour, starting at the middle of the L phase of day 1.
 		D: Continuous wavelet transform spectrogram for the two representative animals.
 		E: Average spectrogram across 40 males and 40 females.
-		In D and E, the lines on the right shows the marginal power spectra corresponding to the shown spectrograms (average across all time).
+		In D and E, the lines on the right show the marginal power spectra corresponding to the shown spectrograms (average across all time).
 		The male data was collected and described in our previous study \cite{geissmann_ethoscopes_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
@@
 Interestingly, in females, this low period pace appears to be frequency modulated during the L phase, suggesting a slower walking speed around ZT6~h.
-In contrast, males shows only a high frequency rhythm around the phase transitions (L to D and D to L).
-Surprisingly, the peak of high frequency rhythm imply a faster pace in males (approximately 60~s) than in females (approximately 120~s) --
+In contrast, males display only a high-frequency rhythm around the phase transitions (L to D and D to L).
+Surprisingly, the peak of high-frequency rhythm implies a faster pace in males (approximately 60~s) than in females (approximately 120~s) --
 indicating that, when active, males walk faster than females.

### 2018-04-20 manuscript.Rnw
few fixes after another round of comments

@@
 
-The biological significance and determinism of animal behaviours have long been a subject of prime scientific interest.
-In the 1960s, behavioural geneticists, such as Seymour Benzer showed that some apparently complex behaviours could be in fact governed by simple genetic determinant, which connected genetics and behaviour\cite{sokolowski_drosophila_2001}.
+The biological implications and the determinism of animal behaviour have long been a subject of significant scientific interest.
+In the 1960s, behavioural geneticists, such as Seymour Benzer, showed that some apparently complex behaviours could in fact be governed by simple genetic determinants, which connected genetics and behaviour\cite{sokolowski_drosophila_2001}.
 In the last few decades, our ability to acquire and analyse vast amounts of biological data has tremendously increased\cite{stephens_big_2015},
-which has been deeply transforming both genetics\cite{schatz_biological_2015} and neuroscience\cite{sejnowski_putting_2014}.
+deeply transforming both genetics\cite{schatz_biological_2015} and neuroscience\cite{sejnowski_putting_2014}.
 %The rise of high-throughput methods and computational approaches, have particularly transformed genetics and neurobiology.
-In fact, ethology itself is undergoing its own transition towards data sciences, which has prompted the terms of `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
+In fact, ethology itself is undergoing its own transition towards data sciences, which has prompted the terms `ethomics' \cite{branson_high-throughput_2009, reiser_ethomics_2009} and `computational ethology'\cite{anderson_toward_2014}.
 It is now accepted that the study of behaviour can also benefit from quantitative sciences such as machine learning, physics and computational linguistics \cite{brown_ethology_2018, berman_measuring_2018}.
@@
 the multiplicity of model organisms, hypotheses and paradigms has led to the existence of a very diverse palette of specific recording techniques.
-Some tools were developed, for instance:
+Some tools were developed, for instance,
 to continuously record simple behavioural features such as walking activity\cite{faville_how_2015} and  position\cite{pelkowski_novel_2011} over long durations (days or weeks);
@@
 At first, it may seem as though behavioural experiments are prohibitively heterogeneous -- in terms of model organisms, paradigm and timescale -- for a similar community to arise.
-However, some low-level conceptual consistencies and methodological challenges are common across experiments.
+However, low-level conceptual consistencies and methodological challenges are shared across experiments.
 For instance, the results (\emph{i.e.} the `data')  feature a set of long time series (sometimes multivariate and irregular), but also contain a formal description of the treatment applied to each individual, the `metadata'.
@@
 can be shared between acquisition platforms and experimental paradigms.
-% it broaden the scope for developing of utilities an shared principles.
 
@@
 \section*{Design and Implementation}
-\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}, which is widely taught and adopted by computational biologists,
+\texttt{rethomics} is implemented in \texttt{R}\cite{r_core_team_r_2017}
 as a collection of packages (Fig~\ref{fig:fig-1}).
@@
 
-Ethomics results can easily scale and data structure therefore gains from being computationally efficient -- both in term of memory footprint and processing speed.
+Ethomics results can easily scale and data structure therefore gains from being computationally efficient -- both in terms of memory footprint and processing speed.
 For instance, there could be very long time series, sampled several times per second, over multiple days, for each individual. %(typically, $k_i > 10^8, \forall i \in [1,n]$) 
@@
 For instance, metadata stores information regarding the date and location of the experiment, treatment, genotype, sex, \emph{post hoc} observations and other arbitrary metavariables.
-A large set of metavariables is an important asset since they can later be used as covariates.
+A large set of metavariables is an important asset since they can later be treated as covariates.
 
-\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2}A~and~B).
+\texttt{behavr} tables link metadata and data within the same object, extending the syntax of \texttt{data.table} to manipulate, join and access metadata (Fig~\ref{fig:fig-2} and \nameref{S1-Fig}).
 This approach guarantees that any data point can be mapped correctly to its parent metadata thanks to a shared key (\texttt{id}).
 Furthermore, it allows implicit update of metadata when data is altered.
-For instance, when data is filtered, only the remaining individuals should be in the new metadata. 
+For instance, when data is subsetted, only the remaining individuals should be in the new metadata. 
 It is also important that metadata and data can interoperate --
-for example, when updating a variable according to the value of a metavariable (say, alter the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
+for example, when updating a variable according to the value of a metavariable (say, altering the variable \texttt{x} only for animals with the metavariable \texttt{sex = `male'}).
 The online tutorials and documentation provide a detailed set of examples and concrete use cases of \texttt{behavr}.
@@
 	\caption{{\bf \texttt{behavr} table.}
-	A: Illustration of a \texttt{behavr} object, the core data structure in \texttt{rethomics}.
+	Illustration of a \texttt{behavr} object, the core data structure in \texttt{rethomics}.
 		The metadata holds a single row for each of the $n$ individuals. 
@@
 		Note that the names used for variables and metavariable in this example are only plausible cases which will likely differ in practice. 
-		B: Non-exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}). 
-		In addition to operations on data, which are inherited from \texttt{data.table},
-		we provide utilities designed specifically to act on both metadata and data.  
-		Commented examples are prefixed by `\texttt{>}'.
 	}
@@
 \subsection*{Data import}
-Data import packages translate results from a specific recording platform (\emph{e.g.} text files and databases) into a single \texttt{behavr} object.
-Currently, we provide two packages: one to import results from single and multi-beam Drosophila Activity Monitor Systems (Trikinetics Inc.) 
+Data import packages translate results from a specific recording platform (\emph{e.g.} text files and relational databases) into a single \texttt{behavr} object.
+Currently, we provide two packages: one to import data from single and multi-beam Drosophila Activity Monitor Systems (Trikinetics Inc.) 
 and another one for Ethoscopes \cite{geissmann_ethoscopes_2017}.
@@
 In all cases, users must provide a metadata table, with one row per individual,
-and featuring both mandatory and optional columns (Fig~\ref{fig:fig-2}A).
+and featuring both mandatory and optional columns (Fig~\ref{fig:fig-2}).
 The mandatory ones are the necessary and sufficient information to fetch data (\emph{e.g.} machine id, region of interest and date). 
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+
 \paragraph*{S1 Fig.}
 \label{S1-Fig}
+{\bf Non-exhaustive list of uses of a \texttt{behavr} table (referred as \texttt{dt}).}
+In addition to operations on data, which are inherited from \texttt{data.table},
+we provide utilities designed specifically to act on both metadata and data.  
+Commented examples are prefixed by `\texttt{>}'.
+
+\paragraph*{S2 Fig.}
+\label{S2-Fig}
 {\bf Complete version of Fig~\ref{fig:fig-4}.}
@@
 In particular, Han Kim, for his invaluable comments on the early versions of \texttt{rethomics} and his dedicated contribution to the tutorials;
-Maite Ogueta, for making the DAMS results data available;
+Maite Ogueta and Ralf Stanewsky, for making the DAMS results data available;
 Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;

### 2018-04-20 manuscript.tex
few fixes after another round of comments

@@
 In particular, we outline the necessary steps to analyse two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
-The \texttt{metadata.csv} file and all the associated result files can be downloaded at \href{https://zenodo.org/record/1172980}{https://zenodo.org/record/1172980}.
+The \texttt{metadata.csv} file and all the associated result files can be downloaded on zenodo\cite{ogueta_ll_2018}.
 
-We start by downloading the data and extract the zip archive in our working directory.
-Then, we load the necessary \texttt{rethomics} packages (see the webpage for installation instructions):
+Fist, we install the \texttt{rethomics} packages (see the webpage for installation instructions) and download the data to extract the zip archive in our working directory.
+Then, we load the necessary \texttt{rethomics} packages:
 
@@
 \end{knitrout}
-The \texttt{id} is a long and exhaustive string of character, which incidentally 
-makes it difficult to read and display as a label on a plot.
-To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+
+The \texttt{id} is a long string of character (for instance, `\texttt{2013-11-19~09:00:00|Monitor36.txt|01}'), 
+which makes it difficult to read and display as a label on a plot.
+To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype} (\emph{e.g.} `\texttt{1.NKCCOX/+}').
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
@@
 			The red `\textbf{+}' symbols show animals that were removed from the subsequent analysis as they did not survive five complete days in LL.}
+			The white and black rectangles on the time axis show L and D phases, respectively.
+			In the LL regime (for $t > 0$), grey rectangles represent subjective nights.
+
 	\label{fig:fig-3}
@@
 which may be erroneously interpreted as inactivity (see, for instance, individual labelled \texttt{30} and \texttt{18} in Fig~\ref{fig:fig-3}A).
-The \texttt{sleepr} package offers a tool to detect and remove such artefactual data.
+The \texttt{sleepr} package offers a tool to detect and remove such data.
 It proceeds by detecting the first time an animal is immobile for more than 99~\% of the time (the default) for at least \texttt{time\_window} seconds and then discard any subsequent data.
@@
 
-For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have lived for \emph{at least five days in LL}.
+For the purpose of this example, we also exclude animals that died prematurely, and keep only individuals that have lived for at least five days in LL.
 An overview of the curate data can be visualised in Fig~\ref{fig:fig-3}B.
@@
 
-Note that as a result, we now have 53 `valid' individuals.
+Note that, as a consequence of the curation procedure, we now have 53 `valid' individuals.
 
 \subsubsection*{Double-plotted actograms}
-`Double-plotted actograms' are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
+Double-plotted actograms are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
+In  \nameref{S2-Fig}A, we show a double-plotted actogram for each animal.
 A selected sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
@@
 \hlcom{# we also show a subset of this figure in Fig 4A}
-\hlstd{figS1A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
+\hlstd{figS2A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
             \hlcom{# one could also use stat_tile_etho}
@@
 Ultimately, in order to quantify periodicity and rhythmicity, we compute periodograms.
-Several methods are implemented in \texttt{zeitbebr}: $\chi{}^2$, Lomb-Scargle, autocorrelation 
+Several methods are implemented in \texttt{zeitgebr}: $\chi{}^2$, Lomb-Scargle, autocorrelation 
 and Fourier.
 In this example, we generate $\chi{}^2$ periodograms and lay them out in a grid.
-Periodograms for the subset of eight animals used in Fig~\ref{fig:fig-4}A is shown in Fig~\ref{fig:fig-4}B. 
-See \nameref{S1-Fig}B for the visualisation of all individuals.
+Periodograms for the subset of eight animals used in Fig~\ref{fig:fig-4}A are shown in Fig~\ref{fig:fig-4}B. 
+See \nameref{S2-Fig}B for the visualisation of all individuals.
 
@@
 \hlcom{# we also show a subset of this figure in supplementary materials}
-\hlstd{figS1B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak} \hlstd{= peak))} \hlopt{+}
+\hlstd{figS2B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak} \hlstd{= peak))} \hlopt{+}
                   \hlcom{# periododogram drawn as a line}
@@
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
-		The blue cross represents the highest peak (if present) above the significance threshold (red line).
+		The blue cross represents the highest peak (if present) above the significance threshold, at $\alpha = 0.05$, (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
-		See \nameref{S1-Fig} for all 53 animals.
+		See \nameref{S2-Fig} for all 53 animals.
 	}
@@
       C: Peak periodicity power and average.
-			      Values of the peak period for animals with a significant peak (\emph{i.e.} rhythmic).
+			      Values of the peak period for animals with a peak above the significance threshold, at $\alpha = 0.05$(\emph{i.e.} rhythmic).
 			      Individual animals are shown by dots whose size represent the relative power of the peak period.
@@
 
-\texttt{summary\_dt} is just a regular data frame with one row per individual, containing both metadata and our summary statistics. 
+\texttt{summary\_dt} is a regular data frame with one row per individual, containing both metadata and our summary statistics. 
 It can therefore be used directly by \texttt{ggplot}:
@@
 
-The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state-of-the-art statistics.
+The case study described so far shows how \texttt{rethomics} can be employed to generate publication-quality figures and state-of-the-art statistics on pre-existing data.
 We were able to comprehensively analyse the data from a circadian experiment with a few lines of code,
@@
 In other words, a behavioural variable can be modulated by the circadian rhythm, but also by co-occurring ultradian and infradian rhythms.
-For instance, an animal could have rhythmic bursts of locomotor activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian  (\emph{i.e.} 24~h) regulation of the same variable.
-In this example, both rhythms happen at timescales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate into the same analysis.
+For instance, an animal could have rhythmic bursts of locomotor activity recurring at high frequency (\emph{e.g.} 1~min), but also a circadian (\emph{i.e.} 24~h) regulation of the same variable,
+both rhythms would then happen at timescales separated by approximately three orders of magnitudes, which makes them difficult to visualise and integrate into the same analysis.
 Being able to keep frequency information over multiple scales is however important in some cases.
@@
 	\caption{{\bf Wavelet analysis of positional data.}
-		A: Raw position data for a representative female (top) and male (bottom) \emph{Drosophila} over five days, in black. 
+		A: Raw position data for a representative female (top) and male (bottom) \emph{Drosophila~melanogaster} over five days, in black. 
 		The thick coloured lines show the average position every two hours.
 		The green rectangles in the background show the two time windows selected for B and C.
-		B: Close up of A, showing position over one hour, starting at the beginning of the L phase of day 1.
-		C: Close up of A, showing position over one hour, starting at the middle of the L phase of day 1.
+		B: Close up of the first time window in A, showing position over one hour, starting at the beginning of the L phase of day 1.
+		C: Close up of the second time window in A, showing position over one hour, starting at the middle of the L phase of day 1.
 		D: Continuous wavelet transform spectrogram for the two representative animals.

### 2018-04-20 results.Rnw
few fixes after another round of comments

@@
 @
-The \texttt{id} is a long and exhaustive string of character, which incidentally 
-makes it difficult to read and display as a label on a plot.
-To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype}.
+
+The \texttt{id} is a long string of character (for instance, `\texttt{2013-11-19~09:00:00|Monitor36.txt|01}'), 
+which makes it difficult to read and display as a label on a plot.
+To address this issue, we create our own \texttt{label} metavariable, as the combination of a number and \texttt{genotype} (\emph{e.g.} `\texttt{1.NKCCOX/+}').
 In the restricted context of this analysis, \texttt{label} acts as a unique identifier.
@@
 valid_ids <- valid_dt[valid == T, id]
-fig3B <- fig3B + geom_text(data = dt[!id %in% valid_ids,.(label), meta = T], aes(x = days(6), y = label), inherit.aes = F, colour = "red", label = "+", fontface = "bold")
+fig3B <- fig3B + geom_text(data = dt[!id %in% valid_ids,.(label), meta = T], aes(x = days(6.1), y = label), inherit.aes = F, colour = "red", label = "+", fontface = "bold", size=10, vjust=0.35)
 @
@@
 
-Note that as a result, we now have \Sexpr{nrow(dt[meta=T])} `valid' individuals.
+Note that, as a consequence of the curation procedure, we now have \Sexpr{nrow(dt[meta=T])} `valid' individuals.
 
 \subsubsection*{Double-plotted actograms}
-`Double-plotted actograms' are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
-In  \nameref{S1-Fig}A, we show a double-plotted actogram for each animal.
+Double-plotted actograms are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
+In  \nameref{S2-Fig}A, we show a double-plotted actogram for each animal.
 A selected sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
@@
 # we also show a subset of this figure in Fig 4A
-figS1A <- ggetho(dt, aes(z = moving), multiplot = 2) +
+figS2A <- ggetho(dt, aes(z = moving), multiplot = 2) +
             # one could also use stat_tile_etho
@@
 # we also show a subset of this figure in supplementary materials
-figS1B <- ggperio(per_dt, aes(y = power, peak = peak)) + 
+figS2B <- ggperio(per_dt, aes(y = power, peak = peak)) + 
                   # periododogram drawn as a line
@@
 		B: $\chi{}^2$ periodograms over the LL part of the experiment matching the animals in A.
-		The blue cross represents the highest peak (if present) above the significance threshold (red line).
+		The blue cross represents the highest peak (if present) above the significance threshold, at $\alpha = 0.05$, (red line).
 		Titles on top of each facet refer to the label allocated to each individual.
-		See \nameref{S1-Fig} for all \Sexpr{nrow(dt[meta=TRUE])} animals.
+		See \nameref{S2-Fig} for all \Sexpr{nrow(dt[meta=TRUE])} animals.
 	}
@@
 ################################ CODE BELOW FOR COSMETIC ALTERATIONS OF FIGURES
+
+MUTANT_PALETTES <- c("#babdb6", "#729fcf")
+
+mutant_colours <- list(scale_colour_manual(values = MUTANT_PALETTES),
+                      scale_fill_manual(values = MUTANT_PALETTES))
+
 library(cowplot)
@@
 #=================================================
-figS1A <- figS1A +  theme(strip.text.x = element_text(size =8))
-figS1B <- figS1B +  theme(strip.text.x = element_text(size =8))
+figS2A <- figS2A +  theme(strip.text.x = element_text(size =8))
+figS2B <- figS2B +  theme(strip.text.x = element_text(size =8))
 
-figS1 <- cowplot::plot_grid(
-                          figS1A,
-                          figS1B,
+figS2 <- cowplot::plot_grid(
+                          figS2A,
+                          figS2B,
                           nrow=1,
@@
 
-pdf("../fig/fig-S1.pdf",w=12,h=16)
-print(figS1)
+pdf("../fig/fig-S2.pdf",w=12,h=16)
+print(figS2)
 tmp <- dev.off()
@@
 fig4 <- cowplot::plot_grid(
-				fig4A + theme(legend.position = "none"),
-				fig4B+ theme(legend.position = "none"),
+				fig4A + theme(legend.position = "none") + mutant_colours,
+				fig4B+ theme(legend.position = "none")  + mutant_colours ,
 				nrow=1,
@@
 
+#scale_colour_discrete(palette = MUTANT_PALETTES)
 fill_legend <- 
@@
     cowplot::get_legend(
-        fig4A + theme(legend.position = "bottom", legend.direction = "horizontal") + scale_fill_discrete(name="Genotype "))
+        fig4A + theme(legend.position = "bottom", legend.direction = "horizontal") + scale_fill_manual(name="Genotype ", values = MUTANT_PALETTES))
         )
@@
 
-fig5 <- cowplot::plot_grid(fig5A + theme(legend.position="none"),
-                           fig5B + theme(legend.position="none") + scale_y_continuous(name="Number") + scale_x_discrete(name="Genotype"),
-                           fig5C + theme(legend.position="none") + scale_x_discrete(name="Genotype"),
+fig5 <- cowplot::plot_grid(fig5A + mutant_colours +theme(legend.position="none"),
+                           fig5B + mutant_colours  + theme(legend.position="none") + scale_y_continuous(name="Number") + scale_x_discrete(name="Genotype"),
+                           fig5C + mutant_colours  + theme(legend.position="none") + scale_x_discrete(name="Genotype"),
                            nrow=1,
@@
   geom_jitter(aes(is_rhythmic, first_peak_period, size=first_peak_rel_power), inherit.aes = F) +
-   scale_fill_discrete(name="Genotype")+scale_size(name= "Peak relative power") +  scale_alpha_discrete(name= "") +
+   scale_fill_manual(name="Genotype", values=MUTANT_PALETTES)+scale_size(name= "Peak relative power") +  scale_alpha_discrete(name= "") +
   theme(legend.position = "bottom")  + 

### 2018-04-20 manuscript.tex
another small rewording to introduce first expls

@@
 
-Fist, we install the \texttt{rethomics} packages (see the webpage for installation instructions) and download the data to extract the zip archive in our working directory.
+Fist, we install the \texttt{rethomics} packages (see the webpage for installation instructions), download the zip archive containing the raw data and extract all files into our working directory.
 Then, we load the necessary \texttt{rethomics} packages:

### 2018-04-23 manuscript.Rnw
revert line numbers and remove todos
author names formated like in plos
oops spelling mistake in marcus' name

@@
 \reversemarginpar
-\usepackage[colorinlistoftodos]{todonotes}
-%\usepackage[colorinlistoftodos,disable]{todonotes}
+%\usepackage[colorinlistoftodos]{todonotes}
+\usepackage[colorinlistoftodos,disable]{todonotes}
 
@@
 Quentin Geissmann\textsuperscript{1*},
-Luis G Rodriguez\textsuperscript{2},
-Esteban J Beckwith\textsuperscript{1},
-Giorgio F Gilestro\textsuperscript{1*}
+Luis Garcia Rodriguez\textsuperscript{2},
+Esteban J. Beckwith\textsuperscript{1},
+Giorgio F. Gilestro\textsuperscript{1*}
 \\
@@
 Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
-Marcus Gosh and Tara Kane for their feedback on the manuscript;
+Marcus Ghosh and Tara Kane for their feedback on the manuscript;
 Brenna Williams, for her help to support multi-beam DAMS;

### 2018-04-24 manuscript.tex
revert line numbers and remove todos
author names formated like in plos
preparation for plos submission!

@@
 To illustrate the ability of \texttt{rethomics} to analyse pre-existing results,
-we gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_ll_2018}.
+we gathered a subset of the data from a recent publication\cite{buhl_quasimodo_2016}, kindly made publicly available by the authors\cite{ogueta_maite_2018_1172980}.
 
@@
 In particular, we outline the necessary steps to analyse two repetitions of the same experiment in which a total of 58 animals were recorded for three to four days in LD and then subjected to constant light for six or seven days.
-The \texttt{metadata.csv} file and all the associated result files can be downloaded on zenodo\cite{ogueta_ll_2018}.
+The \texttt{metadata.csv} file and all the associated result files can be downloaded on zenodo \cite{ogueta_maite_2018_1172980}.
 

### 2018-04-24 results.Rnw
explicit link to zenodo url not ref properly in the biblio
preparation for plos submission!

@@
 
-MUTANT_PALETTES <- c("#babdb6", "#729fcf")
+MUTANT_PALETTES <- c("#babdb6", "#95cc60")
 

### 2018-04-25 manuscript.Rnw
forgot to thank everyone

@@
 Maite Ogueta and Ralf Stanewsky, for making the DAMS results data available;
-Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
+Anne Petzold, Alice French, Hannah Jones, Diana Bicazan and Florencia Fernandez-Chiappe for their comments as early users;
 Marcus Ghosh and Tara Kane for their feedback on the manuscript;

### 2018-05-16 manuscript.Rnw
availability section +CRAN

--- 2018-04-23 manuscript.Rnw
@@
 \section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at
-\href{https://github.com/rethomics}{https://github.com/rethomics/}.
+All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and are listed at
+\href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
+\texttt{behavr} and \texttt{damr} are already hosted on the Comprehensive R Archive Network (CRAN), and all the other packages are either under review or in preparation.
 Extensive installation instructions, as well as reproducible demos and tutorials, are available at

### 2018-05-16 results.Rnw
update fig6 legend

@@
 		E: Average spectrogram across 40 males and 40 females.
-		In D and E, the lines on the right show the marginal power spectra corresponding to the shown spectrograms (average across all time).
+		In D and E, the lines on the right show the marginal power spectra corresponding to the spectrograms shown -- that is the average across 12 hours in either the light (white bar) or dark (black bar) phase.
 		The male data was collected and described in our previous study \cite{geissmann_ethoscopes_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.

### 2018-05-16 cwt/script.R
update all figures specifically fig 6

@@ -4,7 +4,7 @@ library(scopr)
 library(ggetho)
 library(sleepr)
 library(gtools)
-library(TSclust)
+#library(TSclust)
 
 FEMALE_MALE_PALETTE <- c("#be2828ff", "#282896ff")
 CONTROL_SD_PALETTE <- c( "#969696ff", "#3caa3cff")
@@ -74,7 +74,7 @@ library(cowplot)
 
 dt_target <- dt
 
-pdf("/tmp/test_wt.pdf", w=16, h=9)
+#~ pdf("/tmp/test_wt.pdf", w=16, h=9)
 
 #for(i in 1:nrow(dt_target[meta=T])){
 #~ for(i in 1:10){
@@ -112,7 +112,7 @@ wt_analysis <- function(dd){
 	plot_dt <- plot_dt[, .(z=mean(z)), by="x,y"]
 	plot_dt <- plot_dt[, z := sqrt(z)]
 
-#~ 	brks <- c(mins(2), hours(12), hours(24))
+	#brks <- c(mins(2), hours(12), hours(24))
 #~ 	a <- ggplot(dtest[t %between% hours(c(24, 48))], aes(x=t, y=x_rel)) + geom_line() + scale_x_hours() + stat_ld_annotations() +ggtitle(dt_target[i,id, meta=T])
 #~ 	b <- ggplot(out, aes(period, power)) + scale_x_log10(breaks=brks) + geom_line()
 #~ 	c <- ggplot(plot_dt, aes(x=x,y=y,fill=z)) + geom_raster() + scale_x_hours() + 
@@ -168,15 +168,44 @@ for(i in unique(out[id %in% selected , id, meta=T])){
 
 tmp_dt
 out[ , .(z=mean(z)), by=y]
+#out_bak <- copy(out)
+out[, phase := ifelse(x < hours(12), "A", "B")]
+tmp_dt <- rejoin(out[ , .SD[ , .(z=mean(z)), by="y,phase"], by=id])#[,.(z = mean(z)),by="y,sex"]
 
-tmp_dt <- rejoin(out[ , .SD[ , .(z=mean(z)), by=y], by=id])#[,.(z = mean(z)),by="y,sex"]
-dm <- ggplot(tmp_dt[sex=="M"], aes(y=z, x=y)) + stat_pop_etho(method= mean_cl_boot) + coord_flip() +
-		scale_x_continuous(breaks = log2(brks/10))   + scale_y_continuous(limits=c(0,1))
+
[…]

### 2018-05-16 manuscript.tex
all ms recompiled

@@
 
-Several users, in different research groups, have already adopted and are contributing to the future development framework.
+Several users, in different research groups, have already adopted and are contributing to the future development of the framework.
 Several new packages in the \texttt{rethomics} framework are currently envisaged. 

### 2018-09-14 results.Rnw
small typos suggetsed by rev.2

@@
 In order to illustrate the usefulness of \texttt{rethomics}, we provide two examples.
-The first one is a detailed and reproducible description of the loading and analysis activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
+The first one is a detailed and reproducible description of the loading and analysis of activity data in the context of circadian rhythm, using DAM2 (Trikinetics Inc.) data.
 The second one shows how \texttt{rethomics} integrates within \texttt{R} to perform a multi-scale analysis of a periodic behaviour, using continuous wavelet transform, on data generated with ethoscopes\cite{geissmann_ethoscopes_2017}.
@@
 
-The \texttt{id} is a long string of character (for instance, `\texttt{2013-11-19~09:00:00|Monitor36.txt|01}'), 
+The \texttt{id} is a long string of characters (for instance, `\texttt{2013-11-19~09:00:00|Monitor36.txt|01}'), 
 which makes it difficult to read and display as a label on a plot.
@@
 <<cache=TRUE, warning=FALSE, result="hide", message=FALSE>>=
-# we also show a subset of this figure in Fig 4A
+# we also show a subset of this figure as Figure 4
 figS2A <- ggetho(dt, aes(z = moving), multiplot = 2) +
@@
 
-Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
+Lastly, we can generate a table that computes arbitrary population statistics for each genotype:
 <<warning=FALSE>>=
@@
 Being able to keep frequency information over multiple scales is however important in some cases.
-In particular, when interested in the frequency modulation of a rhythm by another --
+In particular, this is relevant to study the frequency modulation of a rhythm by another --
 that is, if the periodicity of a high-frequency rhythm itself can be a function of a lower frequency one.

### 2018-09-14 manuscript.tex
add discussion

@@
 \begin{alltt}
-\hlcom{# we also show a subset of this figure in Fig 4A}
+\hlcom{# we also show a subset of this figure as Figure 4}
 \hlstd{figS2A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
@@
 \end{kframe}
+\end{knitrout}
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in ans[test \& ok] <- rep(yes, length.out = length(ans))[test \& ok]: replacement has length zero}}
+
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in check\_breaks\_labels(breaks, labels): object 'new\_labs' not found}}\end{kframe}
 \end{knitrout}
@@
 
-Lastly, we can generate a table that compute arbitrary population statistics for each genotype:
+Lastly, we can generate a table that computes arbitrary population statistics for each genotype:
 \begin{knitrout}
@@
 % raw data and metadata files are publicly available TODO cite zenodo
-%
+
+\section*{Discussion}
+
+something here
 

### 2018-09-15 functionalities_table.tex
corrections

\documentclass{standalone}
\usepackage{makecell}

\begin{document}
\pagenumbering{gobble}
\begin{tabular}{|l|l|l|}
	\hline
	Package & Role & Main functions \\ \hline \hline
	\multicolumn{ 1}{|c|}{behavr} & utilities & \texttt{behavr()}, \texttt{meta()},\texttt{setbehavr()},\texttt{setmeta()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & metadata operations \textbf{see S1 Fig.} & \texttt{xmd()}, \texttt{rejoin()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & merge and check several tables & \texttt{bind\_behavr\_list()}, \texttt{stitch\_on()} \\ \hline
	\multicolumn{ 1}{|c|}{ggetho} & interface to ggplot (\textbf{see main text}) & \texttt{ggetho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|c|}{} & tile and ``bar tile'' plots (\textbf{\emph{e.g.} Fig. 3 and 4A}) & \texttt{stat\_tile\_etho}, \texttt{stat\_bar\_tile\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & population trends as lines and error bars & \texttt{stat\_pop\_etho()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & circadian biology utilities (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{stat\_ld\_annotations()}, \texttt{geom\_peak()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & smart time scales (auto-adjust labels and ticks) & \texttt{scale\_x\_days()}, \texttt{scale\_x\_hours()}, \dots\\ \hline
	\multicolumn{ 1}{|c|}{damr} & ``link'' metadata to DAM files (\textbf{see main text}) & \texttt{link\_dam\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & \makecell[vl]{``load'' relevant DAM (2 or 5) data as behavr.\\Detects daylight-saving bugs and such} & \texttt{load\_dam()} \\ \hline
	\multicolumn{ 1}{|c|}{scopr} & ``link'' metadata to ethoscope files & \texttt{link\_ethoscope\_metadata()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & ``load'' relevant ethoscope data as behavr & \texttt{load\_ethoscope()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & utilities (list files, fetch remote data, ) & \texttt{experiment\_info()}, \texttt{list\_result\_files()}, \dots \\ \hline
	\multicolumn{ 1}{|c|}{zeitgebr} & compute periodograms (AC, $\chi^2$,LS,Fourrier) & \small{\texttt{*\_periodogram()} (\texttt{*} $\in$ \texttt{\{ac,chi\_sqr,ls,fourrier\}})}\\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & preprocess and standardise methods & \texttt{periodogram()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & find peak periods in periodograms (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{find\_peaks()} \\ \hline
	\multicolumn{ 1}{|c|}{sleepr} & motion thresholding/velocity detection & \texttt{motion\_detectors()}  \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & sleep definition using N minute rule & \texttt{sleep\_annotation()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & data curation: dead animals scored as asleep & \texttt{curate\_dead\_animals()} \\ \cline{ 2- 3}
	\multicolumn{ 1}{|l|}{} & discrete behavioural state bout architecture & \texttt{bout\_analysis()} \\ \hline
\end{tabular}

\end{document}

### 2018-09-15 manuscript.Rnw
update fig6 legend
add discussion
corrections

--- 2018-04-25 manuscript.Rnw
@@
 as a collection of packages (Fig~\ref{fig:fig-1}).
+A summary of the functionality of all packages is represented in a supplementary table (\nameref{S1-Table}).
 Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
@@
 % raw data and metadata files are publicly available TODO cite zenodo
-%
 
-\section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and listed at
-\href{https://github.com/rethomics}{https://github.com/rethomics/}.
+\section*{Discussion}
+The framework  we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data.
+We showed that \texttt{rethomics} can readily be used to analyse specifically circadian data.
+As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to the rigid array of GUI tools which cannot interoperate.
+In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
+data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistics (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
+In addition, to the canonical analysis we demonstrated, the richness of the \texttt{R} environment renders rethomics readily available for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).
+
+Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation tools whilst embracing the diversity of recording equipment.
+By design, rethomics is very flexible and not limited to the study of circadian rhythms.
+In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general solution to link data to metadata that come with no assumption regarding the type and number of variables stored.
+Therefore, we anticipate it will apply to  a variety contexts and scales.
+
+\section*{Availability}
+All \texttt{rethomics} packages are listed at \href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
+and are hosted on the official Comprehensive \texttt{R} Archive Network (CRAN) under the terms of the GPLv3 license. 
 Extensive installation instructions, as well as reproducible demos and tutorials, are available at
@@
 
-Several users, in different research groups, have already adopted and are contributing to the future development framework.
-Several new packages in the \texttt{rethomics} framework are currently envisaged. 
-They include utilities to input new behaviour tracking methods and analyse multi-animal interactions.
-
 \section*{Supporting information}
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+
+\paragraph*{S1 Table.}
+\label{S1-Table}
+{\bf List of packages and main functionalities implemented in \texttt{rethomics}.}
+The last column shows a list of user-space functions for each role.
+Extensive list of all functions is available in the documentation of each package.
 

### 2018-09-15 manuscript.tex
corrections

@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in ans[test \& ok] <- rep(yes, length.out = length(ans))[test \& ok]: replacement has length zero}}
-
-{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in check\_breaks\_labels(breaks, labels): object 'new\_labs' not found}}\end{kframe}
-\end{knitrout}
-
 \subsubsection*{Periodograms}
@@
 \section*{Discussion}
+The framework  we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data.
+We showed that \texttt{rethomics} can readily be used to analyse specifically circadian data.
+As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to the rigid array of GUI tools which cannot interoperate.
+In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
+data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistics (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
+In addition, to the canonical analysis we demonstrated, the richness of the \texttt{R} environment renders rethomics readily available for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).
 
-something here
+Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation tools whilst embracing the diversity of recording equipment.
+By design, rethomics is very flexible and not limited to the study of circadian rhythms.
+In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general solution to link data to metadata that come with no assumption regarding the type and number of variables stored.
+Therefore, we anticipate it will apply to  a variety contexts and scales.
 
-\section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and are listed at
-\href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
-\texttt{behavr} and \texttt{damr} are already hosted on the Comprehensive R Archive Network (CRAN), and all the other packages are either under review or in preparation.
+\section*{Availability}
+All \texttt{rethomics} packages are listed at \href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
+and are hosted on the official Comprehensive \texttt{R} Archive Network (CRAN) under the terms of the GPLv3 license. 
 Extensive installation instructions, as well as reproducible demos and tutorials, are available at
@@
 
-Several users, in different research groups, have already adopted and are contributing to the future development of the framework.
-Several new packages in the \texttt{rethomics} framework are currently envisaged. 
-They include utilities to input new behaviour tracking methods and analyse multi-animal interactions.
-
 \section*{Supporting information}

### 2018-09-15 cwt/script.R

@@ -30,8 +30,6 @@ dt <- load_ethoscope(met,
 					   ncores=1)
 
 
-
-
 curate_data <- function(data){
   #data[, t := t - days(xmv(baseline_days))]
   #data <- data[is_interpolated == F]
@@ -112,7 +110,7 @@ wt_analysis <- function(dd){
 	plot_dt <- plot_dt[, .(z=mean(z)), by="x,y"]
 	plot_dt <- plot_dt[, z := sqrt(z)]
 
-	#brks <- c(mins(2), hours(12), hours(24))
+# brks <- c(mins(2), hours(12), hours(24))
 #~ 	a <- ggplot(dtest[t %between% hours(c(24, 48))], aes(x=t, y=x_rel)) + geom_line() + scale_x_hours() + stat_ld_annotations() +ggtitle(dt_target[i,id, meta=T])
 #~ 	b <- ggplot(out, aes(period, power)) + scale_x_log10(breaks=brks) + geom_line()
 #~ 	c <- ggplot(plot_dt, aes(x=x,y=y,fill=z)) + geom_raster() + scale_x_hours() + 

### 2018-09-23 functionalities_table.tex
work++

@@
 	\multicolumn{ 1}{|l|}{} & circadian biology utilities (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{stat\_ld\_annotations()}, \texttt{geom\_peak()} \\ \cline{ 2- 3}
+	\multicolumn{ 1}{|l|}{} & visualise spectrograms (\textbf{\emph{e.g.} CWT, Fig. 6}) & \texttt{ggspectro()} \\ \cline{ 2- 3}
 	\multicolumn{ 1}{|l|}{} & smart time scales (auto-adjust labels and ticks) & \texttt{scale\_x\_days()}, \texttt{scale\_x\_hours()}, \dots\\ \hline
@@
 	\multicolumn{ 1}{|l|}{} & utilities (list files, fetch remote data, ) & \texttt{experiment\_info()}, \texttt{list\_result\_files()}, \dots \\ \hline
-	\multicolumn{ 1}{|c|}{zeitgebr} & compute periodograms (AC, $\chi^2$,LS,Fourrier) & \small{\texttt{*\_periodogram()} (\texttt{*} $\in$ \texttt{\{ac,chi\_sqr,ls,fourrier\}})}\\ \cline{ 2- 3}
+	\multicolumn{ 1}{|c|}{zeitgebr} & compute periodograms (CWT, AC, $\chi^2$,LS,Fourrier) & \small{\texttt{*\_periodogram()} (\texttt{*} $\in$ \texttt{\{cwt, ac,chi\_sqr,ls,fourrier\}})}\\ \cline{ 2- 3}
 	\multicolumn{ 1}{|l|}{} & preprocess and standardise methods & \texttt{periodogram()} \\ \cline{ 2- 3}
-	\multicolumn{ 1}{|l|}{} & find peak periods in periodograms (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{find\_peaks()} \\ \hline
+	\multicolumn{ 1}{|l|}{} & find peak periods in periodograms (\textbf{\emph{e.g.} Fig. 4 and 5A}) & \texttt{find\_peaks()} \\ \cline{ 2- 3}
+	\multicolumn{ 1}{|l|}{} & compute CWT spectrograms & \texttt{spectrogram()} \\ \hline	
 	\multicolumn{ 1}{|c|}{sleepr} & motion thresholding/velocity detection & \texttt{motion\_detectors()}  \\ \cline{ 2- 3}
-	\multicolumn{ 1}{|l|}{} & sleep definition using N minute rule & \texttt{sleep\_annotation()} \\ \cline{ 2- 3}
+	\multicolumn{ 1}{|l|}{} & sleep definition using N-minute rule & \texttt{sleep\_annotation()} \\ \cline{ 2- 3}
 	\multicolumn{ 1}{|l|}{} & data curation: dead animals scored as asleep & \texttt{curate\_dead\_animals()} \\ \cline{ 2- 3}
-	\multicolumn{ 1}{|l|}{} & discrete behavioural state bout architecture & \texttt{bout\_analysis()} \\ \hline
+	\multicolumn{ 1}{|l|}{} & discrete behavioural state architecture & \texttt{bout\_analysis()} \\ \hline
 \end{tabular}

### 2018-09-23 manuscript.Rnw
work++

@@
 as a collection of packages (Fig~\ref{fig:fig-1}).
-A summary of the functionality of all packages is represented in a supplementary table (\nameref{S1-Table}).
+A summary of the functionalities of all packages is represented in a supplementary table (\nameref{S1-Table}).
 Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
@@
 The framework  we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data.
-We showed that \texttt{rethomics} can readily be used to analyse specifically circadian data.
-As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to the rigid array of GUI tools which cannot interoperate.
+We showed that \texttt{rethomics} can readily be used to analyse circadian data.
+As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to a rigid array of GUI tools which cannot interoperate.
 In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
-data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistics (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
-In addition, to the canonical analysis we demonstrated, the richness of the \texttt{R} environment renders rethomics readily available for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).
+data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
+In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment renders rethomics particularly interesting for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).
 
-Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation tools whilst embracing the diversity of recording equipment.
+Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation solutions whilst embracing the diversity of recording equipment.
 By design, rethomics is very flexible and not limited to the study of circadian rhythms.
-In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general solution to link data to metadata that come with no assumption regarding the type and number of variables stored.
-Therefore, we anticipate it will apply to  a variety contexts and scales.
+In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general structure to link data to metadata that come with no assumption regarding the type and number of variables stored.
+Therefore, we anticipate it will apply to  a variety of contexts and scales.
+For instance, in the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (\emph{i.e.} feeding, metabolic activity, heart rate, position and so on).
+In collaboration with the teams providing of these methods, we envisage the development of device-specific packages that will load data into \texttt{rethomics}, making visualisation and analysis immediately available.
+For instance, the developer any behaviour recording method can simply write a small package to read their data in a \texttt{behavr} table and benefit from the functionalities available.
+Furthermore, such an approach becomes particularily useful in order to compare methods or study behaviours that are recorded by different devices, on the same individuals.
+Indeed, in these contexts, it becomes apparent that analysis must be unified at some level.
+
+In conclusion, \texttt{rethomics} already delivers state-of-the-art tools to study circadian rhythm using conventional tools, but was designed in an inclusive and extensible manner.
+As such, if fills a unique niche in a community that is becoming increasingly quantitative.
 

### 2018-09-23 manuscript.tex
work++

@@
 To illustrate how \texttt{rethomic} integrates with other packages and render such non-mainstream analysis possible, 
-we performed a wavelet analysis of the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
-We used the package \texttt{scopr}, part of \texttt{rethomics}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
+we developped, as part of the \texttt{zeitgebr} package, utilities to perform wavelet analysis, which we then applied to the position of 80 naive fruit flies (40 females and 40 males) in their glass tubes.
+We used the package \texttt{scopr}, to load five days of ethoscope positional data, which we sampled at 0.1~Hz.
 Our variable of interest is the position of animals in their tube (from the food end, $Position = 0$,  to the cotton end, $Position = 1$).
@@
 
-In order to compute CWT, we used the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018}.
-We then averaged the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
+In order to compute CWT, we interfaced the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018} in \texttt{rethomics}. 
+This interface allowed us to average the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
 and for the population(Fig~\ref{fig:fig-6}E).

### 2018-09-23 results.Rnw
corrections
work++

@@
 
-ylabs <- ggplot_build(fig4A)$layout$panel_ranges[[1]]$y.labels
+pl <- ggplot_build(fig4A)
+
+if(is.null(pl$layout$panel_ranges[[1]]))
+  ## old ggplot (< 3.0.0)
+  ylabs <- pl$layout$panel_params[[1]]$y.labels else
+  ylabs <- pl$layout$panel_ranges[[1]]$y.labels
 

### 2018-10-04 manuscript.tex
rebutal++

@@
 
-In order to compute CWT, we interfaced the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018} in \texttt{rethomics}. 
-This interface allowed us to average the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
+We implemented specific utilities to compute and display CWT spectrograms in \texttt{rethomics}, based on the \texttt{WaveletComp} package\cite{schmidbauer_waveletcomp_2018}. 
+We then used this interface to average the result of the five consecutive days in the time/period domain over one circadian day both for the two representative animals (Fig~\ref{fig:fig-6}D)
 and for the population(Fig~\ref{fig:fig-6}E).

### 2018-10-09 manuscript.Rnw
almost resubmitted

--- 2018-05-16 manuscript.Rnw
@@
 as a collection of packages (Fig~\ref{fig:fig-1}).
+A summary of the functionalities of all packages is represented in a supplementary table (\nameref{S1-Table}).
 Such modular architecture follows the model of modern frameworks such as the \texttt{tidyverse}\cite{wickham_tidyverse_2017}, which results in increased testability, maintainability and adaptability.
@@
 % raw data and metadata files are publicly available TODO cite zenodo
-%
 
-\section*{Availability and Future Directions}
-All packages in the \texttt{rethomics} framework are available under the terms of the GPLv3 license and are listed at
-\href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
-\texttt{behavr} and \texttt{damr} are already hosted on the Comprehensive R Archive Network (CRAN), and all the other packages are either under review or in preparation.
-Extensive installation instructions, as well as reproducible demos and tutorials, are available at
+\section*{Discussion}
+
+The framework we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data. 
+As a practical case study, in this article, we showed how \texttt{rethomics} can readily be used to analyse circadian and sleep data.
+As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to a rigid array of GUI tools which cannot interoperate.
+In particular, within the R environment, rethomic features a streamlined workflow (Fig.~\ref{fig:fig-1}): data loading; quality control (Fig.~\ref{fig:fig-3}); per-individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
+In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment makes \texttt{rethomics} particularly interesting for exploratory work: as a case in point, we showed here the use of continuous wavelet transform to study the modulation of walking pace (Fig.~\ref{fig:fig-6}).
+Altogether, \texttt{rethomics} will allow the community to share solutions for the analysis and visualisation of behavioural data originated from a plethora of recording equipment.
+By design, \texttt{rethomics} is meant to be flexible and certainly not limited to the study of circadian rhythms.
+In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general structure to link data to metadata that come with no assumption regarding the type and number of variables stored.
+Therefore, we anticipate it will apply to a variety of contexts and scales.
+In the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (i.e.
+feeding, metabolic activity, heart rate, position and so on).
+All these devices and behavioural paradigm could take advantage of the skeletal structure in \texttt{rethomics} and lead to  the development of device-specific packages.
+Such an approach becomes particularly useful in order to compare methods or study behaviours that are recorded by different devices on the same individual animals.
+
+\section*{Availability}
+All \texttt{rethomics} packages are listed at \href{https://rethomics.github.io/intro.html}{https://rethomics.github.io/intro.html}.
+and are hosted on the official Comprehensive \texttt{R} Archive Network (CRAN) under the terms of the GPLv3 license. 
+Installation instructions, as well as reproducible demos and tutorials, are available at
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
 All packages are continuously integrated and unit tested on several versions of \texttt{R} to minimise the risk of present and future issues.
-
-Several users, in different research groups, have already adopted and are contributing to the future development framework.
-Several new packages in the \texttt{rethomics} framework are currently envisaged. 
-They include utilities to input new behaviour tracking methods and analyse multi-animal interactions.
 

### 2018-11-21 manuscript-final.tex
final ms

--- 2018-10-04 manuscript.tex
@@
 % Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
+%\bibliographystyle{plos2015}
 %\bibliographystyle{apalike}
@@
 \section*{Discussion}
+
 The framework we presented in this article describes a general computational set of tools to store, manipulate and analyse behavioural data. 
-We showed that \texttt{rethomics} can readily be used to analyse circadian data.
+As a practical case study, in this article, we showed how \texttt{rethomics} can readily be used to analyse circadian and sleep data.
 As such, it already fills a clear gap in the sleep and circadian community by providing an open, extensible and programmatic toolbox as opposed to a rigid array of GUI tools which cannot interoperate.
-In particular, within the \texttt{R} environment, \texttt{rethomic} features a streamlined workflow (Fig.~\ref{fig:fig-1}):
-data loading; quality control (Fig.~\ref{fig:fig-3}); per individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
-In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment renders rethomics particularly interesting for exploratory work such as the use of continuous wavelet transform to study the modulation of walking pace(Fig.~\ref{fig:fig-6}).
-
-Altogether, our tool provides a needed opportunity for the community to share analytic and visualisation solutions whilst embracing the diversity of recording equipment.
-By design, rethomics is very flexible and not limited to the study of circadian rhythms.
+In particular, within the R environment, rethomic features a streamlined workflow (Fig.~\ref{fig:fig-1}): data loading; quality control (Fig.~\ref{fig:fig-3}); per-individual visualisation (Fig.~\ref{fig:fig-4}A) and computation (Fig.~\ref{fig:fig-4}B); group averages and statistical analysis (Fig.~\ref{fig:fig-5}), ultimately resulting in publication-quality figures.
+In addition to the canonical analysis we presented in this study, the richness of the \texttt{R} environment makes \texttt{rethomics} particularly interesting for exploratory work: as a case in point, we showed here the use of continuous wavelet transform to study the modulation of walking pace (Fig.~\ref{fig:fig-6}).
+Altogether, \texttt{rethomics} will allow the community to share solutions for the analysis and visualisation of behavioural data originated from a plethora of recording equipment.
+By design, \texttt{rethomics} is meant to be flexible and certainly not limited to the study of circadian rhythms.
 In fact, \texttt{behavr} tables (Fig.~\ref{fig:fig-2}) are a general structure to link data to metadata that come with no assumption regarding the type and number of variables stored.
 Therefore, we anticipate it will apply to a variety of contexts and scales.
-For instance, in the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (\emph{i.e.} feeding, metabolic activity, heart rate, position and so on).
-In collaboration with the teams providing of these methods, we envisage the development of device-specific packages that will load data into \texttt{rethomics}, making visualisation and analysis immediately available.
-For instance, the developer any behaviour recording method can simply write a small package to read their data in a \texttt{behavr} table and benefit from the functionalities available.
-Furthermore, such an approach becomes particularily useful in order to compare methods or study behaviours that are recorded by different devices, on the same individuals.
-Indeed, in these contexts, it becomes apparent that analysis must be unified at some level.
-
-In conclusion, \texttt{rethomics} already delivers state-of-the-art tools to study circadian rhythm using conventional tools, but was designed in an inclusive and extensible manner.
-As such, if fills a unique niche in a community that is becoming increasingly quantitative.
+In the last decade, multiple devices have been developed -- either commercially or academically -- to record longitudinal data on several animal models (i.e.
+feeding, metabolic activity, heart rate, position and so on).
+All these devices and behavioural paradigm could take advantage of the skeletal structure in \texttt{rethomics} and lead to  the development of device-specific packages.
+Such an approach becomes particularly useful in order to compare methods or study behaviours that are recorded by different devices on the same individual animals.
 
@@
 and are hosted on the official Comprehensive \texttt{R} Archive Network (CRAN) under the terms of the GPLv3 license. 
-Extensive installation instructions, as well as reproducible demos and tutorials, are available at
+Installation instructions, as well as reproducible demos and tutorials, are available at
 \href{https://rethomics.github.io/}{https://rethomics.github.io/}.
@@
 
-\bibliography{manuscript}{}
+\begin{thebibliography}{10}
+
+	\bibitem{sokolowski_drosophila_2001}
+	Sokolowski MB.
+	\newblock {\emph{Drosophila}}: {{Genetics}} Meets Behaviour.
+	\newblock Nature Reviews Genetics. 2001;2(11):879--890.
+	\newblock doi:{10.1038/35098592}.
+
+	\bibitem{stephens_big_2015}
+	Stephens ZD, Lee SY, Faghri F, Campbell RH, Zhai C, Efron MJ, et~al.
+	\newblock Big {{Data}}: {{Astronomical}} or {{Genomical}}?
+	\newblock PLOS Biology. 2015;13(7):e1002195.
+	\newblock doi:{10.1371/journal.pbio.1002195}.
+
+	\bibitem{schatz_biological_2015}
+	Schatz MC.
+	\newblock Biological Data Sciences in Genome Research.
+	\newblock Genome Research. 2015;25(10):1417--1422.
+	\newblock doi:{10.1101/gr.191684.115}.
+
+	\bibitem{sejnowski_putting_2014}
+	Sejnowski TJ, Churchland PS, Movshon JA.
+	\newblock {Putting big data to good use in neuroscience}.
+	\newblock Nature Neuroscience. 2014;17(11):1440--1441.
+	\newblock doi:{10.1038/nn.3839}.
+
+	\bibitem{branson_high-throughput_2009}
+	Branson K, Robie AA, Bender J, Perona P, Dickinson MH.
+	\newblock High-Throughput Ethomics in Large Groups of {{{\emph{Drosophila}}}}.
+	\newblock Nature Methods. 2009;6(6):451--457.
+	\newblock doi:{10.1038/nmeth.1328}.
+
+	\bibitem{reiser_ethomics_2009}
+	Reiser M.
+	\newblock The Ethomics Era?
+	\newblock Nature Methods. 2009;6(6):413--414.
+	\newblock doi:{10.1038/nmeth0609-413}.
+
+	\bibitem{anderson_toward_2014}
+	Anderson DJ, Perona P.
+	\newblock Toward a {{Science}} of {{Computational Ethology}}.
+	\newblock Neuron. 2014;84(1):18--31.
+	\newblock doi:{10.1016/j.neuron.2014.09.005}.
+
+	\bibitem{brown_ethology_2018}
+	Brown AEX, de~Bivort B.
+	\newblock Ethology as a Physical Science.
+	\newblock Nature Physics. 2018; p.~1.
+	\newblock doi:{10.1038/s41567-018-0093-0}.
+
+	\bibitem{berman_measuring_2018}
+	Berman GJ.
+	\newblock Measuring Behavior across Scales.
+	\newblock BMC Biology. 2018;16:23.
+	\newblock doi:{10.1186/s12915-018-0494-7}.
+
+	\bibitem{faville_how_2015}
+	Faville R, Kottler B, Goodhill GJ, Shaw PJ, {van Swinderen} B.
+	\newblock How Deeply Does Your Mutant Sleep? {{Probing}} Arousal to Better
+	Understand Sleep Defects in {{{\emph{Drosophila}}}}.
+	\newblock Scientific Reports. 2015;5:8454.
+	\newblock doi:{10.1038/srep08454}.
+
+	\bibitem{pelkowski_novel_2011}
+	Pelkowski SD, Kapoor M, Richendrfer HA, Wang X, Colwill RM, Creton R.
+	\newblock A Novel High-Throughput Imaging System for Automated Analyses of
+	Avoidance Behavior in Zebrafish Larvae.
+	\newblock Behavioural Brain Research. 2011;223(1):135--144.
+	\newblock doi:{10.1016/j.bbr.2011.04.033}.
+
+	\bibitem{itskov_automated_2014}
+	Itskov PM, Moreira JM, Vinnik E, Lopes G, Safarik S, Dickinson MH, et~al.
+	\newblock Automated Monitoring and Quantitative Analysis of Feeding Behaviour
+	in {{{\emph{Drosophila}}}}.
+	\newblock Nature Communications. 2014;5:4560.
+	\newblock doi:{10.1038/ncomms5560}.
+
+	\bibitem{ro_flic_2014}
+	Ro J, Harvanek ZM, Pletcher SD.
+	\newblock {{FLIC}}: {{High}}-{{Throughput}}, {{Continuous Analysis}} of
+	{{Feeding Behaviors}} in {{Drosophila}}.
+	\newblock PLOS ONE. 2014;9(6):e101107.
+	\newblock doi:{10.1371/journal.pone.0101107}.
+
+	\bibitem{dankert_automated_2009}
+	Dankert H, Wang L, Hoopfer ED, Anderson DJ, Perona P.
+	\newblock Automated {{Monitoring}} and {{Analysis}} of {{Social Behavior}} in
+	{{Drosophila}}.
+	\newblock Nature methods. 2009;6(4):297--303.
+	\newblock doi:{10.1038/nmeth.1310}.
+
+	\bibitem{tsai_image_2012}
+	Tsai HY, Huang YW.
+	\newblock Image {{Tracking Study}} on {{Courtship Behavior}} of {{Drosophila}}.
+	\newblock PLOS ONE. 2012;7(4):e34784.
+	\newblock doi:{10.1371/journal.pone.0034784}.
+
+	\bibitem{swierczek_high-throughput_2011}
+	Swierczek NA, Giles AC, Rankin CH, Kerr RA.
+	\newblock High-Throughput Behavioral Analysis in {{{\emph{C}}}}{\emph{.
+			Elegans}}.
+	\newblock Nature Methods. 2011;8(7):592--598.
+	\newblock doi:{10.1038/nmeth.1625}.
+
+	\bibitem{perez-escudero_idtracker_2014}
+	P{\'e}rez-Escudero A, Vicente-Page J, Hinz RC, Arganda S, {de Polavieja} GG.
+	\newblock {{idTracker}}: Tracking Individuals in a Group by Automatic
+	Identification of Unmarked Animals.
+	\newblock Nature Methods. 2014;11(7):743--748.
+	\newblock doi:{10.1038/nmeth.2994}.
+
+	\bibitem{robie_mapping_2017}
+	Robie AA, Hirokawa J, Edwards AW, Umayam LA, Lee A, Phillips ML, et~al.
+	\newblock Mapping the {{Neural Substrates}} of {{Behavior}}.
+	\newblock Cell. 2017;170(2):393--406.e28.
+	\newblock doi:{10.1016/j.cell.2017.06.032}.
+
+	\bibitem{kabra_jaaba_2013}
+	Kabra M, Robie AA, Rivera-Alba M, Branson S, Branson K.
+	\newblock {{JAABA}}: Interactive Machine Learning for Automatic Annotation of
+	Animal Behavior.
+	\newblock Nature Methods. 2013;10(1):64--67.
+	\newblock doi:{10.1038/nmeth.2281}.
+
+	\bibitem{lopes_bonsai_2015}
+	Lopes G, Bonacchi N, Fraz{\~a}o J, Neto JP, Atallah BV, Soares S, et~al.
+	\newblock Bonsai: An Event-Based Framework for Processing and Controlling Data
+	Streams.
+	\newblock Frontiers in Neuroinformatics. 2015;9.
+	\newblock doi:{10.3389/fninf.2015.00007}.
+
+	\bibitem{geissmann_ethoscopes_2017}
+	Geissmann Q, Rodriguez LG, Beckwith EJ, French AS, Jamasb AR, Gilestro GF.
+	\newblock Ethoscopes: {{An}} Open Platform for High-Throughput Ethomics.
+	\newblock PLOS Biology. 2017;15(10):e2003026.
+	\newblock doi:{10.1371/journal.pbio.2003026}.
+
+	\bibitem{roumpeka_review_2017}
+	Roumpeka DD, Wallace RJ, Escalettes F, Fotheringham I, Watson M.
+	\newblock A {{Review}} of {{Bioinformatics Tools}} for {{Bio}}-{{Prospecting}}
+	from {{Metagenomic Sequence Data}}.
+	\newblock Frontiers in Genetics. 2017;8.
+	\newblock doi:{10.3389/fgene.2017.00023}.
+
+	\bibitem{huber_orchestrating_2015}
+	Huber W, Carey VJ, Gentleman R, Anders S, Carlson M, Carvalho BS, et~al.
+	\newblock Orchestrating High-Throughput Genomic Analysis with {{Bioconductor}}.
+	\newblock Nature Methods. 2015;12(2):115--121.
+	\newblock doi:{10.1038/nmeth.3252}.
+
+	\bibitem{leipzig_review_2017}
+	Leipzig J.
+	\newblock A Review of Bioinformatic Pipeline Frameworks.
+	\newblock Briefings in Bioinformatics. 2017;18(3):530--536.
+	\newblock doi:{10.1093/bib/bbw020}.
+
+	\bibitem{peng_reproducible_2011}
+	Peng RD.
+	\newblock Reproducible {{Research}} in {{Computational Science}}.
+	\newblock Science. 2011;334(6060):1226--1227.
+	\newblock doi:{10.1126/science.1213847}.
+
+	\bibitem{stodden_toward_2013}
+	Stodden V, Guo P, Ma Z.
+	\newblock Toward {{Reproducible Computational Research}}: {{An Empirical
+			Analysis}} of {{Data}} and {{Code Policy Adoption}} by {{Journals}}.
+	\newblock PLOS ONE. 2013;8(6):e67111.
+	\newblock doi:{10.1371/journal.pone.0067111}.
+
+	\bibitem{hashem_rise_2015}
+	Hashem IAT, Yaqoob I, Anuar NB, Mokhtar S, Gani A, Ullah~Khan S.
+	\newblock The Rise of ``Big Data'' on Cloud Computing: {{Review}} and Open
+	Research Issues.
+	\newblock Information Systems. 2015;47:98--115.
+	\newblock doi:{10.1016/j.is.2014.07.006}.
+
+	\bibitem{r_core_team_r_2017}
+	{R Core Team}.
+	\newblock R: {{A Language}} and {{Environment}} for {{Statistical Computing}}.
+	\newblock Vienna, Austria: {R Foundation for Statistical Computing}; 2017.
+
+	\bibitem{wickham_tidyverse_2017}
+	Wickham H.
+	\newblock Tidyverse: {{Easily Install}} and {{Load}} the '{{Tidyverse}}'; 2017.
+
+	\bibitem{dowle_data.table_2017}
+	Dowle M, Srinivasan A.
+	\newblock Data.Table: {{Extension}} of `data.Frame`; 2017.
+
+	\bibitem{wickham_ggplot2_2016}
+	Wickham H.
+	\newblock Ggplot2: {{Elegant Graphics}} for {{Data Analysis}}.
+	\newblock {Springer}; 2016.
+
+	\bibitem{sokolove_chi_1978}
+	Sokolove PG, Bushell WN.
+	\newblock The Chi Square Periodogram: {{Its}} Utility for Analysis of Circadian
+	Rhythms.
+	\newblock Journal of Theoretical Biology. 1978;72(1):131--160.
+	\newblock doi:{10.1016/0022-5193(78)90022-X}.
+
+	\bibitem{ruf_lomb-scargle_1999}
+	Ruf T.
+	\newblock The {{Lomb}}-{{Scargle Periodogram}} in {{Biological Rhythm
+			Research}}: {{Analysis}} of {{Incomplete}} and {{Unequally Spaced
+			Time}}-{{Series}}.
+	\newblock Biological Rhythm Research. 1999;30(2):178--201.
+	\newblock doi:{10.1076/brhm.30.2.178.1422}.
+
+	\bibitem{dubowy_circadian_2017}
+	Dubowy C, Sehgal A.
+	\newblock Circadian {{Rhythms}} and {{Sleep}} in {{Drosophila}} Melanogaster.
+	\newblock Genetics. 2017;205(4):1373--1397.
+	\newblock doi:{10.1534/genetics.115.185157}.
+
+	\bibitem{buhl_quasimodo_2016}
+	Buhl E, Bradlaugh A, Ogueta M, Chen KF, Stanewsky R, Hodge JJL.
+	\newblock Quasimodo Mediates Daily and Acute Light Effects on {{Drosophila}}
+	Clock Neuron Excitability.
+	\newblock Proceedings of the National Academy of Sciences.
+	2016;113(47):13486--13491.
+	\newblock doi:{10.1073/pnas.1606547113}.
+
+	\bibitem{ogueta_maite_2018_1172980}
+	Ogueta M, Stanewsky R.
+	\newblock {{LL Behaviour}} of {{TIM Gal4}} $>$ {{NKCC OX}} and {{NKCC OX}} / +
+	Flies.
+	\newblock Zenodo. 2018;doi:{10.5281/zenodo.1172980}.
+
+	\bibitem{aguiar-conraria_business_2011}
+	Aguiar-Conraria L, Joana~Soares M.
+	\newblock Business Cycle Synchronization and the {{Euro}}: {{A}} Wavelet
+	Analysis.
+	\newblock Journal of Macroeconomics. 2011;33(3):477--489.
+	\newblock doi:{10.1016/j.jmacro.2011.02.005}.
+
+	\bibitem{lau_climate_1995}
+	Lau KM, Weng H.
+	\newblock Climate {{Signal Detection Using Wavelet Transform}}: {{How}} to
+	{{Make}} a {{Time Series Sing}}.
+	\newblock Bulletin of the American Meteorological Society.
+	1995;76(12):2391--2402.
+	\newblock doi:{10.1175/1520-0477(1995)076<2391:CSDUWT>2.0.CO;2}.
+
+	\bibitem{cazelles_wavelet_2008}
+	Cazelles B, Chavez M, Berteaux D, M{\'e}nard F, Vik JO, Jenouvrier S, et~al.
+	\newblock Wavelet Analysis of Ecological Time Series.
+	\newblock Oecologia. 2008;156(2):287--304.
+	\newblock doi:{10.1007/s00442-008-0993-2}.
+
+	\bibitem{grossmann_decomposition_1984}
+	Grossmann A, Morlet J.
+	\newblock Decomposition of {{Hardy Functions}} into {{Square Integrable
+			Wavelets}} of {{Constant Shape}}.
+	\newblock SIAM Journal on Mathematical Analysis. 1984;15(4):723--736.
+	\newblock doi:{10.1137/0515056}.
+
+	\bibitem{leise_wavelet_2013}
+	Leise TL.
+	\newblock Wavelet Analysis of Circadian and Ultradian Behavioral Rhythms.
+	\newblock Journal of Circadian Rhythms. 2013;11:5.
+	\newblock doi:{10.1186/1740-3391-11-5}.
+
+	\bibitem{schmidbauer_waveletcomp_2018}
+	Schmidbauer AR, Harald. {{WaveletComp}}: {{Computational Wavelet Analysis}};
+	2018.
+
+\end{thebibliography}
 

### 2018-12-10 manuscript-final.tex
plos editorial tasks

@@
 % Use the PLoS provided BiBTeX style
-%\bibliographystyle{plos2015}
+\bibliographystyle{plos2015}
 %\bibliographystyle{apalike}
@@
 Double-plotted actograms are a common choice to visualise the periodicity and rhythmicity in circadian experiments.
-In \nameref{S2-Fig}A, we show a double-plotted actogram for each animal.
+In \hyperref[S2-Fig]{S2A~Fig.}, we show a double-plotted actogram for each animal.
 A selected sample of four individuals for each genotype is shown in Fig~\ref{fig:fig-4}A.
@@
 Periodograms for the subset of eight animals used in Fig~\ref{fig:fig-4}A are shown in Fig~\ref{fig:fig-4}B. 
-See \nameref{S2-Fig}B for the visualisation of all individuals.
+See \hyperref[S2-Fig]{S2B~Fig.} for the visualisation of all individuals.
 
@@
 		In D and E, the lines on the right show the marginal power spectra corresponding to the spectrograms shown -- that is the average across 12 hours in either the light (white bar) or dark (black bar) phase.
-		The male data was collected and described in our previous study \cite{geissmann_ethoscopes_2017} (controls in figure~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
+		The male data was collected and described in our previous study \cite{geissmann_ethoscopes_2017} (controls in Fig~5M-P) and the females data was acquired in parallel, in the same experimental conditions, but not previously published.
 	}

### 2018-12-12 manuscript-final.tex
change names of the variables refering to sup. figs.

@@
 \hlcom{# we also show a subset of this figure as Figure 4}
-\hlstd{figS2A} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
+\hlstd{S2A_fig} \hlkwb{<-} \hlkwd{ggetho}\hlstd{(dt,} \hlkwd{aes}\hlstd{(}\hlkwc{z} \hlstd{= moving),} \hlkwc{multiplot} \hlstd{=} \hlnum{2}\hlstd{)} \hlopt{+}
             \hlcom{# one could also use stat_tile_etho}
@@
 \hlcom{# we also show a subset of this figure in supplementary materials}
-\hlstd{figS2B} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak} \hlstd{= peak))} \hlopt{+}
+\hlstd{S2B_fig} \hlkwb{<-} \hlkwd{ggperio}\hlstd{(per_dt,} \hlkwd{aes}\hlstd{(}\hlkwc{y} \hlstd{= power,} \hlkwc{peak} \hlstd{= peak))} \hlopt{+}
                   \hlcom{# periododogram drawn as a line}

### 2018-12-12 results.Rnw
change names of the variables refering to sup. figs.

@@
 # we also show a subset of this figure as Figure 4
-figS2A <- ggetho(dt, aes(z = moving), multiplot = 2) +
+S2A_fig <- ggetho(dt, aes(z = moving), multiplot = 2) +
             # one could also use stat_tile_etho
@@
 # we also show a subset of this figure in supplementary materials
-figS2B <- ggperio(per_dt, aes(y = power, peak = peak)) + 
+S2B_fig <- ggperio(per_dt, aes(y = power, peak = peak)) + 
                   # periododogram drawn as a line
@@
 #=================================================
-figS2A <- figS2A +  theme(strip.text.x = element_text(size =8))
-figS2B <- figS2B +  theme(strip.text.x = element_text(size =8))
+
+S2A_fig <- S2A_fig +  theme(strip.text.x = element_text(size =8))
+S2B_fig <- S2B_fig +  theme(strip.text.x = element_text(size =8))
 
 figS2 <- cowplot::plot_grid(
-                          figS2A,
-                          figS2B,
+                          S2A_fig,
+                          S2B_fig,
                           nrow=1,
